---
title: "Ensemble Model Using LSEI"
author: "Ensemble Team"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# SETTING UP LIBRARIES
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(knitr)
library(scales)
library(limSolve)  # For lsei() function
```

# Ensemble Model Overview

## Methodology: Stacking with LSEI

**Ensemble Approach:**
- Combine predictions from multiple base models (RF, XGBoost, Lasso, Ridge, PCR, Elastic Net)
- Learn optimal weights using constrained least squares (LSEI)
- Constraints: Weights sum to 1, all weights non-negative

**Training Process:**
1. Load training predictions from each base model (Q1-Q206)
2. Use `lsei()` to find optimal combination weights
3. Minimize: ||actual_returns - weighted_predictions||²
4. Subject to: sum(weights) = 1, weights ≥ 0

**Testing Process:**
1. Apply learned weights to test predictions (Q207-Q258)
2. Generate ensemble forecast for each quarter
3. Evaluate performance against individual models

---

# 1. Load Training Predictions

## 1.1 Load Individual Model Predictions (Training Set)
```{r load-training-predictions}
cat("\n=== LOADING TRAINING PREDICTIONS FOR ENSEMBLE ===\n\n")

# Each model team provides their training predictions (Q1-Q206)
# File format: date, log_ret, [model]_forecast_ret

# Random Forest training predictions
rf_train <- read_csv("../data/rf_train_predictions.csv") %>%
  select(date, log_ret, rf_forecast_ret) %>%
  arrange(date)

cat("✓ Random Forest loaded:", nrow(rf_train), "observations\n")

# XGBoost training predictions
xgb_train <- read_csv("../data/xgb_train_predictions.csv") %>%
  select(date, log_ret, xgb_forecast_ret) %>%
  arrange(date)

cat("✓ XGBoost loaded:", nrow(xgb_train), "observations\n")

# Lasso training predictions
lasso_train <- read_csv("../data/lasso_train_predictions.csv") %>%
  select(date, lasso_forecast_ret) %>%
  arrange(date)

cat("✓ Lasso loaded:", nrow(lasso_train), "observations\n")

# Ridge training predictions
ridge_train <- read_csv("../data/ridge_train_predictions.csv") %>%
  select(date, ridge_forecast_ret) %>%
  arrange(date)

cat("✓ Ridge loaded:", nrow(ridge_train), "observations\n")

# PCR training predictions
pcr_train <- read_csv("../data/pcr_train_predictions.csv") %>%
  select(date, pcr_forecast_ret) %>%
  arrange(date)

cat("✓ PCR loaded:", nrow(pcr_train), "observations\n")

# Elastic Net training predictions
elastic_train <- read_csv("../data/elastic_train_predictions.csv") %>%
  select(date, elastic_forecast_ret) %>%
  arrange(date)

cat("✓ Elastic Net loaded:", nrow(elastic_train), "observations\n")

cat("\n=== ALL TRAINING PREDICTIONS LOADED ===\n")
cat("Date range:", min(rf_train$date), "to", max(rf_train$date), "\n\n")
```

## 1.2 Combine Training Predictions
```{r combine-training}
cat("=== COMBINING TRAINING PREDICTIONS ===\n\n")

# Merge all training predictions by date
train_ensemble <- rf_train %>%
  left_join(xgb_train %>% select(date, xgb_forecast_ret), by = "date") %>%
  left_join(lasso_train, by = "date") %>%
  left_join(ridge_train, by = "date") %>%
  left_join(pcr_train, by = "date") %>%
  left_join(elastic_train, by = "date")

# Remove any rows with missing predictions
train_ensemble <- train_ensemble %>% drop_na()

cat("Combined training dataset:\n")
cat("  Total observations:", nrow(train_ensemble), "\n")
cat("  Number of models:", ncol(train_ensemble) - 2, "\n")  # Subtract date and log_ret
cat("  Date range:", min(train_ensemble$date), "to", max(train_ensemble$date), "\n")
cat("\nColumns included:\n")
print(names(train_ensemble))

cat("\n=== SUMMARY STATISTICS (Training Set) ===\n")
summary(train_ensemble %>% select(-date))
```

## 1.3 Visualize Training Predictions
```{r plot-training-predictions, fig.width=12, fig.height=8}
# Reshape data for plotting
train_long <- train_ensemble %>%
  pivot_longer(
    cols = ends_with("_forecast_ret"),
    names_to = "Model",
    values_to = "Prediction"
  ) %>%
  mutate(Model = gsub("_forecast_ret", "", Model))

# Plot predictions vs actual
ggplot(train_long, aes(x = date)) +
  geom_line(aes(y = log_ret), color = "black", linewidth = 1, alpha = 0.5) +
  geom_line(aes(y = Prediction, color = Model), alpha = 0.7) +
  labs(
    title = "Training Set: Model Predictions vs Actual Returns",
    subtitle = paste("Period:", min(train_ensemble$date), "to", max(train_ensemble$date)),
    x = "Date",
    y = "Log Return",
    color = "Model"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )
```

---

# 2. Learn Ensemble Weights Using LSEI

## 2.1 Prepare Data for LSEI
```{r prepare-lsei-data}
cat("\n=== PREPARING DATA FOR LSEI ===\n\n")

# Create prediction matrix (X)
# Each column is a model's predictions
X_train <- train_ensemble %>%
  select(rf_forecast_ret, xgb_forecast_ret, lasso_forecast_ret, 
         ridge_forecast_ret, pcr_forecast_ret, elastic_forecast_ret) %>%
  as.matrix()

# Actual returns vector (y)
y_train <- train_ensemble$log_ret

# Get model information
n_models <- ncol(X_train)
model_names <- colnames(X_train)

cat("LSEI Problem Setup:\n")
cat("  Number of models:", n_models, "\n")
cat("  Model names:", paste(gsub("_forecast_ret", "", model_names), collapse = ", "), "\n")
cat("  Training observations:", nrow(X_train), "\n")
cat("  Target variable: Actual log returns (log_ret)\n\n")

cat("Objective:\n")
cat("  Minimize: ||y_actual - (w1*model1 + w2*model2 + ... + wn*modeln)||²\n\n")

cat("Constraints:\n")
cat("  1. Equality: sum(w1, w2, ..., wn) = 1 (weights sum to 100%)\n")
cat("  2. Inequality: w1, w2, ..., wn ≥ 0 (non-negative weights)\n\n")
```

## 2.2 Set Up Constraints
```{r setup-constraints}
cat("=== SETTING UP LSEI CONSTRAINTS ===\n\n")

# Equality constraint: sum(weights) = 1
# E * weights = f
# [1, 1, 1, ..., 1] * [w1, w2, ..., wn]' = 1
E <- matrix(1, nrow = 1, ncol = n_models)
f <- 1

cat("Equality constraint matrix (E):\n")
print(E)
cat("Equality constraint value (f):", f, "\n\n")

# Inequality constraint: weights >= 0
# G * weights >= h
# Identity matrix ensures each weight >= 0
G <- diag(n_models)
h <- rep(0, n_models)

cat("Inequality constraint matrix (G):\n")
print(G)
cat("Inequality constraint values (h):", h, "\n\n")
```

## 2.3 Solve for Optimal Weights
```{r solve-lsei}
cat("=== SOLVING CONSTRAINED LEAST SQUARES ===\n\n")

# Run LSEI optimization
# Minimize ||X*weights - y||^2 subject to E*weights=f, G*weights>=h
lsei_result <- lsei(A = X_train, B = y_train, E = E, F = f, G = G, H = h)

# Extract optimal weights
ensemble_weights <- lsei_result$X
names(ensemble_weights) <- model_names

cat("✓ LSEI optimization completed successfully!\n\n")

cat("=================================================\n")
cat("           OPTIMAL ENSEMBLE WEIGHTS              \n")
cat("=================================================\n\n")

# Display weights in a formatted table
weights_display <- data.frame(
  Model = gsub("_forecast_ret", "", model_names),
  Weight = ensemble_weights,
  Percentage = paste0(round(ensemble_weights * 100, 2), "%")
)

print(weights_display, row.names = FALSE)

cat("\n-------------------------------------------------\n")
cat("Sum of weights:", round(sum(ensemble_weights), 10), "\n")
cat("(Should be exactly 1.0)\n")
cat("=================================================\n\n")

# Check which models are most important
top_models <- weights_display %>%
  arrange(desc(Weight)) %>%
  head(3)

cat("Top 3 models:\n")
print(top_models, row.names = FALSE)
```

## 2.4 In-Sample Performance
```{r in-sample-performance}
cat("\n=== IN-SAMPLE PERFORMANCE EVALUATION ===\n\n")

# Generate ensemble predictions on training set
train_ensemble_pred <- as.vector(X_train %*% ensemble_weights)
train_ensemble$ensemble_forecast_ret <- train_ensemble_pred

# Calculate performance metrics
train_errors <- train_ensemble$log_ret - train_ensemble$ensemble_forecast_ret
train_mse <- mean(train_errors^2)
train_rmse <- sqrt(train_mse)
train_mae <- mean(abs(train_errors))
train_r2 <- cor(train_ensemble$log_ret, train_ensemble$ensemble_forecast_ret)^2

cat("ENSEMBLE Performance (Training Set):\n")
cat("  RMSE:", round(train_rmse, 6), "\n")
cat("  MAE :", round(train_mae, 6), "\n")
cat("  R²  :", round(train_r2, 4), "\n\n")

cat("=================================================\n")
cat("      COMPARISON WITH INDIVIDUAL MODELS         \n")
cat("=================================================\n")
cat(sprintf("%-25s   RMSE      MAE       R²\n", "Model"))
cat("-------------------------------------------------\n")

# Calculate metrics for each individual model
for (i in 1:n_models) {
  model_errors <- train_ensemble$log_ret - X_train[, i]
  model_rmse <- sqrt(mean(model_errors^2))
  model_mae <- mean(abs(model_errors))
  model_r2 <- cor(train_ensemble$log_ret, X_train[, i])^2
  
  model_name <- gsub("_forecast_ret", "", model_names[i])
  cat(sprintf("%-25s  %.6f  %.6f  %.4f\n", 
              model_name, model_rmse, model_mae, model_r2))
}

cat("-------------------------------------------------\n")
cat(sprintf("%-25s  %.6f  %.6f  %.4f  <-- ENSEMBLE\n", 
            "ENSEMBLE", train_rmse, train_mae, train_r2))
cat("=================================================\n\n")

# Calculate improvement over best individual model
individual_rmses <- sapply(1:n_models, function(i) {
  sqrt(mean((train_ensemble$log_ret - X_train[, i])^2))
})
best_individual_rmse <- min(individual_rmses)
improvement_pct <- (best_individual_rmse - train_rmse) / best_individual_rmse * 100

if (improvement_pct > 0) {
  cat(sprintf("✓ Ensemble improves RMSE by %.2f%% over best individual model\n", improvement_pct))
} else {
  cat(sprintf("Note: Ensemble RMSE is %.2f%% higher than best individual model\n", abs(improvement_pct)))
}
```

## 2.5 Visualize Ensemble Weights
```{r plot-weights, fig.width=10, fig.height=6}
# Create bar plot of ensemble weights
weights_df <- data.frame(
  Model = gsub("_forecast_ret", "", model_names),
  Weight = ensemble_weights
) %>%
  arrange(desc(Weight))

weights_df$Model <- factor(weights_df$Model, levels = weights_df$Model)

ggplot(weights_df, aes(x = Model, y = Weight, fill = Model)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.1f%%", Weight * 100)), 
            vjust = -0.5, size = 4, fontface = "bold") +
  scale_y_continuous(labels = percent_format(), 
                     limits = c(0, max(weights_df$Weight) * 1.15)) +
  labs(
    title = "Ensemble Model Weights (Learned via LSEI)",
    subtitle = "Constrained: Sum to 100%, Non-negative",
    x = "Model",
    y = "Weight"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11)
  )
```

---

# 3. Apply Ensemble to Test Set

## 3.1 Load Test Predictions
```{r load-test-predictions}
cat("\n=== LOADING TEST PREDICTIONS ===\n\n")

# Load test predictions from each model (Q207-Q258)
# These should match the test period used in your backtesting

# Random Forest test predictions
rf_test <- read_csv("../data/rf_evals.csv") %>%
  select(date, log_ret, rf_forecast_ret) %>%
  arrange(date)

cat("✓ Random Forest test loaded:", nrow(rf_test), "observations\n")

# XGBoost test predictions
xgb_test <- read_csv("../data/xgb_evals.csv") %>%
  select(date, xgb_forecast_ret) %>%
  arrange(date)

cat("✓ XGBoost test loaded:", nrow(xgb_test), "observations\n")

# Lasso test predictions
lasso_test <- read_csv("../data/lasso_predictions.csv") %>%
  select(date, lasso_forecast_ret) %>%
  arrange(date)

cat("✓ Lasso test loaded:", nrow(lasso_test), "observations\n")

# Ridge test predictions
ridge_test <- read_csv("../data/ridge_predictions.csv") %>%
  select(date, ridge_forecast_ret) %>%
  arrange(date)

cat("✓ Ridge test loaded:", nrow(ridge_test), "observations\n")

# PCR test predictions
pcr_test <- read_csv("../data/pcr_predictions.csv") %>%
  select(date, pcr_forecast_ret) %>%
  arrange(date)

cat("✓ PCR test loaded:", nrow(pcr_test), "observations\n")

# Elastic Net test predictions
elastic_test <- read_csv("../data/elastic_predictions.csv") %>%
  select(date, elastic_forecast_ret) %>%
  arrange(date)

cat("✓ Elastic Net test loaded:", nrow(elastic_test), "observations\n")

cat("\n=== ALL TEST PREDICTIONS LOADED ===\n")
cat("Date range:", min(rf_test$date), "to", max(rf_test$date), "\n\n")
```

## 3.2 Combine Test Predictions
```{r combine-test}
cat("=== COMBINING TEST PREDICTIONS ===\n\n")

# Merge all test predictions
test_ensemble <- rf_test %>%
  left_join(xgb_test, by = "date") %>%
  left_join(lasso_test, by = "date") %>%
  left_join(ridge_test, by = "date") %>%
  left_join(pcr_test, by = "date") %>%
  left_join(elastic_test, by = "date")

test_ensemble <- test_ensemble %>% drop_na()

cat("Combined test dataset:\n")
cat("  Observations:", nrow(test_ensemble), "\n")
cat("  Date range:", min(test_ensemble$date), "to", max(test_ensemble$date), "\n\n")
```

## 3.3 Generate Ensemble Predictions
```{r generate-ensemble-predictions}
cat("=== GENERATING ENSEMBLE PREDICTIONS ===\n\n")

# Create test prediction matrix
X_test <- test_ensemble %>%
  select(rf_forecast_ret, xgb_forecast_ret, lasso_forecast_ret, 
         ridge_forecast_ret, pcr_forecast_ret, elastic_forecast_ret) %>%
  as.matrix()

# Apply learned weights to generate ensemble predictions
test_ensemble_pred <- as.vector(X_test %*% ensemble_weights)
test_ensemble$ensemble_forecast_ret <- test_ensemble_pred

cat("✓ Ensemble predictions generated!\n")
cat("  Number of predictions:", length(test_ensemble_pred), "\n\n")

cat("Sample ensemble predictions:\n")
print(head(test_ensemble %>% select(date, log_ret, ensemble_forecast_ret), 10))
```

## 3.4 Out-of-Sample Performance
```{r out-of-sample-performance}
cat("\n=== OUT-OF-SAMPLE PERFORMANCE EVALUATION ===\n\n")

# Calculate performance metrics
test_errors <- test_ensemble$log_ret - test_ensemble$ensemble_forecast_ret
test_mse <- mean(test_errors^2)
test_rmse <- sqrt(test_mse)
test_mae <- mean(abs(test_errors))
test_r2 <- cor(test_ensemble$log_ret, test_ensemble$ensemble_forecast_ret)^2

cat("ENSEMBLE Performance (Test Set):\n")
cat("  RMSE:", round(test_rmse, 6), "\n")
cat("  MAE :", round(test_mae, 6), "\n")
cat("  R²  :", round(test_r2, 4), "\n\n")

cat("=================================================\n")
cat("      COMPARISON WITH INDIVIDUAL MODELS         \n")
cat("=================================================\n")
cat(sprintf("%-25s   RMSE      MAE       R²\n", "Model"))
cat("-------------------------------------------------\n")

# Calculate metrics for each individual model
for (i in 1:n_models) {
  model_errors <- test_ensemble$log_ret - X_test[, i]
  model_rmse <- sqrt(mean(model_errors^2))
  model_mae <- mean(abs(model_errors))
  model_r2 <- cor(test_ensemble$log_ret, X_test[, i])^2
  
  model_name <- gsub("_forecast_ret", "", model_names[i])
  cat(sprintf("%-25s  %.6f  %.6f  %.4f\n", 
              model_name, model_rmse, model_mae, model_r2))
}

cat("-------------------------------------------------\n")
cat(sprintf("%-25s  %.6f  %.6f  %.4f  <-- ENSEMBLE\n", 
            "ENSEMBLE", test_rmse, test_mae, test_r2))
cat("=================================================\n\n")

# Calculate improvement over best individual model
individual_rmses_test <- sapply(1:n_models, function(i) {
  sqrt(mean((test_ensemble$log_ret - X_test[, i])^2))
})
best_individual_rmse_test <- min(individual_rmses_test)
improvement_pct_test <- (best_individual_rmse_test - test_rmse) / best_individual_rmse_test * 100

if (improvement_pct_test > 0) {
  cat(sprintf("✓ Ensemble improves RMSE by %.2f%% over best individual model\n", improvement_pct_test))
} else {
  cat(sprintf("Note: Ensemble RMSE is %.2f%% higher than best individual model\n", abs(improvement_pct_test)))
}
```

## 3.5 Visualize Test Predictions
```{r plot-test-predictions, fig.width=12, fig.height=8}
# Plot actual vs ensemble predictions
ggplot(test_ensemble, aes(x = date)) +
  geom_line(aes(y = log_ret, color = "Actual"), linewidth = 1) +
  geom_line(aes(y = ensemble_forecast_ret, color = "Ensemble"), linewidth = 1, alpha = 0.8) +
  scale_color_manual(values = c("Actual" = "black", "Ensemble" = "red")) +
  labs(
    title = "Test Set: Ensemble Predictions vs Actual Returns",
    subtitle = paste("Period:", min(test_ensemble$date), "to", max(test_ensemble$date)),
    x = "Date",
    y = "Log Return",
    color = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )
```

## 3.6 Prediction Errors
```{r plot-errors, fig.width=12, fig.height=6}
# Plot prediction errors
test_ensemble$error <- test_ensemble$log_ret - test_ensemble$ensemble_forecast_ret

ggplot(test_ensemble, aes(x = date, y = error)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Ensemble Prediction Errors (Test Set)",
    subtitle = paste("Mean Error:", round(mean(test_ensemble$error), 6), 
                     "| Std Dev:", round(sd(test_ensemble$error), 6)),
    x = "Date",
    y = "Prediction Error"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))
```

---

# 4. Save Ensemble Predictions

## 4.1 Save for Backtesting
```{r save-predictions}
cat("\n=== SAVING ENSEMBLE PREDICTIONS ===\n\n")

# Save test ensemble predictions for use in strategy_backtest.Rmd
ensemble_output <- test_ensemble %>%
  select(date, log_ret, ensemble_forecast_ret)

write_csv(ensemble_output, "../data/ensemble_predictions.csv")

cat("✓ Ensemble predictions saved to: ../data/ensemble_predictions.csv\n")
cat("  Columns: date, log_ret, ensemble_forecast_ret\n")
cat("  Rows:", nrow(ensemble_output), "\n\n")

cat("=================================================\n")
cat("This file can now be loaded in strategy_backtest.Rmd\n")
cat("for tactical asset allocation backtesting!\n")
cat("=================================================\n")
```

## 4.2 Save Ensemble Weights
```{r save-weights}
# Save weights for documentation
weights_output <- data.frame(
  model = gsub("_forecast_ret", "", model_names),
  weight = ensemble_weights,
  percentage = ensemble_weights * 100
)

write_csv(weights_output, "../data/ensemble_weights.csv")

cat("\n✓ Ensemble weights saved to: ../data/ensemble_weights.csv\n\n")
```

---

# 5. Summary

## 5.1 Key Findings
```{r summary}
cat("\n=================================================\n")
cat("           ENSEMBLE MODEL SUMMARY                \n")
cat("=================================================\n\n")

cat("METHODOLOGY:\n")
cat("  - Technique: Stacking with LSEI (constrained least squares)\n")
cat("  - Models combined:", n_models, "\n")
cat("  - Training period:", min(train_ensemble$date), "to", max(train_ensemble$date), 
    "(", nrow(train_ensemble), "quarters)\n")
cat("  - Test period:", min(test_ensemble$date), "to", max(test_ensemble$date), 
    "(", nrow(test_ensemble), "quarters)\n\n")

cat("WEIGHTS LEARNED:\n")
for (i in 1:n_models) {
  model_name <- gsub("_forecast_ret", "", model_names[i])
  cat(sprintf("  %-20s: %5.1f%%\n", model_name, ensemble_weights[i] * 100))
}

cat("\nPERFORMANCE (RMSE):\n")
cat(sprintf("  Training set : %.6f\n", train_rmse))
cat(sprintf("  Test set     : %.6f\n", test_rmse))

if (improvement_pct_test > 0) {
  cat(sprintf("\n✓ SUCCESS: Ensemble outperforms best individual model by %.2f%%\n", 
              improvement_pct_test))
} else {
  cat(sprintf("\nNote: Best individual model outperforms ensemble by %.2f%%\n", 
              abs(improvement_pct_test)))
}

cat("\n=================================================\n")
cat("Next step: Use ensemble_predictions.csv in\n")
cat("           strategy_backtest.Rmd for backtesting!\n")
cat("=================================================\n")
```
