---
title: "Lasso Regression"
author: "marth"
output: html_document
date: "2025-11-10"
---

```{r setup, include=FALSE}

# SETTING UP LIBRARIES
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(Metrics)
library(forecast)
library(glmnet)
```

## Loading Datasets
```{r}

data = read.csv("../data/data_merged.csv")
data_mean = read.csv("../data/prevail_mean.csv")

data$date <- as.Date(data$date)
data_mean$date <- as.Date(data_mean$date)

# ============================================================================
# CREATE MULTI-HORIZON TARGET VARIABLES
# ============================================================================
# We want to forecast 3-month, 6-month, and 12-month ahead returns
# 6-month = sum of 2 consecutive quarters
# 12-month = sum of 4 consecutive quarters

data <- data %>%
  arrange(date) %>%
  mutate(
    # Create lead log returns for each horizon
    log_ret_lead_3m = lead(log_ret, 1),           # 3-month ahead (1 quarter)
    log_ret_lead_6m = lead(log_ret, 1) + lead(log_ret, 2),  # 6-month ahead (2 quarters)
    log_ret_lead_12m = lead(log_ret, 1) + lead(log_ret, 2) + lead(log_ret, 3) + lead(log_ret, 4)  # 12-month ahead (4 quarters)
  )

# ============================================================================
# CREATE LAGGED VARIABLES FOR MULTI-HORIZON TARGETS
# ============================================================================
# Create 5 lags for each horizon's returns

# Lags for 3-month returns
data <- data %>%
  mutate(
    log_ret_3m_L1 = lag(log_ret_lead_3m, 1),
    log_ret_3m_L2 = lag(log_ret_lead_3m, 2),
    log_ret_3m_L3 = lag(log_ret_lead_3m, 3),
    log_ret_3m_L4 = lag(log_ret_lead_3m, 4),
    log_ret_3m_L5 = lag(log_ret_lead_3m, 5)
  )

# ============================================================================
# SELECT FINAL FEATURES FOR MODELING
# ============================================================================

data <- data %>%
  select(
    date,
    log_ret,
    # Multi-horizon targets
    log_ret_lead_3m,
    log_ret_lead_6m,
    log_ret_lead_12m,
    # All lagged y's for 3-month horizon
    matches("^log_ret_3m_L"),
    # ALL lagged x variables (L2-L5 only to avoid leakage with 1-quarter lead)
    matches("_L[2-5]$")
  )

cat("\n=== Multi-Horizon Target Variables Created ===\n")
cat("✓ log_ret_lead_3m: 3-month ahead returns (1 quarter)\n")
cat("✓ log_ret_lead_6m: 6-month ahead returns (sum of 2 quarters)\n")
cat("✓ log_ret_lead_12m: 12-month ahead returns (sum of 4 quarters)\n")
cat("✓ 5 lags created for each horizon\n")

# ============================================================================
# DIAGNOSTIC: Show all available columns with lag variables
# ============================================================================

cat("\n=== All Columns in Data ===\n")
all_cols <- colnames(data)
cat("Total columns:", length(all_cols), "\n")
cat("Columns:", paste(all_cols, collapse = ", "), "\n")

cat("\n=== Lag Columns (containing _L2, _L3, _L4, or _L5) ===\n")
lag_cols <- grep("_L[2-5]$", all_cols, value = TRUE)
cat("Total lag variables (L2-L5):", length(lag_cols), "\n")
print(lag_cols)

cat("\n=== 3-month Horizon Lag Columns ===\n")
lag_3m <- grep("^log_ret_3m_L", all_cols, value = TRUE)
cat("Total 3m lag variables:", length(lag_3m), "\n")
print(lag_3m)

cat("\n=== FEATURE SUMMARY ===\n")
cat("log_ret_3m_L* (3m horizon lags):", length(lag_3m), "\n")
cat("_L[2-5]$ (all L2-L5 variables):", length(lag_cols), "\n")
cat("TOTAL FEATURES FOR MODELING:", length(lag_3m) + length(lag_cols), "\n")

```

## Helper Functions: BIC-based Lambda Selection

```{r}
# ============================================================================
# BIC-based Lambda Selection Functions (Kock & Callot)
# BIC(λ) = log(RSS(λ)) + (log(T)/T)·df(λ)
# ============================================================================

# Function 1: Compute degrees of freedom for Lasso Regression
# For Lasso: df(λ) ≈ number of non-zero coefficients (excluding intercept)
compute_lasso_df <- function(fit_object, lambda) {
  # Get coefficients at this lambda
  coefs <- coef(fit_object, s = lambda)
  # df = number of non-zero coefficients (excluding intercept)
  df <- sum(coefs[-1] != 0)
  return(df)
}

# Function 2: Fit Lasso Regression and compute BIC for each lambda
fit_lasso_bic <- function(X, y, alpha = 1, lambdas = NULL) {
  
  # ---- Input validation ----
  T <- nrow(X)
  
  if (length(y) != T) stop("X and y must have same number of rows")
  if (!is.matrix(X)) X <- as.matrix(X)
  
  # Fit glmnet (Lasso: alpha=1, standardize=TRUE, intercept=TRUE)
  fit <- glmnet(X, y, alpha = alpha, standardize = TRUE, intercept = TRUE, 
                lambda = lambdas, thresh = 1e-7)
  
  # ---- Compute RSS on training data ----
  # Get predictions for each lambda
  y_pred_matrix <- predict(fit, newx = X, s = fit$lambda)
  
  # RSS for each lambda
  RSS_vec <- colSums((y - y_pred_matrix)^2)
  
  # ---- Compute degrees of freedom for each lambda ----
  df_vec <- sapply(fit$lambda, function(lam) {
    compute_lasso_df(fit, lam)
  })
  
  # ---- Compute BIC: log(RSS) + (log(T)/T) * df ----
  bic_penalty <- log(T) / T
  BIC_vec <- log(RSS_vec) + bic_penalty * df_vec
  
  # ---- Find optimal lambda ----
  idx_best <- which.min(BIC_vec)
  lambda_best <- fit$lambda[idx_best]
  
  # ---- Create results table ----
  bic_tbl <- data.frame(
    lambda = fit$lambda,
    RSS = RSS_vec,
    df = df_vec,
    BIC = BIC_vec,
    is_best = (fit$lambda == lambda_best)
  )
  
  # ---- Return results ----
  return(list(
    lambda_best = lambda_best,
    bic_tbl = bic_tbl,
    fit = fit
  ))
}

# Function 3: Helper to select lambda by BIC
# Dispatches based on alpha (0=Ridge, 1=Lasso)
select_lambda_bic <- function(X, y, alpha = 1, lambdas = NULL) {
  
  if (is.null(lambdas)) {
    # Generate default lambda sequence (glmnet default behavior)
    lambdas <- 10^seq(5, -2, length.out = 100)
  }
  
  if (alpha == 0) {
    # Ridge Regression (not used here)
    stop("Use alpha=1 for Lasso. For Ridge, use ridge_regression.Rmd")
  } else if (alpha == 1) {
    # LASSO
    result <- fit_lasso_bic(X, y, alpha = 1, lambdas = lambdas)
  } else {
    stop("alpha must be 0 (Ridge) or 1 (LASSO)")
  }
  
  return(result)
}

```

## Build Lasso Models for Each Horizon

```{r}
# ---- Fit Lasso Regression Models for Each Horizon (3m, 6m, 12m) ----
# Each horizon gets its own independent BIC optimization

# Lambda sequence for BIC optimization
lambdas <- 10^seq(3, -1, length.out = 50)

# Store results for each horizon
lasso_results <- list()

# Loop through each horizon
for (horizon in c("3m", "6m", "12m")) {
  cat("\n========== HORIZON:", horizon, "==========\n")
  
  # Extract datasets for this horizon (horizon-specific features only)
  X_train_h <- get(paste0("X_train_", horizon))
  y_train_h <- get(paste0("y_train_", horizon))
  X_test_h <- get(paste0("X_test_", horizon))
  y_test_h <- get(paste0("y_test_", horizon))
  
  cat("Features for", horizon, "horizon:", ncol(X_train_h), "variables\n")
  cat("Sample features:", paste(colnames(X_train_h)[seq_len(min(3, ncol(X_train_h)))], collapse = ", "), "\n")
  
  # Fit Lasso regression with BIC lambda selection
  # BIC optimization is done INDEPENDENTLY for each horizon
  lasso_fit <- select_lambda_bic(X_train_h, y_train_h, 
                                 alpha = 1, 
                                 lambdas = lambdas)
  
  # Store the result
  lasso_results[[horizon]] <- lasso_fit
  
  # Extract optimal lambda and model
  lambda_opt <- lasso_fit$lambda_best
  fit_glmnet <- lasso_fit$fit
  
  cat("\nOptimal lambda (BIC) for", horizon, ":", lambda_opt, "\n")
  cat("Top 10 BIC values:\n")
  print(head(lasso_fit$bic_tbl, 10))
  
  # Make predictions on test set
  y_pred_test <- as.vector(predict(fit_glmnet, newx = X_test_h, s = lambda_opt))
  
  # Make predictions on training set
  y_pred_train <- as.vector(predict(fit_glmnet, newx = X_train_h, s = lambda_opt))
  
  # Compute test metrics (RMSE = sqrt(mean(squared errors)))
  rmse_test <- sqrt(mean((y_test_h - y_pred_test)^2))
  
  # Compute training metrics
  rmse_train <- sqrt(mean((y_train_h - y_pred_train)^2))
  
  # Display metrics
  cat("\nTest Set Metrics:\n")
  cat("  RMSE:", rmse_test, "\n")
  
  cat("\nTraining Set Metrics:\n")
  cat("  RMSE:", rmse_train, "\n")
  
  # Extract coefficients
  coefficients <- as.vector(coef(fit_glmnet, s = lambda_opt))
  
  # Create results dataframe for test predictions
  test_results <- data.frame(
    actual = y_test_h,
    predicted = y_pred_test,
    residual = y_test_h - y_pred_test,
    abs_error = abs(y_test_h - y_pred_test)
  )
  
  # Create results dataframe for training predictions
  train_results <- data.frame(
    actual = y_train_h,
    predicted = y_pred_train,
    residual = y_train_h - y_pred_train,
    abs_error = abs(y_train_h - y_pred_train)
  )
  
  # Create metrics summary
  metrics_summary <- data.frame(
    Horizon = horizon,
    Dataset = c("Training", "Test"),
    RMSE = c(rmse_train, rmse_test),
    Lambda_Optimal = lambda_opt,
    N_NonZero_Coefs = sum(coefficients[-1] != 0)  # Exclude intercept
  )
  
  # Store results and metrics in lists for later visualization
  lasso_results[[horizon]]$train_results <- train_results
  lasso_results[[horizon]]$test_results <- test_results
  lasso_results[[horizon]]$metrics_summary <- metrics_summary
  lasso_results[[horizon]]$train_data_dates <- train_data$date
  lasso_results[[horizon]]$test_data_dates <- test_data$date
  
  cat("\n")
}

cat("\n========== LASSO REGRESSION MODELS COMPLETE ==========\n")

# Create cross-horizon summary
cat("\nCross-Horizon Summary:\n")
summary_list <- lapply(c("3m", "6m", "12m"), function(horizon) {
  lasso_results[[horizon]]$metrics_summary
})
summary_table <- do.call(rbind, summary_list)
print(summary_table)

```

## Visualizations: Lasso Regression Results for Each Horizon

```{r}
# ---- Visualize Test Set: Forecasted vs Actual for Each Horizon ----

for (horizon in c("3m", "6m", "12m")) {
  
  test_results <- lasso_results[[horizon]]$test_results
  test_dates <- lasso_results[[horizon]]$test_data_dates
  
  # Add dates to results
  test_results$date <- test_dates
  
  metrics <- lasso_results[[horizon]]$metrics_summary
  test_metrics <- metrics[metrics$Dataset == "Test", ]
  
  # Line plot: Forecasted vs Actual for test set
  p <- ggplot(test_results, aes(x = date)) +
    geom_line(aes(y = actual, color = "Actual"), linewidth = 1) +
    geom_line(aes(y = predicted, color = "Lasso Forecast"), linewidth = 0.8, linetype = "dashed") +
    labs(
      title = paste("Lasso Regression:", toupper(horizon), "Horizon - Test Set"),
      x = "Date",
      y = "Log Return",
      color = "Series",
      subtitle = paste("RMSE =", round(test_metrics$RMSE, 4))
    ) +
    scale_color_manual(values = c("Actual" = "steelblue", "Lasso Forecast" = "coral")) +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  print(p)
}

```

```{r}
# ---- Visualize Training Set: Forecasted vs Actual for Each Horizon ----

for (horizon in c("3m", "6m", "12m")) {
  
  train_results <- lasso_results[[horizon]]$train_results
  train_dates <- lasso_results[[horizon]]$train_data_dates
  
  # Add dates to results
  train_results$date <- train_dates
  
  metrics <- lasso_results[[horizon]]$metrics_summary
  train_metrics <- metrics[metrics$Dataset == "Training", ]
  
  # Line plot: Forecasted vs Actual for training set
  p <- ggplot(train_results, aes(x = date)) +
    geom_line(aes(y = actual, color = "Actual"), linewidth = 1) +
    geom_line(aes(y = predicted, color = "Lasso Forecast"), linewidth = 0.8, linetype = "dashed") +
    labs(
      title = paste("Lasso Regression:", toupper(horizon), "Horizon - Training Set"),
      x = "Date",
      y = "Log Return",
      color = "Series",
      subtitle = paste("RMSE =", round(train_metrics$RMSE, 4))
    ) +
    scale_color_manual(values = c("Actual" = "steelblue", "Lasso Forecast" = "coral")) +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  print(p)
}

```

```{r}
# ---- Scatter Plots: Forecast Accuracy for Each Horizon (Test Set) ----

for (horizon in c("3m", "6m", "12m")) {
  
  test_results <- lasso_results[[horizon]]$test_results
  metrics <- lasso_results[[horizon]]$metrics_summary
  test_metrics <- metrics[metrics$Dataset == "Test", ]
  
  p <- ggplot(test_results, aes(x = actual, y = predicted)) +
    geom_point(alpha = 0.6, color = "steelblue", size = 2) +
    geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", linewidth = 1) +
    labs(
      title = paste("Lasso Regression:", toupper(horizon), "Horizon - Test Set Accuracy"),
      x = "Actual Return",
      y = "Forecasted Return",
      subtitle = paste("RMSE =", round(test_metrics$RMSE, 4))
    ) +
    theme_minimal() +
    coord_fixed(ratio = 1)
  
  print(p)
}

```

## Store Lasso Regression Results

```{r}
# ---- Combine Results Across All Horizons ----

cat("\n========== STORING LASSO REGRESSION RESULTS ==========\n")

# Get dates for train and test sets
train_dates <- train_data$date
test_dates <- test_data$date

# Initialize combined dataframes
lasso_train_combined <- data.frame(
  date = train_dates,
  actual_3m = y_train_3m,
  predicted_3m = lasso_results[["3m"]]$train_results$predicted,
  actual_6m = y_train_6m,
  predicted_6m = lasso_results[["6m"]]$train_results$predicted,
  actual_12m = y_train_12m,
  predicted_12m = lasso_results[["12m"]]$train_results$predicted
)

lasso_test_combined <- data.frame(
  date = test_dates,
  actual_3m = y_test_3m,
  predicted_3m = lasso_results[["3m"]]$test_results$predicted,
  actual_6m = y_test_6m,
  predicted_6m = lasso_results[["6m"]]$test_results$predicted,
  actual_12m = y_test_12m,
  predicted_12m = lasso_results[["12m"]]$test_results$predicted
)

# Save to CSV
write.csv(lasso_train_combined, "../data/lasso_train_predictions.csv", row.names = FALSE)
write.csv(lasso_test_combined, "../data/lasso_test_predictions.csv", row.names = FALSE)

cat("\n=== Results Saved ===\n")
cat("✓ Training predictions saved to: ../data/lasso_train_predictions.csv\n")
cat("✓ Test predictions saved to: ../data/lasso_test_predictions.csv\n")
cat("✓ Columns: date, actual_3m, predicted_3m, actual_6m, predicted_6m, actual_12m, predicted_12m\n")

# Display sample of results
cat("\n=== Training Set Results (First 5 rows) ===\n")
print(head(lasso_train_combined, 5))

cat("\n=== Test Set Results (First 5 rows) ===\n")
print(head(lasso_test_combined, 5))

```

```{r}
# ---- Scatter Plots: Forecast Accuracy for Each Horizon (Training Set) ----

for (horizon in c("3m", "6m", "12m")) {
  
  train_results <- lasso_results[[horizon]]$train_results
  metrics <- lasso_results[[horizon]]$metrics_summary
  train_metrics <- metrics[metrics$Dataset == "Training", ]
  
  p <- ggplot(train_results, aes(x = actual, y = predicted)) +
    geom_point(alpha = 0.6, color = "steelblue", size = 2) +
    geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", linewidth = 1) +
    labs(
      title = paste("Lasso Regression:", toupper(horizon), "Horizon - Training Set Accuracy"),
      x = "Actual Return",
      y = "Forecasted Return",
      subtitle = paste("RMSE =", round(train_metrics$RMSE, 4))
    ) +
    theme_minimal() +
    coord_fixed(ratio = 1)
  
  print(p)
}

```
cat("Number of non-zero coefficients:", sum(coefs_df$coefficient != 0), "\n")
print(coefs_df)

```

```{r}
# ---- Visualize coefficients ----

ggplot(coefs_df, aes(x = reorder(feature, coefficient), y = coefficient, fill = coefficient > 0)) +
  geom_col() +
  scale_fill_manual(values = c("steelblue", "coral"), guide = "none") +
  labs(
    title = "LASSO Regression Coefficients",
    x = "Feature",
    y = "Coefficient Value"
  ) +
  coord_flip() +
  theme_minimal()

```

```{r}
# ---- Visualize forecasted vs actual returns on test set ----

# Create a data frame for plotting
test_results <- data.frame(
  date = test_data$date,
  actual = y_test,
  forecasted = as.numeric(y_pred_test),
  obs_num = seq_along(y_test)
)

# Line plot: Forecasted vs Actual
ggplot(test_results, aes(x = date)) +
  geom_line(aes(y = actual, color = "Actual"), linewidth = 1) +
  geom_line(aes(y = forecasted, color = "LASSO Forecast"), linewidth = 0.8, linetype = "dashed") +
  labs(
    title = "LASSO Regression: Forecasted vs Actual Returns (Test Set)",
    x = "Date",
    y = "Log Return",
    color = "Series"
  ) +
  scale_color_manual(values = c("Actual" = "steelblue", "LASSO Forecast" = "coral")) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

```{r}
# ---- Scatter plot: Forecasted vs Actual ----

ggplot(test_results, aes(x = actual, y = forecasted)) +
  geom_point(alpha = 0.6, color = "steelblue", size = 2) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "LASSO Regression: Forecast Accuracy (Test Set)",
    x = "Actual Return",
    y = "Forecasted Return",
    subtitle = paste("R² =", round(r2_lasso, 4), "| RMSE =", round(rmse_lasso, 6))
  ) +
  theme_minimal() +
  coord_fixed(ratio = 1)

```

```{r}
# ---- Save LASSO Regression Predictions and Results ----

# Create results dataframe with TRAINING predictions
y_pred_train <- predict(fit, newx = X_train, s = lambda_opt)

lasso_train_predictions <- data.frame(
  date = train_data$date,
  lasso_actual = y_train,
  lasso_forecast = as.numeric(y_pred_train)
)

# Save training predictions to CSV
write.csv(lasso_train_predictions, "../data/lasso_train_predictions.csv", row.names = FALSE)

# Create results dataframe with TEST predictions
lasso_predictions <- test_results %>%
  select(date, actual, forecasted) %>%
  rename(
    lasso_forecast = forecasted,
    lasso_actual = actual
  )

# Save test predictions to CSV
write.csv(lasso_predictions, "../data/lasso_predictions.csv", row.names = FALSE)

# Save model evaluation metrics
lasso_metrics <- data.frame(
  model = "LASSO Regression",
  optimal_lambda = lasso_result$lambda_best,
  MAE = mae_lasso,
  RMSE = rmse_lasso,
  R2 = r2_lasso
)

write.csv(lasso_metrics, "../data/lasso_metrics.csv", row.names = FALSE)

# Save BIC table
write.csv(lasso_result$bic_tbl, "../data/lasso_bic_table.csv", row.names = FALSE)

cat("\n=== LASSO Regression Results Saved ===\n")
cat("✓ Training predictions saved to: ../data/lasso_train_predictions.csv\n")
cat("✓ Test predictions saved to: ../data/lasso_predictions.csv\n")
cat("✓ Metrics saved to: ../data/lasso_metrics.csv\n")
cat("✓ BIC table saved to: ../data/lasso_bic_table.csv\n")

```

```{r}
# ---- Visualize training set: Forecasted vs Actual ----

# Create a data frame for training set plotting
train_results <- data.frame(
  date = train_data$date,
  actual = y_train,
  forecasted = as.numeric(y_pred_train),
  obs_num = seq_along(y_train)
)

# Line plot: Forecasted vs Actual for training set
ggplot(train_results, aes(x = date)) +
  geom_line(aes(y = actual, color = "Actual"), linewidth = 1) +
  geom_line(aes(y = forecasted, color = "LASSO Forecast"), linewidth = 0.8, linetype = "dashed") +
  labs(
    title = "LASSO Regression: Forecasted vs Actual Returns (Training Set)",
    x = "Date",
    y = "Log Return",
    color = "Series"
  ) +
  scale_color_manual(values = c("Actual" = "steelblue", "LASSO Forecast" = "coral")) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

```{r}
# ---- Scatter plot: Training set Forecasted vs Actual ----

# Compute training metrics for display
mae_train <- mae(y_train, y_pred_train)
rmse_train <- rmse(y_train, y_pred_train)
r2_train <- cor(y_train, y_pred_train)^2

ggplot(train_results, aes(x = actual, y = forecasted)) +
  geom_point(alpha = 0.6, color = "steelblue", size = 2) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "LASSO Regression: Training Set Forecast Accuracy",
    x = "Actual Return",
    y = "Forecasted Return",
    subtitle = paste("R² =", round(r2_train, 4), "| RMSE =", round(rmse_train, 6))
  ) +
  theme_minimal() +
  coord_fixed(ratio = 1)

```
