---
title: "Lasso Regression Forecast"
author: "Nadhif"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Set working directory to project root so ../data paths work
if (exists("knitr")) {
  knitr::opts_knit$set(root.dir = normalizePath('..'))
}

# SETTING UP LIBRARIES
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(purrr)
library(zoo)
library(Metrics)
library(forecast)
library(glmnet)  # for Lasso regression
library(here)    # for robust file paths
```

## Loading Datasets
```{r}

# Use here() for robust paths (works in knit AND interactive sessions)
library(here)

data_path <- here("data", "data_merged.csv")

if (!file.exists(data_path)) {
  stop("Data file not found at: ", data_path, 
       "\nCurrent working directory: ", getwd())
}

data = read.csv(data_path)
data$date <- as.Date(data$date)

# Join with prevailing mean benchmark from previous analysis
# We'll compute it here for comparison
window_size <- 60

data <- data %>%
  mutate(
    mean_prevailing_ret = map_dbl(seq_along(log_ret), function(i) {
      if (i <= window_size) return(NA_real_)
      mean(log_ret[(i - window_size):(i - 1)], na.rm = TRUE)
    })
  )

```

## Lasso Regression Model

Lasso (L1-regularized regression) performs automatic feature selection by shrinking some coefficients to exactly zero. This is useful when we have many lagged predictors and want to identify which features are most relevant for forecasting returns.

```{r}
# Build Lasso feature matrix: log_ret + lagged predictors *_L1..*_L4
# We use lagged predictors to prevent look-ahead bias

lasso_features <- data %>%
  select(
    date,
    log_ret,
    matches("_L[1-4]$")
  )

# Drop rows missing any predictors
lasso_df <- lasso_features %>% drop_na()

set.seed(123)
window_size <- 60
min_history <- 60
n <- nrow(lasso_df)

lasso_pred <- rep(NA_real_, n)
lasso_coefs_list <- list()  # store coefficients for interpretation

for (t in seq(min_history + 1, n)) {
  
  # Rolling training window
  start_idx <- max(1, t - window_size)
  end_idx   <- t - 1
  
  train_data <- lasso_df[start_idx:end_idx, ]
  test_data  <- lasso_df[t, , drop = FALSE]
  
  X_train <- train_data %>% select(-date, -log_ret) %>% as.matrix()
  y_train <- train_data$log_ret
  
  X_test  <- test_data %>% select(-date, -log_ret) %>% as.matrix()
  
  # Fit Lasso with cross-validation to choose optimal lambda
  cv_lasso <- cv.glmnet(
    x = X_train,
    y = y_train,
    alpha = 1,           # alpha=1 is Lasso, alpha=0 is Ridge
    nfolds = 5,
    standardize = TRUE   # standardize predictors
  )
  
  # Use lambda.1se (more regularized, simpler model)
  lasso_model <- glmnet(
    x = X_train,
    y = y_train,
    alpha = 1,
    lambda = cv_lasso$lambda.1se,
    standardize = TRUE
  )
  
  # Store last model coefficients for interpretation
  if (t == n) {
    last_lasso_model <- lasso_model
    last_cv_lasso <- cv_lasso
    last_X_train <- X_train
  }
  
  # Predict
  lasso_pred[t] <- predict(lasso_model, newx = X_test, s = cv_lasso$lambda.1se)[1]
}

# Check Lasso predictions
cat("\n=== Lasso Predictions Diagnostic ===\n")
cat("First index with non-NA Lasso:", which(!is.na(lasso_pred))[1], "\n")
cat("First 70 Lasso values:\n")
for (i in 1:70) {
  if (i %% 10 == 1) cat("\n[",i,"-", min(i+9, 70), "]: ", sep="")
  cat(sprintf("%s ", if(is.na(lasso_pred[i])) "NA" else sprintf("%.3f", lasso_pred[i])))
}
cat("First 70 Lasso values:\n")
print(lasso_pred[1:70])

# Attach forecasts
lasso_df$lasso_forecast_ret <- lasso_pred

# Diagnostic: Check Lasso alignment
cat("\n=== Lasso Alignment Check ===\n")
cat("Lasso pred - First non-NA index:", min(which(!is.na(lasso_pred))), "\n")
cat("Lasso pred - Last NA index:", max(which(is.na(lasso_pred))), "\n")

```

## Post-Lasso Model

Post-Lasso re-estimates the model using only the features that Lasso selected (non-zero coefficients). This often improves predictions by eliminating the bias from the L1 penalty on the selected features, while still benefiting from feature selection.

```{r}
# First, identify which features are selected by Lasso
set.seed(123)
window_size <- 60
min_history <- 60
n <- nrow(lasso_df)

postlasso_pred <- rep(NA_real_, n)

# Track errors
postlasso_errors <- 0

for (t in seq(min_history + 1, n)) {
  
  # Rolling training window
  start_idx <- max(1, t - window_size)
  end_idx   <- t - 1
  
  train_data <- lasso_df[start_idx:end_idx, ]
  test_data  <- lasso_df[t, , drop = FALSE]
  
  X_train <- train_data %>% select(-date, -log_ret) %>% as.matrix()
  y_train <- train_data$log_ret
  
  X_test  <- test_data %>% select(-date, -log_ret) %>% as.matrix()
  
  # Step 1: Fit Lasso to identify selected features
  tryCatch({
    cv_lasso_temp <- cv.glmnet(
      x = X_train,
      y = y_train,
      alpha = 1,
      nfolds = 5,
      standardize = TRUE
    )
    
    lasso_model_temp <- glmnet(
      x = X_train,
      y = y_train,
      alpha = 1,
      lambda = cv_lasso_temp$lambda.1se,
      standardize = TRUE
    )
    
    # Get non-zero coefficients (selected features)
    coef_lasso_temp <- coef(lasso_model_temp, s = cv_lasso_temp$lambda.1se)
    selected_features <- which(coef_lasso_temp[-1] != 0)  # exclude intercept
    
    # Step 2: Re-estimate using only selected features with OLS
    if (length(selected_features) > 0) {
      # Subset - ensure we get a matrix even if only one feature selected
      X_train_selected <- X_train[, selected_features, drop = FALSE]
      X_test_selected  <- X_test[, selected_features, drop = FALSE]
      
      # Fit OLS on selected features
      postlasso_df_temp <- as.data.frame(cbind(y_train, X_train_selected))
      colnames(postlasso_df_temp)[1] <- "y"
      
      postlasso_model <- lm(y ~ ., data = postlasso_df_temp)
      
      # Predict for test point
      X_test_selected_df <- as.data.frame(X_test_selected)
      colnames(X_test_selected_df) <- colnames(X_train_selected)
      postlasso_pred[t] <- predict(postlasso_model, newdata = X_test_selected_df)
      
      # Store last model info
      if (t == n) {
        last_postlasso_model <- postlasso_model
        last_selected_features <- selected_features
        last_X_train_selected <- X_train_selected
      }
    } else {
      # If Lasso selected no features, use mean of training data
      postlasso_pred[t] <- mean(y_train)
    }
  }, error = function(e) {
    # On any error, just use the mean of training data as fallback
    postlasso_pred[t] <<- mean(y_train)
    postlasso_errors <<- postlasso_errors + 1
  })
}

# Check the predictions before attaching
cat("\n=== Diagnostic: postlasso_pred ===\n")
cat("Length:", length(postlasso_pred), "\n")
cat("Non-NA count:", sum(!is.na(postlasso_pred)), "\n")
cat("Total errors caught:", postlasso_errors, "\n")
if (sum(!is.na(postlasso_pred)) > 0) {
  cat("First non-NA index:", min(which(!is.na(postlasso_pred))), "\n")
  cat("Last NA index:", max(which(is.na(postlasso_pred))), "\n")
} else {
  cat("ALL postlasso_pred are NA!\n")
}

# Attach forecasts to lasso_df
lasso_df$postlasso_forecast_ret <- postlasso_pred

# Diagnostic: Check Post-Lasso alignment
cat("\n=== Post-Lasso Alignment Check ===\n")
cat("Post-Lasso pred - First non-NA index:", min(which(!is.na(postlasso_pred))), "\n")
cat("Post-Lasso pred - Last NA index:", max(which(is.na(postlasso_pred))), "\n")
cat("Post-Lasso pred - Total NAs:", sum(is.na(postlasso_pred)), "\n")

```

## Visualization: Lasso + Post-Lasso vs Prevailing Mean vs Actual

```{r}
# Join both Lasso and Post-Lasso forecasts to actual returns and prevailing mean benchmark
plot_lasso_df <- lasso_df %>%
  left_join(
    data %>% select(date, log_ret, mean_prevailing_ret),
    by = "date"
  )

# DON'T filter - ggplot will skip NAs automatically
# This way, both Lasso and Post-Lasso start at the same row (row 61)

cat("\n=== Before Visualization ===\n")
cat("plot_lasso_df rows:", nrow(plot_lasso_df), "\n")
cat("Rows with non-NA lasso_forecast_ret:", sum(!is.na(plot_lasso_df$lasso_forecast_ret)), "\n")
cat("Rows with non-NA postlasso_forecast_ret:", sum(!is.na(plot_lasso_df$postlasso_forecast_ret)), "\n")
cat("First non-NA lasso at row:", min(which(!is.na(plot_lasso_df$lasso_forecast_ret))), "Date:", format(plot_lasso_df$date[min(which(!is.na(plot_lasso_df$lasso_forecast_ret)))], "%Y-%m-%d"), "\n")
cat("First non-NA postlasso at row:", min(which(!is.na(plot_lasso_df$postlasso_forecast_ret))), "Date:", format(plot_lasso_df$date[min(which(!is.na(plot_lasso_df$postlasso_forecast_ret)))], "%Y-%m-%d"), "\n")

# Use BOTH - this way ggplot skips NAs automatically for each line independently
# This ensures Lasso and Post-Lasso lines both use same time period
plot_filtered <- plot_lasso_df %>%
  filter(!is.na(lasso_forecast_ret) & !is.na(postlasso_forecast_ret))

cat("\nAfter filtering for BOTH non-NA:\n")
cat("Filtered rows:", nrow(plot_filtered), "\n")
if (nrow(plot_filtered) > 0) {
  cat("Date range:", format(min(plot_filtered$date), "%Y-%m-%d"), "to", format(max(plot_filtered$date), "%Y-%m-%d"), "\n")
}

ggplot(plot_filtered, 
       aes(x = date)) +
  geom_line(aes(y = log_ret.x,                color = "Actual returns")) +
  geom_line(aes(y = mean_prevailing_ret,     color = "Prevailing-mean forecast")) +
  geom_line(aes(y = lasso_forecast_ret,      color = "Lasso forecast")) +
  geom_line(aes(y = postlasso_forecast_ret,  color = "Post-Lasso forecast"), linetype = "dashed") +
  labs(
    title = "S&P 500 Log Return Forecasts: Lasso + Post-Lasso vs Prevailing Mean vs Actual",
    x = "Year",
    y = "Log Return",
    color = ""
  ) +
  theme_minimal()


```

## Performance Metrics

```{r}

lasso_eval_df <- plot_lasso_df %>%
  filter(
    !is.na(log_ret.x),
    !is.na(lasso_forecast_ret),
    !is.na(postlasso_forecast_ret),
    !is.na(mean_prevailing_ret)
  ) %>%
  mutate(
    err_lasso = log_ret.x - lasso_forecast_ret,
    err_postlasso = log_ret.x - postlasso_forecast_ret,
    err_mean = log_ret.x - mean_prevailing_ret
  )

# Lasso Metrics
MAE_lasso  <- mae(lasso_eval_df$log_ret.x, lasso_eval_df$lasso_forecast_ret)
RMSE_lasso <- rmse(lasso_eval_df$log_ret.x, lasso_eval_df$lasso_forecast_ret)
R2_lasso   <- cor(lasso_eval_df$log_ret.x, lasso_eval_df$lasso_forecast_ret)^2

# Post-Lasso Metrics
MAE_postlasso  <- mae(lasso_eval_df$log_ret.x, lasso_eval_df$postlasso_forecast_ret)
RMSE_postlasso <- rmse(lasso_eval_df$log_ret.x, lasso_eval_df$postlasso_forecast_ret)
R2_postlasso   <- cor(lasso_eval_df$log_ret.x, lasso_eval_df$postlasso_forecast_ret)^2

cat("\n=== Lasso Regression Forecast Performance ===\n")
cat("MAE :", round(MAE_lasso, 6),
    "\nRMSE:", round(RMSE_lasso, 6),
    "\nR²  :", round(R2_lasso, 4))

cat("\n=== Post-Lasso Regression Forecast Performance ===\n")
cat("MAE :", round(MAE_postlasso, 6),
    "\nRMSE:", round(RMSE_postlasso, 6),
    "\nR²  :", round(R2_postlasso, 4))

# DM Test: Lasso vs Prevailing Mean
dm_test_lasso <- dm.test(
  e1 = lasso_eval_df$err_lasso,
  e2 = lasso_eval_df$err_mean,
  alternative = "less"
)

cat("\n=== DM Test: Lasso VS Prevailing Mean ===\n")
cat("DM Statistic:", round(as.numeric(dm_test_lasso$statistic), 3),
    "\nP-value:", signif(dm_test_lasso$p.value, 4), "\n")

# DM Test: Post-Lasso vs Prevailing Mean
dm_test_postlasso <- dm.test(
  e1 = lasso_eval_df$err_postlasso,
  e2 = lasso_eval_df$err_mean,
  alternative = "less"
)

cat("\n=== DM Test: Post-Lasso VS Prevailing Mean ===\n")
cat("DM Statistic:", round(as.numeric(dm_test_postlasso$statistic), 3),
    "\nP-value:", signif(dm_test_postlasso$p.value, 4), "\n")

# DM Test: Post-Lasso vs Lasso
dm_test_postlasso_vs_lasso <- dm.test(
  e1 = lasso_eval_df$err_postlasso,
  e2 = lasso_eval_df$err_lasso,
  alternative = "less"
)

cat("\n=== DM Test: Post-Lasso VS Lasso ===\n")
cat("DM Statistic:", round(as.numeric(dm_test_postlasso_vs_lasso$statistic), 3),
    "\nP-value:", signif(dm_test_postlasso_vs_lasso$p.value, 4), "\n")

if(dm_test_postlasso_vs_lasso$p.value < 0.05) {
  cat("\nConclusion: Post-Lasso significantly outperforms Lasso.\n")
} else {
  cat("\nConclusion: Post-Lasso does not significantly improve over Lasso.\n")
}
```

## Lasso Coefficient Plot (Feature Selection)

One of the key advantages of Lasso is automatic feature selection. Let's visualize which features have non-zero coefficients in the final model.

```{r}
# Extract coefficients from the last trained model
coef_matrix <- coef(last_lasso_model, s = last_cv_lasso$lambda.1se)
coef_df <- data.frame(
  Feature = rownames(coef_matrix),
  Coefficient = as.numeric(coef_matrix)
) %>%
  filter(Feature != "(Intercept)", Coefficient != 0) %>%
  arrange(desc(abs(Coefficient)))

if (nrow(coef_df) > 0) {
  # Plot non-zero coefficients
  ggplot(coef_df, aes(x = reorder(Feature, abs(Coefficient)), y = Coefficient)) +
    geom_col(aes(fill = Coefficient > 0)) +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "coral"),
                      labels = c("Negative", "Positive")) +
    labs(
      title = "Lasso: Non-Zero Coefficients (Selected Features)",
      x = "Feature",
      y = "Coefficient Value",
      fill = "Direction"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  cat("\n=== Number of Non-Zero Features Selected by Lasso ===\n")
  cat("Total features:", ncol(last_X_train), "\n")
  cat("Selected (non-zero):", nrow(coef_df), "\n")
  cat("Sparsity:", round(100 * (1 - nrow(coef_df) / ncol(last_X_train)), 1), "%\n")
} else {
  cat("\nNote: Lasso shrunk all coefficients to zero (very high regularization).\n")
  cat("Consider using lambda.min instead of lambda.1se for less aggressive shrinkage.\n")
}

```

## Cross-Validation Lambda Path

This plot shows how the model's cross-validation error changes with different levels of regularization (lambda).

```{r}
plot(last_cv_lasso, main = "Lasso Cross-Validation: MSE vs Log(Lambda)")
abline(v = log(last_cv_lasso$lambda.1se), col = "red", lty = 2)
abline(v = log(last_cv_lasso$lambda.min), col = "blue", lty = 2)
legend("topright", 
       legend = c("lambda.1se (used)", "lambda.min"), 
       col = c("red", "blue"), 
       lty = 2, 
       cex = 0.8)
```

The red line shows the lambda we used (lambda.1se), which gives a simpler, more regularized model. The blue line (lambda.min) minimizes CV error but may overfit slightly.

## Summary

- **Lasso regression** performs automatic feature selection through L1 regularization
- Uses rolling 60-period windows for realistic out-of-sample forecasting
- Cross-validation selects optimal regularization strength (lambda)
- Evaluated against the prevailing mean benchmark using MAE, RMSE, R², and Diebold-Mariano test
- Coefficient plot shows which lagged features are most predictive (non-zero coefficients)
