---
title: "RF and XGBoost"
author: "Dawn"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}

# SETTING UP LIBRARIES
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(Metrics)
library(forecast)
library(randomForest)
library(xgboost)
library(pdp)
```

## Loading Datasets
```{r}

data = read.csv("../data/data_merged.csv") %>%
  select(
    date,
    log_ret,
    matches("_L[1-4]$")       # only lagged features
  ) %>%
  select(-matches("price")) %>% 
  drop_na()

```

## Benchmark Model - Prevailing Mean 
Our first benchmark will be a prevailing mean of S&P500 returns
```{r}
data$date <- as.Date(data$date)

window_size <- 60

data <- data %>%
  mutate(
    mean_prevailing_ret = map_dbl(seq_along(log_ret), function(i) {
      if (i <= window_size) return(NA_real_)
      mean(log_ret[(i - window_size):(i - 1)], na.rm = TRUE)
    })
  )

# Remove first rows that have NA forecast
plot_df <- data %>% filter(!is.na(mean_prevailing_ret))

# Plot
ggplot(plot_df, aes(x = date)) +
  geom_line(aes(y = log_ret, color = "Actual S&P 500 return")) +
  geom_line(aes(y = mean_prevailing_ret, color = "Prevailing mean return forecast")) +
  labs(
    title = "S&P 500 Return Forecast: Prevailing Mean Benchmark",
    x = "Date",
    y = "Log Return",
    color = ""
  ) +
  theme_minimal()
```

##Prevailing Mean Metrics
```{r}
PM_df <- data %>%
  mutate(lag_ret = dplyr::lag(log_ret)
  ) %>%
  filter(!is.na(log_ret), !is.na(mean_prevailing_ret), !is.na(lag_ret)) %>%
  mutate(
    err_mean = log_ret - mean_prevailing_ret,  # prevailing-mean forecast errors
    err_rw   = log_ret - lag_ret               # random-walk benchmark errors
  )

# Metrics
MAE_mean  <- mae(PM_df$log_ret, PM_df$mean_prevailing_ret)
RMSE_mean <- rmse(PM_df$log_ret, PM_df$mean_prevailing_ret)
R2_mean   <- cor(PM_df$log_ret, PM_df$mean_prevailing_ret)^2

cat("\n=== Prevailing Mean Return Forecast ===\n")
cat("MAE :", round(MAE_mean, 6),
    "\nRMSE:", round(RMSE_mean, 6),
    "\nR²  :", round(R2_mean, 4), "\n")

# Diebold–Mariano Test (Prevailing Mean vs Random Walk)
dm_test_mean <- dm.test(
  e1 = PM_df$err_mean,
  e2 = PM_df$err_rw,
  alternative = "less",  # H1: prevailing mean has lower expected loss
  h = 1, power = 2
)

cat("\n=== DM Test: Prevailing Mean vs Lag-Return (Random Walk) ===\n")
cat("DM Statistic:", round(as.numeric(dm_test_mean$statistic), 3),
    "\nP-value:", signif(dm_test_mean$p.value, 4), "\n")

# === Conclusion Interpretation ===
if(dm_test_mean$p.value < 0.05) {
  cat("\nConclusion: The p-value < 0.05 → We reject the null hypothesis.\n")
  cat("The prevailing mean forecast has significantly lower forecast error\n")
  cat("than the random-walk (last return) benchmark.\n")
} else {
  cat("\nConclusion: The p-value >= 0.05 → We fail to reject the null hypothesis.\n")
  cat("The prevailing mean forecast does not significantly outperform\n")
  cat("the random-walk (last return) benchmark.\n")
}


```
## Random Forest with Hyperparameter Tuning

# ============================================================================
# MODELING APPROACH: 80/20 Train-Test Split w/ Recursive CV
# ============================================================================


```{r}
# --- STEP 1: Prepare data ---

rf_df <- data %>%
  arrange(date) %>% 
  select(-mean_prevailing_ret) %>%
  drop_na()

set.seed(123)

n <- nrow(rf_df)
train_size <- floor(0.8 * n)  # 204 quarters (training)
test_size <- n - train_size   # 52 quarters (testing)

cat("\n=== Data Split ===\n")
cat("Training:", train_size, "quarters\n")
cat("Testing :", test_size, "quarters\n")

train_data <- rf_df[1:train_size, ]
test_data  <- rf_df[(train_size + 1):n, ]

# --- STEP 2: Hyperparameter tuning  ---
# will conduct hyperparameter tuning to get a set of best params, then recursive CV using these

rf_param_grid <- expand.grid(
  mtry = c(5, 10, 20, 30, 40, 50),    # important: feature sampling 
  nodesize = c(1, 2, 3, 5),           # regularisation/tree complexity
  ntree = 200                         # fixed: given train size 165, we can afford 200 trees
)

train_tune_size <- floor(0.8 * train_size)
val_tune_size   <- train_size - train_tune_size

X_train_tune <- train_data[1:train_tune_size, ] %>% select(-date, -log_ret)
y_train_tune <- train_data$log_ret[1:train_tune_size]

X_val_tune <- train_data[(train_tune_size + 1):train_size, ] %>% select(-date, -log_ret)
y_val_tune <- train_data$log_ret[(train_tune_size + 1):train_size]

best_val_rmse <- Inf
best_params_rf <- NULL


for (i in 1:nrow(rf_param_grid)) {
  temp_rf <- randomForest(
    x = X_train_tune, y = y_train_tune,
    ntree = rf_param_grid$ntree[i],
    mtry = rf_param_grid$mtry[i],
    nodesize = rf_param_grid$nodesize[i]
  )
  
  val_preds <- predict(temp_rf, newdata = X_val_tune)
  val_rmse <- sqrt(mean((y_val_tune - val_preds)^2))
  
  if (val_rmse < best_val_rmse) {
    best_val_rmse <- val_rmse
    best_params_rf <- rf_param_grid[i, ]
  }
}

cat("\n=== Best RF Params ===\n")
cat("mtry     :", best_params_rf$mtry, "\n")
cat("nodesize :", best_params_rf$nodesize, "\n")
cat("Validation RMSE:", round(best_val_rmse, 5), "\n")

# --- STEP 3: Recursive CV ---
# will use an expanding window, loop over 

cat("\n=== Starting Recursive Forecast Evaluation ===\n")

rf_recursive_preds <- rep(NA_real_, test_size)

for (t in 1:test_size) {
  
  train_subset <- rf_df[1:(train_size + t - 1), ]
  test_point   <- rf_df[train_size + t, , drop = FALSE]
  
  X_train <- train_subset %>% select(-date, -log_ret)
  y_train <- train_subset$log_ret
  X_test  <- test_point  %>% select(-date, -log_ret)
  
  # train only on past data
  rf_model_final <- randomForest(
    x = X_train, y = y_train,
    ntree = best_params_rf$ntree,
    mtry = best_params_rf$mtry,
    nodesize = best_params_rf$nodesize
  )
  
  # Predict for current test observation
  rf_recursive_preds[t] <- predict(rf_model_final, newdata = X_test)

}

cat("Prediction complete!\n")

# Store forecasts in df
rf_df$rf_forecast_ret <- c(rep(NA, train_size), rf_recursive_preds)


```

```{r}
# Join RF forecasts to actual returns and prevailing mean benchmark
plot_rf_df <- rf_df %>%
  left_join(
    data %>% select(date, mean_prevailing_ret),
    by = "date"
  )

# ONLY PLOT TEST SET (last 52 quarters)

ggplot(plot_rf_df %>% tail(test_size), aes(x = date)) +
  geom_line(aes(y = log_ret,             color = "Actual returns")) +
  geom_line(aes(y = mean_prevailing_ret, color = "Prevailing-mean forecast")) +
  geom_line(aes(y = rf_forecast_ret,     color = "Random Forest forecast")) +
  labs(
    title = "S&P 500 Log Return Forecasts: RF vs Prevailing Mean vs Actual (TEST SET)",
    subtitle = paste("Test period:", test_size, "quarters (Q207-Q258)"),
    x = "Year",
    y = "Log Return",
    color = ""
  ) +
  theme_minimal()


```

```{r}
# EVALUATE ONLY ON TEST SET
rf_eval_df <- plot_rf_df %>%
  tail(test_size) %>%   # last 52 quarters = test set
  filter(
    !is.na(log_ret),
    !is.na(rf_forecast_ret),
    !is.na(mean_prevailing_ret)
  ) %>%
  mutate(
    err_rf   = log_ret - rf_forecast_ret,
    err_mean = log_ret - mean_prevailing_ret
  )

cat("\n=== TEST SET PERFORMANCE (Q207-Q258, n =", nrow(rf_eval_df), "quarters) ===\n\n")

# MAE, RMSE, R²
actual <- tail(rf_df$log_ret, test_size)
pred   <- rf_recursive_preds

MAE_rf  <- mean(abs(actual - pred))
RMSE_rf <- sqrt(mean((actual - pred)^2))
R2_rf   <- cor(actual, pred, use = "complete.obs")^2


cat("=== Random Forest Forecast Performance ===\n")
cat("MAE :", round(MAE_rf, 6),
    "\nRMSE:", round(RMSE_rf, 6),
    "\nR²  :", round(R2_rf, 4))

#DM Test
dm_test_rf <- dm.test(
  e1 = rf_eval_df$err_rf,
  e2 = rf_eval_df$err_mean,
  alternative = "less"   # H1: RF has lower expected error
)

cat("\n\n=== DM Test: Random Forest VS Prevailing Mean ===\n")
cat("DM Statistic:", round(as.numeric(dm_test_rf$statistic), 3),
    "\nP-value:", signif(dm_test_rf$p.value, 4), "\n")

# === Conclusion Interpretation ===
if(dm_test_rf$p.value < 0.05) {
  cat("\nConclusion: The p-value < 0.05 → We reject the null hypothesis.\n")
  cat("Random Forest has significantly lower forecast error\n")
  cat("than the Prevailing Mean model.\n")
} else {
  cat("\nConclusion: The p-value >= 0.05 → We fail to reject the null hypothesis.\n")
  cat("Random Forest does not significantly outperform the Prevailing Mean Model.\n")
}
```

##Feature Importance Plot
```{r}
# Extract importance values from final RF model 
imp <- importance(rf_model_final)

# Convert to clean dataframe
imp_df <- data.frame(
  Feature = rownames(imp),
  Importance = imp[, 1]   # %IncMSE for regression
)

# Select Top 10 features to display in the plot
imp_top10 <- imp_df %>%
  arrange(desc(Importance)) %>%
  slice_head(n = 10)

ggplot(imp_top10, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Feature Importance (%IncMSE)",
    x = "Predictor (Lagged Features)",
    y = "Importance"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(face = "bold")
  )

```
##PDP Plot
```{r}
top3 <- imp_top10$Feature[1:3]

for (feat in top3) {
  pd <- partial(
    rf_model_final,
    pred.var = feat,
    train = train_data %>% select(-date, -log_ret),
    grid.resolution = 40
  )
  
  p <- autoplot(pd) +
    labs(
      title = paste("Partial Dependence Plot:", feat),
      x = feat,
      y = "Effect on Predicted Return"
    ) +
    theme_minimal()
  
  print(p)
}

```
##OOB Error
```{r}
plot(rf_model_final, main = "OOB Error vs Number of Trees (Random Forest)")
```
##From the plot, we see that Eeror drops quickly in the first ~50 trees, and the curve flattens and stabilizes around 100 trees. Afteraround 100 trees, the OOB error does not meaningfully decrease anymor (the model is basically just averaging noise (flat tail)). So, adding more trees beyond ~150 does not improve accuracy. Hence, we conclude that we could have choosen about 150 trees for a stable and low variance random forest model.

##XGBoost with Hyperparameter Tuning
```{r}

# Use same train/test split as Random Forest
# train_data and test_data already defined (206 / 52 split)

xgb_df <- data %>%
  arrange(date) %>% 
  select(-mean_prevailing_ret) %>%
  drop_na()

set.seed(123)

n <- nrow(xgb_df)
train_size <- floor(0.8 * n)  # 206 quarters
test_size  <- n - train_size  # 52 quarters

train_data <- xgb_df[1:train_size, ]
test_data  <- xgb_df[(train_size + 1):n, ]

cat("\n=== Data Split ===\n")
cat("Training set:", train_size, "quarters\n")
cat("Test set    :", test_size, "quarters\n\n")

# --- STEP 2: Hyperparameter tuning ---

train_tune_size <- floor(0.8 * train_size)  # 165 quarters
val_tune_size   <- train_size - train_tune_size  # 41 quarters

X_train_tune <- train_data[1:train_tune_size, ] %>% select(-date, -log_ret)
y_train_tune <- train_data$log_ret[1:train_tune_size]

X_val_tune   <- train_data[(train_tune_size + 1):train_size, ] %>% select(-date, -log_ret)
y_val_tune   <- train_data$log_ret[(train_tune_size + 1):train_size]

dtrain_tune <- xgb.DMatrix(data = as.matrix(X_train_tune), label = y_train_tune)
dval_tune   <- xgb.DMatrix(data = as.matrix(X_val_tune),   label = y_val_tune)

xgb_param_grid <- expand.grid(
  max_depth        = c(3, 4, 5, 6),
  eta              = c(0.01, 0.05, 0.1),
  nrounds          = c(100, 200, 300),
  subsample        = c(0.7, 0.8, 0.9),
  colsample_bytree = c(0.6, 0.7, 0.8)
)

best_val_rmse_xgb <- Inf
best_params_xgb   <- NULL
best_nrounds      <- NULL

cat("\n=== XGBoost Hyperparameter Tuning ===\n")

for (i in 1:nrow(xgb_param_grid)) {
  params <- list(
    objective        = "reg:squarederror",
    max_depth        = xgb_param_grid$max_depth[i],
    eta              = xgb_param_grid$eta[i],
    subsample        = xgb_param_grid$subsample[i],
    colsample_bytree = xgb_param_grid$colsample_bytree[i]
  )
  
  model_temp <- xgb.train(
    params = params,
    data   = dtrain_tune,
    nrounds = xgb_param_grid$nrounds[i],
    watchlist = list(train = dtrain_tune, val = dval_tune),
    verbose = 0,
    early_stopping_rounds = 20
  )
  
  preds_val <- predict(model_temp, dval_tune)
  rmse_val  <- sqrt(mean((y_val_tune - preds_val)^2))
  
  if (rmse_val < best_val_rmse_xgb) {
    best_val_rmse_xgb <- rmse_val
    best_params_xgb   <- xgb_param_grid[i, ]
    best_nrounds      <- model_temp$best_iteration
  }
}

cat("\n=== Best XGBoost Parameters ===\n")
cat("Best Val RMSE:", round(best_val_rmse_xgb, 6), "\n\n")

# --- STEP 3: Recursive CV ---

cat("=== Starting Recursive XGBoost Forecasting ===\n")

xgb_recursive_preds <- rep(NA_real_, test_size)

for (t in 1:test_size) {
  # expanding window up to current test quarter
  train_subset <- xgb_df[1:(train_size + t - 1), ]
  test_point   <- xgb_df[train_size + t, , drop = FALSE]
  
  X_train <- train_subset %>% select(-date, -log_ret)
  y_train <- train_subset$log_ret
  X_test  <- test_point  %>% select(-date, -log_ret)
  
  dtrain <- xgb.DMatrix(data = as.matrix(X_train), label = y_train)
  dtest  <- xgb.DMatrix(data = as.matrix(X_test))
  
  params_final <- list(
    objective        = "reg:squarederror",
    max_depth        = best_params_xgb$max_depth,
    eta              = best_params_xgb$eta,
    subsample        = best_params_xgb$subsample,
    colsample_bytree = best_params_xgb$colsample_bytree
  )
  
  xgb_model_final <- xgb.train(
    params  = params_final,
    data    = dtrain,
    nrounds = best_nrounds,
    verbose = 0
  )
  
  xgb_recursive_preds[t] <- predict(xgb_model_final, dtest)
  
}


cat("Prediction complete!\n")

xgb_df$xgb_forecast_ret <- c(rep(NA, train_size), xgb_recursive_preds)

```


## Plot XGBoost model against actual and prevailing mean model 
```{r}

plot_xgb_df <- xgb_df %>%
  left_join(
    data %>% select(date, mean_prevailing_ret),
    by = "date"
  )

# ONLY PLOT TEST SET (last 52 quarters)

ggplot(plot_xgb_df %>% tail(test_size), aes(x = date)) +
  geom_line(aes(y = log_ret,            color = "Actual returns")) +
  geom_line(aes(y = mean_prevailing_ret, color = "Prevailing-mean forecast")) +
  geom_line(aes(y = xgb_forecast_ret,    color = "XGBoost forecast")) +
  labs(
    title = "S&P 500 Log Return Forecasts: XGBoost vs Prevailing Mean vs Actual",
    subtitle = paste("Test period:", test_size, "quarters (Q207-Q258)"),
    x = "Year",
    y = "Log Return",
    color = ""
  ) +
  theme_minimal()
```

```{r}
# EVALUATE ONLY ON TEST SET
xgb_eval_df <- plot_xgb_df %>%
  tail(test_size) %>%  # Only last 52 quarters (test set)
  filter(
    !is.na(log_ret),
    !is.na(xgb_forecast_ret),
    !is.na(mean_prevailing_ret)
  ) %>%
  mutate(
    err_xgb   = log_ret - xgb_forecast_ret,
    err_mean = log_ret - mean_prevailing_ret
  )

cat("\n=== TEST SET PERFORMANCE (Q207-Q258, n =", nrow(xgb_eval_df), "quarters) ===\n\n")

# MAE, RMSE, R²
pred   <- xgb_recursive_preds

MAE_xgb  <- mean(abs(actual - pred))
RMSE_xgb <- sqrt(mean((actual - pred)^2))
R2_xgb   <- cor(actual, pred, use = "complete.obs")^2


cat("=== XGBoost Forecast Performance ===\n")
cat("MAE :", round(MAE_xgb, 6),
    "\nRMSE:", round(RMSE_xgb, 6),
    "\nR²  :", round(R2_xgb, 4))

#DM Test
dm_test_xgb <- dm.test(
  e1 = xgb_eval_df$err_xgb,
  e2 = xgb_eval_df$err_mean,
  alternative = "less"   # H1: XGBoost has lower expected error
)

cat("\n\n=== DM Test: XGBoost VS Prevailing Mean ===\n")
cat("DM Statistic:", round(as.numeric(dm_test_xgb$statistic), 3),
    "\nP-value:", signif(dm_test_xgb$p.value, 4), "\n")

# === Conclusion Interpretation ===
if(dm_test_xgb$p.value < 0.05) {
  cat("\nConclusion: The p-value < 0.05 → We reject the null hypothesis.\n")
  cat("XGBoost has significantly lower forecast error\n")
  cat("than the Prevailing Mean model.\n")
} else {
  cat("\nConclusion: The p-value >= 0.05 → We fail to reject the null hypothesis.\n")
  cat("XGBoost does not significantly outperform the Prevailing Mean Model.\n")
}
```

##Feature Importance
```{r}
# Feature importance (Top 10 by Gain)
imp_tbl <- xgb.importance(model = xgb_model_final, 
                          feature_names = colnames(train_data %>% select(-date, -log_ret)))

imp_top10 <- imp_tbl %>%
  dplyr::slice_max(order_by = Gain, n = 10) %>%
  dplyr::mutate(Feature = factor(Feature, levels = rev(Feature)))

ggplot(imp_top10, aes(x = Feature, y = Gain)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "XGBoost Feature Importance (Top 10)",
       x = "Feature", y = "Gain") +
  theme_minimal()

```

## Model Comparison: Test Set Performance
```{r}
# ============================================================================
# COMPREHENSIVE MODEL COMPARISON ON TEST SET
# ============================================================================
# Compare all three models on the same test period (Q207-Q258, 52 quarters)
# This ensures a fair apples-to-apples comparison

cat("  COMPREHENSIVE MODEL COMPARISON: TEST SET PERFORMANCE\n")
cat("  Test Period: Q207-Q258 (52 quarters, ~13 years)\n")
cat("  Training Period: Q1-Q206 (206 quarters for ML models)\n")


# Prepare comparison dataframe with all three models on test set
comparison_df <- rf_eval_df %>%
  select(date, log_ret, mean_prevailing_ret, rf_forecast_ret, err_rf) %>%
  left_join(
    xgb_eval_df %>%
      select(date, xgb_forecast_ret, err_xgb),
    by = "date"
  ) %>%
  filter(
    !is.na(log_ret),
    !is.na(mean_prevailing_ret),
    !is.na(rf_forecast_ret),
    !is.na(xgb_forecast_ret)
  ) %>%
  mutate(err_prevailing = log_ret - mean_prevailing_ret
         )
  

# ========== PERFORMANCE METRICS TABLE ==========
# Calculate all metrics
metrics_summary <- data.frame(
  Model = c("Prevailing Mean (Benchmark)", "Random Forest (Tuned)", "XGBoost (Tuned)"),
  
  MAE = c(
    MAE_mean,
    MAE_rf,
    MAE_xgb),
  
  RMSE = c(
    RMSE_mean,
    RMSE_rf,
    RMSE_xgb),
  
  R_squared = c(
    R2_mean,
    R2_rf,
    R2_xgb)
)

# Print formatted table
cat("PERFORMANCE METRICS (Test Set: n =", nrow(comparison_df), "quarters)\n")
cat(rep("-", 80), "\n")
print(metrics_summary, row.names = FALSE, digits = 6)
cat(rep("-", 80), "\n\n")

# ========== IDENTIFY BEST MODEL ==========
best_mae_idx <- which.min(metrics_summary$MAE)
best_rmse_idx <- which.min(metrics_summary$RMSE)
best_r2_idx <- which.max(metrics_summary$R_squared)

cat("BEST PERFORMANCE:\n")
cat("  Lowest MAE  :", metrics_summary$Model[best_mae_idx], 
    "(", round(metrics_summary$MAE[best_mae_idx], 6), ")\n")
cat("  Lowest RMSE :", metrics_summary$Model[best_rmse_idx], 
    "(", round(metrics_summary$RMSE[best_rmse_idx], 6), ")\n")
cat("  Highest R²  :", metrics_summary$Model[best_r2_idx], 
    "(", round(metrics_summary$R_squared[best_r2_idx], 4), ")\n\n")

# ========== RELATIVE IMPROVEMENT OVER BENCHMARK ==========
cat("IMPROVEMENT OVER PREVAILING MEAN BENCHMARK:\n")
cat(rep("-", 80), "\n")

rf_mae_improvement <- (metrics_summary$MAE[1] - metrics_summary$MAE[2]) / metrics_summary$MAE[1] * 100
xgb_mae_improvement <- (metrics_summary$MAE[1] - metrics_summary$MAE[3]) / metrics_summary$MAE[1] * 100

rf_rmse_improvement <- (metrics_summary$RMSE[1] - metrics_summary$RMSE[2]) / metrics_summary$RMSE[1] * 100
xgb_rmse_improvement <- (metrics_summary$RMSE[1] - metrics_summary$RMSE[3]) / metrics_summary$RMSE[1] * 100

cat("Random Forest:\n")
cat("  MAE improvement : ", sprintf("%+.2f%%", rf_mae_improvement), 
    ifelse(rf_mae_improvement > 0, " (better)", " (worse)"), "\n")
cat("  RMSE improvement: ", sprintf("%+.2f%%", rf_rmse_improvement), 
    ifelse(rf_rmse_improvement > 0, " (better)", " (worse)"), "\n\n")

cat("XGBoost:\n")
cat("  MAE improvement : ", sprintf("%+.2f%%", xgb_mae_improvement), 
    ifelse(xgb_mae_improvement > 0, " (better)", " (worse)"), "\n")
cat("  RMSE improvement: ", sprintf("%+.2f%%", xgb_rmse_improvement), 
    ifelse(xgb_rmse_improvement > 0, " (better)", " (worse)"), "\n\n")

# ========== DIEBOLD-MARIANO TESTS ==========
cat("DIEBOLD-MARIANO STATISTICAL TESTS:\n")
cat(rep("-", 80), "\n")

# Test 1: RF vs Prevailing Mean

cat("1. Random Forest vs Prevailing Mean:\n")
cat("   DM Statistic:", round(dm_test_rf$statistic, 4), "\n")
cat("   P-value     :", signif(dm_test_rf$p.value, 4), "\n")
cat("   Result      :", 
    ifelse(dm_test_rf$p.value < 0.05, 
           "RF significantly outperforms Prevailing Mean ✓",
           "No significant difference"),
    "\n\n")

# Test 2: XGBoost vs Prevailing Mean

cat("2. XGBoost vs Prevailing Mean:\n")
cat("   DM Statistic:", round(dm_test_xgb$statistic, 4), "\n")
cat("   P-value     :", signif(dm_test_xgb$p.value, 4), "\n")
cat("   Result      :", 
    ifelse(dm_test_xgb$p.value < 0.05, 
           "XGBoost significantly outperforms Prevailing Mean ✓",
           "No significant difference"),
    "\n\n")

# Test 3: RF vs XGBoost
dm_rf_vs_xgb <- dm.test(
  e1 = comparison_df$err_rf,
  e2 = comparison_df$err_xgb,
  alternative = "two.sided",  # H1: Different performance
  h = 1, power = 2
)

cat("3. Random Forest vs XGBoost:\n")
cat("   DM Statistic:", round(dm_rf_vs_xgb$statistic, 4), "\n")
cat("   P-value     :", signif(dm_rf_vs_xgb$p.value, 4), "\n")
cat("   Result      :", 
    ifelse(dm_rf_vs_xgb$p.value < 0.05, 
           ifelse(dm_rf_vs_xgb$statistic < 0,
                  "RF significantly outperforms XGBoost ✓",
                  "XGBoost significantly outperforms RF ✓"),
           "No significant difference between RF and XGBoost"),
    "\n\n")

# ========== VISUAL COMPARISON ==========
cat(rep("=", 80), "\n")
cat("VISUAL COMPARISON PLOT GENERATED BELOW\n")
cat(rep("=", 80), "\n\n")

# Reshape data for plotting
comparison_long <- comparison_df %>%
  select(date, log_ret, mean_prevailing_ret, rf_forecast_ret, xgb_forecast_ret) %>%
  rename(
    Actual = log_ret,
    `Prevailing Mean` = mean_prevailing_ret,
    `Random Forest` = rf_forecast_ret,
    XGBoost = xgb_forecast_ret
  ) %>%
  pivot_longer(
    cols = -date,
    names_to = "Model",
    values_to = "Return"
  ) %>%
  mutate(
    Model = factor(Model, levels = c("Actual", "Prevailing Mean", "Random Forest", "XGBoost"))
  )

# Time series comparison plot
ggplot(comparison_long, aes(x = date, y = Return, color = Model)) +
  geom_line(aes(linetype = Model), linewidth = 0.7) +
  scale_linetype_manual(values = c("solid", "dashed", "solid", "solid")) +
  scale_color_manual(values = c("black", "blue", "red", "darkgreen")) +
  labs(
    title = "Model Comparison: Test Set Forecasts (Q207-Q258)",
    subtitle = paste("Test period:", test_size, "quarters | All models evaluated on same out-of-sample data"),
    x = "Date",
    y = "Log Return",
    color = "Model",
    linetype = "Model"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10)
  )

```

## Error Distribution Comparison
```{r}
# Compare error distributions across models
error_comparison <- comparison_df %>%
  select(date, err_prevailing, err_rf, err_xgb) %>%
  rename(
    `Prevailing Mean` = err_prevailing,
    `Random Forest` = err_rf,
    XGBoost = err_xgb
  ) %>%
  pivot_longer(
    cols = -date,
    names_to = "Model",
    values_to = "Error"
  )

# Boxplot comparison
ggplot(error_comparison, aes(x = Model, y = Error, fill = Model)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  scale_fill_manual(values = c("blue", "red", "darkgreen")) +
  labs(
    title = "Forecast Error Distribution Comparison (Test Set)",
    subtitle = "Lower spread and median closer to zero indicates better performance",
    x = "Model",
    y = "Forecast Error (Actual - Predicted)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14)
  )

# Summary statistics of errors
cat("\nFORECAST ERROR STATISTICS:\n")
cat(rep("-", 80), "\n")

error_stats <- error_comparison %>%
  group_by(Model) %>%
  summarise(
    Mean = mean(Error),
    Median = median(Error),
    SD = sd(Error),
    Min = min(Error),
    Max = max(Error),
    IQR = IQR(Error)
  )

print(error_stats, digits = 6)

```

