---
title: "RF and XGBoost"
author: "Dawn"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}

# SETTING UP LIBRARIES
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)
library(purrr)
library(zoo)
library(Metrics)
library(forecast)
library(randomForest)
library(xgboost)
library(pdp)
```

## Loading Datasets
```{r}

data = read.csv("../data/data_merged.csv")

```

## Benchmark Model - Prevailing Mean 
Our first benchmark will be a prevailing mean of S&P500 returns
```{r}
data$date <- as.Date(data$date)

window_size <- 60

data <- data %>%
  mutate(
    mean_prevailing_ret = map_dbl(seq_along(log_ret), function(i) {
      if (i <= window_size) return(NA_real_)
      mean(log_ret[(i - window_size):(i - 1)], na.rm = TRUE)
    })
  )

# Remove first rows that have NA forecast
plot_df <- data %>% filter(!is.na(mean_prevailing_ret))

# Plot
ggplot(plot_df, aes(x = date)) +
  geom_line(aes(y = log_ret, color = "Actual S&P 500 return")) +
  geom_line(aes(y = mean_prevailing_ret, color = "Prevailing mean return forecast")) +
  labs(
    title = "S&P 500 Return Forecast: Prevailing Mean Benchmark",
    x = "Date",
    y = "Log Return",
    color = ""
  ) +
  theme_minimal()
```

##Prevailing Mean Metrics
```{r}
PM_df <- data %>%
  mutate(lag_ret = dplyr::lag(log_ret)
  ) %>%
  filter(!is.na(log_ret), !is.na(mean_prevailing_ret), !is.na(lag_ret)) %>%
  mutate(
    err_mean = log_ret - mean_prevailing_ret,  # prevailing-mean forecast errors
    err_rw   = log_ret - lag_ret               # random-walk benchmark errors
  )

# Metrics
MAE_mean  <- mae(PM_df$log_ret, PM_df$mean_prevailing_ret)
RMSE_mean <- rmse(PM_df$log_ret, PM_df$mean_prevailing_ret)
R2_mean   <- cor(PM_df$log_ret, PM_df$mean_prevailing_ret)^2

cat("\n=== Prevailing Mean Return Forecast ===\n")
cat("MAE :", round(MAE_mean, 6),
    "\nRMSE:", round(RMSE_mean, 6),
    "\nR²  :", round(R2_mean, 4), "\n")

# Diebold–Mariano Test (Prevailing Mean vs Random Walk)
dm_test_mean <- dm.test(
  e1 = PM_df$err_mean,
  e2 = PM_df$err_rw,
  alternative = "less",  # H1: prevailing mean has lower expected loss
  h = 1, power = 2
)

cat("\n=== DM Test: Prevailing Mean vs Lag-Return (Random Walk) ===\n")
cat("DM Statistic:", round(as.numeric(dm_test_mean$statistic), 3),
    "\nP-value:", signif(dm_test_mean$p.value, 4), "\n")

# === Conclusion Interpretation ===
if(dm_test_mean$p.value < 0.05) {
  cat("\nConclusion: The p-value < 0.05 → We reject the null hypothesis.\n")
  cat("The prevailing mean forecast has significantly lower forecast error\n")
  cat("than the random-walk (last return) benchmark.\n")
} else {
  cat("\nConclusion: The p-value >= 0.05 → We fail to reject the null hypothesis.\n")
  cat("The prevailing mean forecast does not significantly outperform\n")
  cat("the random-walk (last return) benchmark.\n")
}


```
##Random Forest
```{r}
# Make sure sorted by time
data <- data %>% arrange(date)

# Build RF feature matrix: log_ret + lagged predictors *_L1..*_L4
# We use lagged predictors because our goal is real-time forecasting. Using only lagged values ensures the model never uses information that would not be available at the time of prediction, preventing look-ahead bias and producing realistic out-of-sample performance.

rf_features <- data %>%
  select(
    date,
    log_ret,
    matches("_L[1-4]$")
  )

# Drop rows missing any predictors (early periods)
rf_df <- rf_features %>% drop_na()

set.seed(123)
window_size <- 60
min_history <- 60
n <- nrow(rf_df)

rf_pred <- rep(NA_real_, n)

for (t in seq(min_history + 1, n)) {

  # Rolling training window
  start_idx <- max(1, t - window_size)
  end_idx   <- t - 1

  train_data <- rf_df[start_idx:end_idx, ]
  test_data  <- rf_df[t, , drop = FALSE]

  X_train <- train_data %>% select(-date, -log_ret)
  y_train <- train_data$log_ret

  X_test  <- test_data %>% select(-date, -log_ret)

  p <- ncol(X_train)

#We are using the default parameters for random forest
rf_model <- randomForest(
  x = X_train,
  y = y_train,
  keep.forest = TRUE
)

  rf_pred[t] <- predict(rf_model, newdata = X_test)
}

# Attach forecasts
rf_df$rf_forecast_ret <- rf_pred


```

```{r}
# Join RF forecasts to actual returns and prevailing mean benchmark
plot_rf_df <- rf_df %>%
  left_join(
    data %>% select(date, log_ret, mean_prevailing_ret),
    by = "date"
  )


ggplot(plot_rf_df, aes(x = date)) +
  geom_line(aes(y = log_ret.x,             color = "Actual returns")) +
  geom_line(aes(y = mean_prevailing_ret, color = "Prevailing-mean forecast")) +
  geom_line(aes(y = rf_forecast_ret,     color = "Random Forest forecast")) +
  labs(
    title = "S&P 500 Log Return Forecasts: RF vs Prevailing Mean vs Actual",
    x = "Year",
    y = "Log Return",
    color = ""
  ) +
  theme_minimal()


```

```{r}

rf_eval_df <- plot_rf_df %>%
  filter(
    !is.na(log_ret.x),
    !is.na(rf_forecast_ret),
    !is.na(mean_prevailing_ret)
  ) %>%
  mutate(
    err_rf   = log_ret.x - rf_forecast_ret,
    err_mean = log_ret.x - mean_prevailing_ret
  )

# MAE, RMSE, R²
MAE_rf  <- mae(rf_eval_df$log_ret.x, rf_eval_df$rf_forecast_ret)
RMSE_rf <- rmse(rf_eval_df$log_ret.x, rf_eval_df$rf_forecast_ret)
R2_rf   <- cor(rf_eval_df$log_ret.x, rf_eval_df$rf_forecast_ret)^2

cat("\n=== Random Forest Forecast Performance ===\n")
cat("MAE :", round(MAE_rf, 6),
    "\nRMSE:", round(RMSE_rf, 6),
    "\nR²  :", round(R2_rf, 4))

#DM Test
dm_test_rf <- dm.test(
  e1 = rf_eval_df$err_rf,
  e2 = rf_eval_df$err_mean,
  alternative = "less"   # H1: RF has lower expected error
)

cat("\n=== DM Test: Random Forest VS Prevailing Mean ===\n")
cat("DM Statistic:", round(as.numeric(dm_test_rf$statistic), 3),
    "\nP-value:", signif(dm_test_rf$p.value, 4), "\n")

# === Conclusion Interpretation ===
if(dm_test_rf$p.value < 0.05) {
  cat("\nConclusion: The p-value < 0.05 → We reject the null hypothesis.\n")
  cat("Random Forest has significantly lower forecast error\n")
  cat("than the Prevailing Mean model.\n")
} else {
  cat("\nConclusion: The p-value >= 0.05 → We fail to reject the null hypothesis.\n")
  cat("Random Forest does not significantly outperform\n")
  cat("the Prevailing Mean Model.\n")
}
```
##Feature Importance Plot
```{r}
last_rf_model <- rf_model   # store the final trained model for interpretation

# Extract importance values from final RF model
imp <- importance(last_rf_model)

# Convert to clean dataframe
imp_df <- data.frame(
  Feature = rownames(imp),
  Importance = imp[, 1]   # %IncMSE for regression
)

# Select Top 10 features to display in the plot
imp_top10 <- imp_df %>%
  arrange(desc(Importance)) %>%
  slice_head(n = 10)

ggplot(imp_top10, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Feature Importance (%IncMSE)",
    x = "Predictor (Lagged Features)",
    y = "Importance"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(face = "bold")
  )

```
##PDP Plot
```{r}
top3 <- imp_top10$Feature[1:3]

for (feat in top3) {
  pd <- partial(
    last_rf_model,
    pred.var = feat,
    train = last_X_train,
    grid.resolution = 40
  )
  
  p <- autoplot(pd) +
    labs(
      title = paste("Partial Dependence Plot:", feat),
      x = feat,
      y = "Effect on Predicted Return"
    ) +
    theme_minimal()
  
  print(p)
}

```
##OOB Error
```{r}
plot(last_rf_model, main = "OOB Error vs Number of Trees (Random Forest)")
```
##From the plot, we see that Eeror drops quickly in the first ~50–80 trees, and the curve flattens and stabilizes around 100–150 trees. Afteraround 150 trees, the OOB error does not meaningfully decrease anymore.From around 150 to 500 trees, the model is basically just averaging noise (flat tail). So, adding more trees beyond ~150 does not improve accuracy. Hence, we conclude that we could have choosen about 200 trees for a stable and low variance random forest model.

##XGBoost
```{r}

xgb_pred <- rep(NA_real_, n)

for (t in seq(min_history + 1, n)) {

  start_idx <- max(1, t - window_size)
  end_idx   <- t - 1

  train_data <- rf_df[start_idx:end_idx, ]
  test_data  <- rf_df[t, , drop = FALSE]

  X_train <- train_data %>% select(-date, -log_ret)
  y_train <- train_data$log_ret

  X_test  <- test_data %>% select(-date, -log_ret)

  # Convert to xgb.DMatrix
  dtrain <- xgb.DMatrix(data = as.matrix(X_train), label = y_train)
  dtest  <- xgb.DMatrix(data = as.matrix(X_test))

  # use default parameters (low overfit)
  params <- list(
    objective = "reg:squarederror",
    max_depth = 3,
    eta = 0.1,
    subsample = 0.8,
    colsample_bytree = 0.8
  )

  xgb_model <- xgb.train(
    params = params,
    data = dtrain,
    nrounds = 200,
    verbose = 0
  )
  
  # keep last model + last clean training matrix for interpretation
  last_xgb_model   <- xgb_model
  last_X_train_xgb <- X_train %>% tidyr::drop_na()
  last_feat_names  <- colnames(last_X_train_xgb)


  xgb_pred[t] <- predict(xgb_model, dtest)
}

rf_df$xgb_forecast_ret <- xgb_pred

```

```{r}

plot_xgb_df <- rf_df %>%
  left_join(
    data %>% select(date, log_ret, mean_prevailing_ret),
    by = "date"
  )

ggplot(plot_xgb_df, aes(x = date)) +
  geom_line(aes(y = log_ret.x,            color = "Actual returns")) +
  geom_line(aes(y = mean_prevailing_ret, color = "Prevailing-mean forecast")) +
  geom_line(aes(y = xgb_forecast_ret,    color = "XGBoost forecast")) +
  labs(
    title = "S&P 500 Log Return Forecasts: XGBoost vs Randon Forest vs Prevailing Mean vs Actual",
    x = "Year",
    y = "Log Return",
    color = ""
  ) +
  theme_minimal()
```

```{r}
xgb_eval_df <- plot_xgb_df %>%
  filter(
    !is.na(log_ret.x),
    !is.na(xgb_forecast_ret),
    !is.na(mean_prevailing_ret)
  ) %>%
  mutate(
    err_xgb   = log_ret.x - xgb_forecast_ret,
    err_mean = log_ret.x - mean_prevailing_ret
  )

# MAE, RMSE, R²
MAE_rf  <- mae(xgb_eval_df$log_ret.x, xgb_eval_df$xgb_forecast_ret)
RMSE_rf <- rmse(xgb_eval_df$log_ret.x, xgb_eval_df$xgb_forecast_ret)
R2_rf   <- cor(xgb_eval_df$log_ret.x, xgb_eval_df$xgb_forecast_ret)^2

cat("\n=== Random Forest Forecast Performance ===\n")
cat("MAE :", round(MAE_rf, 6),
    "\nRMSE:", round(RMSE_rf, 6),
    "\nR²  :", round(R2_rf, 4))

#DM Test
dm_test_xgb <- dm.test(
  e1 = xgb_eval_df$err_xgb,
  e2 = xgb_eval_df$err_mean,
  alternative = "less"   # H1: RF has lower expected error
)

cat("\n=== DM Test: XGBoost VS Prevailing Mean ===\n")
cat("DM Statistic:", round(as.numeric(dm_test_xgb$statistic), 3),
    "\nP-value:", signif(dm_test_xgb$p.value, 4), "\n")

# === Conclusion Interpretation ===
if(dm_test_xgb$p.value < 0.05) {
  cat("\nConclusion: The p-value < 0.05 → We reject the null hypothesis.\n")
  cat("XGBoost has significantly lower forecast error\n")
  cat("than the Prevailing Mean model.\n")
} else {
  cat("\nConclusion: The p-value >= 0.05 → We fail to reject the null hypothesis.\n")
  cat("XGBoost does not significantly outperform\n")
  cat("the Prevailing Mean Model.\n")
}
```

##Feature Importance
```{r}
# Feature importance (Top 10 by Gain)
imp_tbl <- xgb.importance(model = last_xgb_model, feature_names = last_feat_names)

imp_top10 <- imp_tbl %>%
  dplyr::slice_max(order_by = Gain, n = 10) %>%
  dplyr::mutate(Feature = factor(Feature, levels = rev(Feature)))

ggplot(imp_top10, aes(x = Feature, y = Gain)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "XGBoost Feature Importance (Top 10)",
       x = "Feature", y = "Gain") +
  theme_minimal()

```

##PDP
```{r}
# top-3 from the top10 list
top3_xgb <- imp_top10$Feature[1:3]

# predict wrapper for pdp
xgb_pred_fun <- function(object, newdata) {
  predict(object, newdata = xgb.DMatrix(as.matrix(newdata)))
}

for (feat in top3_xgb) {

  pd <- partial(
    object = last_xgb_model,
    pred.var = feat,
    train = last_X_train_xgb,
    pred.fun = xgb_pred_fun,
    grid.resolution = 100,   # smoother curve
    ice = FALSE             
  )
  
  p <- autoplot(pd) +
    labs(
      title = paste("PDP:", feat),
      x = feat,
      y = "Effect on Predicted Return"
    ) +
    theme_minimal()
  
  print(p)
}

```

