---
title: "RF Tuned Recursive CV"
author: "marth"
output: html_document
date: "2025-11-10"
---

```{r setup, include=FALSE}

# SETTING UP LIBRARIES
library(dplyr)
library(tidyverse)
library(zoo)
library(lubridate)
library(ggplot2)
library(Metrics)
library(forecast)
library(randomForest)
library(pdp)
```

## Loading Datasets
```{r}

data = read.csv("../data/data_merged.csv")
data_mean = read.csv("../data/prevail_mean.csv")

data$date <- as.Date(data$date)
data_mean$date <- as.Date(data_mean$date)

data <- data %>%
  select(
    date,
    log_ret,
    matches("^price_fd_L[1-5]$"),
    matches("^log_ret_L[1-5]$"),
    matches("_L[2-5]$")       # only lagged features from at least 2 periods ago
  ) %>%
  left_join(
    data_mean %>% select(date, mean_prevailing_ret),
    by = "date"
  )

```


## Random Forest with Hyperparameter Tuning

# ============================================================================
# MODELING APPROACH: 80/20 Train-Test Split w/ Recursive CV
# ============================================================================


```{r}
# --- STEP 1: Prepare data ---

rf_df <- data %>%
  arrange(date) %>%
  select(-mean_prevailing_ret) %>%
  drop_na()

set.seed(123)

n <- nrow(rf_df)
train_size <- floor(0.8 * n)  # 204 quarters (training)
test_size <- n - train_size   # 52 quarters (testing)

cat("\n=== Data Split ===\n")
cat("Training:", train_size, "quarters\n")
cat("Testing :", test_size, "quarters\n")

train_data <- rf_df[1:train_size, ]
test_data  <- rf_df[(train_size + 1):n, ]

# --- STEP 2: Hyperparameter tuning  ---
# will conduct hyperparameter tuning to get a set of best params, then recursive CV using these

rf_param_grid <- expand.grid(
  mtry = c(5, 10, 20, 30, 40, 50),    # important: feature sampling 
  nodesize = c(1, 2, 3, 5),           # regularisation/tree complexity
  ntree = c(50, 100, 200)                         # given train size 165, we can afford 200 trees
)

train_tune_size <- floor(0.8 * train_size)
val_tune_size   <- train_size - train_tune_size

X_train_tune <- train_data[1:train_tune_size, ] %>% select(-date, -log_ret)
y_train_tune <- train_data$log_ret[1:train_tune_size]

X_val_tune <- train_data[(train_tune_size + 1):train_size, ] %>% select(-date, -log_ret)
y_val_tune <- train_data$log_ret[(train_tune_size + 1):train_size]

best_val_rmse <- Inf
best_params_rf <- NULL


for (i in 1:nrow(rf_param_grid)) {
  temp_rf <- randomForest(
    x = X_train_tune, y = y_train_tune,
    ntree = rf_param_grid$ntree[i],
    mtry = rf_param_grid$mtry[i],
    nodesize = rf_param_grid$nodesize[i]
  )
  
  val_preds <- predict(temp_rf, newdata = X_val_tune)
  val_rmse <- sqrt(mean((y_val_tune - val_preds)^2))
  
  if (val_rmse < best_val_rmse) {
    best_val_rmse <- val_rmse
    best_params_rf <- rf_param_grid[i, ]
  }
}

cat("\n=== Best RF Params ===\n")
cat("mtry     :", best_params_rf$mtry, "\n")
cat("nodesize :", best_params_rf$nodesize, "\n")
cat("Validation RMSE:", round(best_val_rmse, 5), "\n")

# --- STEP 3: Recursive CV ---
# will use an expanding window, loop over 

cat("\n=== Starting Recursive Forecast Evaluation ===\n")

rf_model_final <- randomForest(
  x = train_data %>% select(-date, -log_ret),
  y = train_data$log_ret,
  ntree = best_params_rf$ntree,
  mtry = best_params_rf$mtry,
  nodesize = best_params_rf$nodesize
)

# --- STEP 4: Predictions ---
# will NOT do recusrive estimation, simple prediction only

rf_train_forecasts <- predict(rf_model_final,
                             newdata = train_data %>% select(-date, -log_ret))

rf_forecasts <- predict(rf_model_final, 
                        newdata = test_data %>% select(-date, -log_ret))


cat("Prediction complete!\n")

# Store forecasts in df
rf_df$rf_forecast_ret <- c(rf_train_forecasts, rf_forecasts)


```


```{r}
# Join RF forecasts to actual returns and prevailing mean benchmark
plot_rf_df <- rf_df %>%
  left_join(
    data %>% select(date, mean_prevailing_ret),
    by = "date"
  ) %>%
  drop_na()

# ONLY PLOT TEST SET (last 52 quarters)

ggplot(plot_rf_df %>% tail(test_size), aes(x = date)) +
  geom_line(aes(y = log_ret,             color = "Actual returns")) +
  geom_line(aes(y = mean_prevailing_ret, color = "Prevailing-mean forecast")) +
  geom_line(aes(y = rf_forecast_ret,     color = "Random Forest forecast")) +
  labs(
    title = "S&P 500 Log Return Forecasts: RF vs Prevailing Mean vs Actual (TEST SET)",
    subtitle = paste("Test period:", test_size, "quarters (Q207-Q258)"),
    x = "Year",
    y = "Log Return",
    color = ""
  ) +
  theme_minimal()


```

```{r}
# EVALUATE ONLY ON TEST SET
rf_eval_df <- plot_rf_df %>%
  tail(test_size) %>%   # last 52 quarters = test set
  filter(
    !is.na(log_ret),
    !is.na(rf_forecast_ret),
    !is.na(mean_prevailing_ret)
  ) %>%
  mutate(
    err_rf   = log_ret - rf_forecast_ret,
    err_mean = log_ret - mean_prevailing_ret
  )

cat("\n=== TEST SET PERFORMANCE (Q207-Q258, n =", nrow(rf_eval_df), "quarters) ===\n\n")

# MAE, RMSE, R²
actual <- tail(rf_df$log_ret, test_size)
pred   <- rf_forecasts

MAE_rf  <- mean(abs(actual - pred))
RMSE_rf <- sqrt(mean((actual - pred)^2))
R2_rf   <- cor(actual, pred, use = "complete.obs")^2


cat("=== Random Forest Forecast Performance ===\n")
cat("MAE :", round(MAE_rf, 6),
    "\nRMSE:", round(RMSE_rf, 6),
    "\nR²  :", round(R2_rf, 4))

dm_test_rf <- dm.test(
  e1 = rf_eval_df$err_rf,
  e2 = rf_eval_df$err_mean,
  alternative = "less"
)

cat("\n\n=== DM Test: Random Forest VS Prevailing Mean ===\n")
cat("DM Statistic:", round(as.numeric(dm_test_rf$statistic), 3),
    "\nP-value:", signif(dm_test_rf$p.value, 4), "\n")

# === Conclusion Interpretation ===
if(dm_test_rf$p.value < 0.05) {
  cat("\nConclusion: The p-value < 0.05 → We reject the null hypothesis.\n")
  cat("Random Forest has significantly lower forecast error\n")
  cat("than the Prevailing Mean model.\n")
} else {
  cat("\nConclusion: The p-value >= 0.05 → We fail to reject the null hypothesis.\n")
  cat("Random Forest does not significantly outperform the Prevailing Mean Model.\n")
}
```

``` {r}

library(sandwich)  # Newey–West
library(lmtest)

# Directional 0-1 loss: 1 if signs disagree, else 0
dir_loss <- function(actual, pred, zero_rule = "tie_ok") {
  # zero_rule: how to treat zeros in actual or pred
  #  - "tie_ok": a zero counts as "no sign" and matches only if both are zero
  #  - "nonpos": treat 0 as non-positive (<=0)
  if (zero_rule == "tie_ok") {
    a <- ifelse(actual > 0,  1, ifelse(actual < 0, -1, 0))
    p <- ifelse(pred   > 0,  1, ifelse(pred   < 0, -1, 0))
  } else {
    a <- ifelse(actual > 0, 1, -1)  # 0 -> -1
    p <- ifelse(pred   > 0, 1, -1)
  }
  as.numeric(a != p)
}

# Robust DM test for *any* scalar loss (e.g., 0-1 directional loss)
# loss1, loss2 must be aligned vectors (same timestamps)
dm_test_hac_loss <- function(loss1, loss2, h = 1, lag = NULL, two_sided = TRUE) {
  stopifnot(length(loss1) == length(loss2))
  d <- loss1 - loss2
  d <- d[is.finite(d)]
  if (length(d) < 5) stop("Not enough observations for DM test.")

  # Regress on intercept and use HAC variance for long-run variance of d
  fit <- lm(d ~ 1)
  # Choose HAC lag: if not supplied, use max(h-1, floor(T^(1/4)))
  Tn <- length(d)
  if (is.null(lag)) lag <- max(h - 1, floor(Tn^(1/4)))
  S <- NeweyWest(fit, lag = lag, prewhite = FALSE, adjust = TRUE)

  dbar <- coef(fit)[1]
  se   <- sqrt(S[1, 1] / Tn)
  stat <- as.numeric(dbar / se)

  if (two_sided) {
    pval <- 2 * (1 - pnorm(abs(stat)))
  } else {
    # one-sided H1: model 1 has *lower* loss than model 2  (mean(d) < 0)
    pval <- pnorm(stat)  # P(Z <= stat)
  }
  list(statistic = stat, p.value = pval, mean_diff = dbar, n = Tn, lag_used = lag)
}


# Example construction if you have a comparison_df already
# with columns: date, log_ret, rf_forecast_ret, xgb_forecast_ret, mean_prevailing_ret
y_test   <- rf_eval_df$log_ret
rf_pred  <- rf_eval_df$rf_forecast_ret
mean_pred<- rf_eval_df$mean_prevailing_ret

# Directional 0-1 losses
rf_loss   <- dir_loss(y_test, rf_pred,  zero_rule = "nonpos")
mean_loss <- dir_loss(y_test, mean_pred,zero_rule = "nonpos")

# --- Robust DM (recommended for recursive estimation, or just to be safe) ---
# H1: first model has *lower* loss than the second (one-sided)
dm_rf_vs_mean  <- dm_test_hac_loss(rf_loss,  mean_loss, h = 1, two_sided = FALSE)

cat("\nDM (RF < Mean):     stat =", round(dm_rf_vs_mean$statistic, 3),
    "p =", signif(dm_rf_vs_mean$p.value, 4), "\n")


```

##Feature Importance Plot
```{r}
# Extract importance values from final RF model 
imp <- importance(rf_model_final)

# Convert to clean dataframe
imp_df <- data.frame(
  Feature = rownames(imp),
  Importance = imp[, 1]   # %IncMSE for regression
)

# Select Top 10 features to display in the plot
imp_top10 <- imp_df %>%
  arrange(desc(Importance)) %>%
  slice_head(n = 10)

ggplot(imp_top10, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Feature Importance (%IncMSE)",
    x = "Predictor (Lagged Features)",
    y = "Importance"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(face = "bold")
  )

```
##PDP Plot
```{r}
top3 <- imp_top10$Feature[1:3]

for (feat in top3) {
  pd <- partial(
    rf_model_final,
    pred.var = feat,
    train = train_data %>% select(-date, -log_ret),
    grid.resolution = 40
  )
  
  p <- autoplot(pd) +
    labs(
      title = paste("Partial Dependence Plot:", feat),
      x = feat,
      y = "Effect on Predicted Return"
    ) +
    theme_minimal()
  
  print(p)
}

```
##OOB Error
```{r}
plot(rf_model_final, main = "OOB Error vs Number of Trees (Random Forest)")
```
##From the plot, we see that Eeror drops quickly in the first ~50 trees, and the curve flattens and stabilizes around 100 trees. Afteraround 100 trees, the OOB error does not meaningfully decrease anymor (the model is basically just averaging noise (flat tail)). So, adding more trees beyond ~150 does not improve accuracy. Hence, we conclude that we could have choosen about 150 trees for a stable and low variance random forest model.

``` {r}
# save model for later use 
write_csv(rf_df, "../data/rf_results.csv")
write_csv(rf_eval_df, "../data/rf_evals.csv")


# for train data and predictions only
rf_train_preds <- rf_df %>%
  head(train_size) %>%  
  filter(
    !is.na(log_ret),
    !is.na(rf_forecast_ret)
  )
  
write_csv(rf_train_preds, "../data/rf_train_predictions.csv")  
```
