---
title: "Multi-Horizon Tactical Asset Allocation Strategy Backtest"
author: "EC4308 - S&P 500 Forecasting Project"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 6)

# Load required libraries
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(knitr)
library(scales)
library(gridExtra)
```

# Executive Summary

This document implements and evaluates **tactical asset allocation strategies** using machine learning forecasts across three rebalancing horizons:

- **Quarterly (3-month)**: Rebalance every quarter
- **Semi-Annual (6-month)**: Rebalance every 6 months
- **Annual (12-month)**: Rebalance every 12 months

## Strategy Design

**Decision Rule:** At each rebalancing date, compare forecasted S&P 500 return with T-bill rate:

- If **Forecasted S&P 500 ≥ T-bill Rate** → Invest in S&P 500
- If **Forecasted S&P 500 < T-bill Rate** → Invest in T-bills

**Key Features:**

- Starting capital: $1.00
- 100% allocation (no partial positions)
- No transaction costs (baseline)
- Test period: 2011-Q4 to 2024-Q3

---

# 1. Load Data

## 1.1 Load Consolidated Forecasts

```{r load-forecasts}
cat("\n=== LOADING CONSOLIDATED FORECAST DATA ===\n\n")

# Load multi-horizon forecasts
forecast_data <- read_csv("../data/consolidated_forecasts_multihorizon.csv")

cat("✓ Forecasts loaded:", nrow(forecast_data), "observations\n")
cat("  Date range:", min(forecast_data$date), "to", max(forecast_data$date), "\n")
cat("  Total columns:", ncol(forecast_data), "\n\n")

cat("Available models:\n")
cat("  - Random Forest (rf)\n")
cat("  - Lasso\n")
cat("  - Ridge\n")
cat("  - PCR\n")
cat("  - Elastic Net\n")
cat("  - Ensemble (Equal-Weight)\n\n")

cat("Available horizons:\n")
cat("  - 3-month (quarterly)\n")
cat("  - 6-month (semi-annual)\n")
cat("  - 12-month (annual)\n\n")

head(forecast_data, 5)
```

## 1.2 Load Ensemble Forecasts

```{r load-ensemble}
cat("\n=== LOADING ENSEMBLE EQUAL-WEIGHT FORECASTS ===\n\n")

# Load ensemble forecasts
ensemble_data <- read_csv("../data/ensemble_equal_weight_forecasts.csv")

cat("✓ Ensemble forecasts loaded:", nrow(ensemble_data), "observations\n")
cat("  Date range:", min(ensemble_data$date), "to", max(ensemble_data$date), "\n")
cat("  Ensemble method: Equal-weight average of Ridge, RF, and PCR\n")
cat("  Weights: 33.33% each\n\n")

head(ensemble_data, 5)
```

## 1.3 Load T-Bill Returns

```{r load-tbills}
cat("\n=== LOADING T-BILL RETURNS ===\n\n")

# Load T-bill returns for all horizons
tbill_data <- read_csv("../data/tbill_returns_multihorizon.csv")

cat("✓ T-Bill returns loaded:", nrow(tbill_data), "observations\n")
cat("  Date range:", min(tbill_data$date), "to", max(tbill_data$date), "\n\n")

cat("T-Bill horizons:\n")
cat("  - tbill_return_3m: 3-month holding period return\n")
cat("  - tbill_return_6m: 6-month holding period return\n")
cat("  - tbill_return_12m: 12-month holding period return\n\n")

head(tbill_data, 5)
```

## 1.4 Merge Forecasts with T-Bills

```{r merge-data}
cat("\n=== MERGING FORECAST AND T-BILL DATA ===\n\n")

# Merge forecasts with ensemble forecasts
forecast_data <- forecast_data %>%
  left_join(ensemble_data %>% select(date, starts_with("ensemble_")), by = "date")

# Merge with T-bills
backtest_data <- forecast_data %>%
  inner_join(tbill_data, by = "date") %>%
  arrange(date) %>%
  drop_na()

cat("✓ Data merged successfully!\n")
cat("  Final observations:", nrow(backtest_data), "\n")
cat("  Date range:", min(backtest_data$date), "to", max(backtest_data$date), "\n")
cat("  Total columns:", ncol(backtest_data), "\n\n")

cat("Dataset structure:\n")
cat("  - Actual returns: actual_3m, actual_6m, actual_12m\n")
cat("  - Model forecasts: [model]_forecast_[horizon]\n")
cat("  - Ensemble forecasts: ensemble_forecast_3m, ensemble_forecast_6m, ensemble_forecast_12m\n")
cat("  - T-Bill returns: tbill_return_3m, tbill_return_6m, tbill_return_12m\n\n")

# Show summary
cat("Summary statistics:\n")
summary(backtest_data %>% select(actual_3m, actual_6m, actual_12m,
                                   tbill_return_3m, tbill_return_6m, tbill_return_12m))
```

---

# 2. Backtesting Framework

## 2.1 Tactical Strategy Function

```{r strategy-function}
#' Tactical Asset Allocation Backtest
#' 
#' @param data Data frame with date, forecast, actual, and tbill columns
#' @param forecast_col Name of forecast column
#' @param actual_col Name of actual return column (ALWAYS use actual_3m for quarterly returns)
#' @param tbill_col Name of T-bill return column (ALWAYS use tbill_return_3m for quarterly returns)
#' @param horizon Rebalancing horizon: "3m", "6m", or "12m"
#' @param initial_capital Starting capital (default: 1)
#' @param strategy_name Name for reporting
#' 
#' @return List containing results dataframe and performance metrics
tactical_backtest <- function(data, forecast_col, actual_col, tbill_col, horizon = "3m",
                               initial_capital = 1, strategy_name = "Strategy") {
  
  cat("\n=== Backtesting:", strategy_name, "===\n")
  
  # Determine rebalancing frequency (quarters to skip between rebalances)
  rebal_freq <- switch(horizon,
                       "3m" = 1,   # Rebalance every quarter
                       "6m" = 2,   # Rebalance every 2 quarters
                       "12m" = 4,  # Rebalance every 4 quarters
                       1)
  
  # Extract relevant columns from full dataset
  full_data <- data %>%
    select(date, 
           forecast = all_of(forecast_col),
           actual = all_of(actual_col), 
           tbill = all_of(tbill_col)) %>%
    arrange(date) %>%
    drop_na()
  
  # For strategy decisions: sample at rebalancing frequency
  # For 6m: use rows 1, 3, 5, 7... (every 2nd quarter)
  # For 12m: use rows 1, 5, 9, 13... (every 4th quarter)
  decision_rows <- seq(1, nrow(full_data), by = rebal_freq)
  decision_df <- full_data[decision_rows, ]
  
  n_rebalances <- nrow(decision_df)
  
  cat("  Rebalancing periods:", n_rebalances, "\n")
  cat("  Date range:", min(decision_df$date), "to", max(decision_df$date), "\n")
  
  # For each rebalancing decision, we hold for rebal_freq quarters
  # We need to track which asset we hold and compound quarterly returns accordingly
  
  # Initialize position tracking
  position <- character(n_rebalances + 1)
  position[1] <- NA
  
  # Build a quarter-by-quarter allocation series
  n_quarters <- nrow(full_data)
  quarterly_position <- character(n_quarters)
  quarterly_returns <- numeric(n_quarters)
  
  # Fill in positions for each quarter based on rebalancing decisions
  for (t in 1:n_rebalances) {
    decision_date_idx <- decision_rows[t]
    
    # Determine position for this rebalancing period
    if (decision_df$forecast[t] >= decision_df$tbill[t]) {
      position[t + 1] <- "SP500"
      current_position <- "SP500"
    } else {
      position[t + 1] <- "TBILL"
      current_position <- "TBILL"
    }
    
    # Apply this position to the next rebal_freq quarters
    end_idx <- min(decision_date_idx + rebal_freq - 1, n_quarters)
    quarterly_position[decision_date_idx:end_idx] <- current_position
    
    # Record which returns to use for each quarter
    if (current_position == "SP500") {
      quarterly_returns[decision_date_idx:end_idx] <- full_data$actual[decision_date_idx:end_idx]
    } else {
      quarterly_returns[decision_date_idx:end_idx] <- full_data$tbill[decision_date_idx:end_idx]
    }
  }
  
  # Now compound quarterly returns just like buy-and-hold
  portfolio_value_quarterly <- initial_capital * cumprod(1 + quarterly_returns)
  portfolio_value_quarterly <- c(initial_capital, portfolio_value_quarterly)
  
  # Extract values at rebalancing dates for reporting
  # For each decision, we want the value AFTER holding for rebal_freq quarters
  # decision_rows[i] is the quarter where we make decision
  # We hold through quarters [decision_rows[i], decision_rows[i] + rebal_freq - 1]
  # So we want portfolio value after quarter [decision_rows[i] + rebal_freq - 1]
  # In portfolio_value_quarterly, that's index [decision_rows[i] + rebal_freq]
  portfolio_value <- portfolio_value_quarterly[c(1, pmin(decision_rows + rebal_freq, length(portfolio_value_quarterly)))]
  
  # Calculate period returns for metrics (between rebalancing dates)
  period_returns <- numeric(n_rebalances)
  for (t in 1:n_rebalances) {
    if (t == 1) {
      period_returns[t] <- (portfolio_value[t + 1] / portfolio_value[1]) - 1
    } else {
      period_returns[t] <- (portfolio_value[t + 1] / portfolio_value[t]) - 1
    }
  }
  
  # Create results dataframe
  # Start date is one quarter before first decision date
  start_date <- decision_df$date[1] %m-% months(3)
  
  results_df <- data.frame(
    date = c(start_date, decision_df$date),
    portfolio_value = portfolio_value,
    position = position,
    returns_earned = c(NA, period_returns),
    forecast = c(NA, decision_df$forecast),
    decision_tbill = c(NA, decision_df$tbill)
  )
  
  # Calculate performance metrics
  total_return <- (portfolio_value[n_rebalances + 1] / portfolio_value[1]) - 1
  
  # Annualization factor depends on horizon
  periods_per_year <- switch(horizon, "3m" = 4, "6m" = 2, "12m" = 1, 4)
  annualized_return <- (1 + total_return)^(periods_per_year / n_rebalances) - 1
  
  # Volatility (annualized)
  returns_sd <- sd(period_returns, na.rm = TRUE)
  annualized_vol <- returns_sd * sqrt(periods_per_year)
  
  # Sharpe ratio (comparing period returns to risk-free rate)
  # For fair comparison, compound T-bill returns over holding period for excess return calc
  excess_returns <- period_returns  # Already net of opportunity cost in decision
  sharpe_ratio <- mean(period_returns, na.rm = TRUE) / sd(period_returns, na.rm = TRUE) * sqrt(periods_per_year)
  
  # Max drawdown
  cummax_value <- cummax(portfolio_value)
  drawdown <- (portfolio_value - cummax_value) / cummax_value
  max_drawdown <- min(drawdown)
  
  # Win rate
  win_rate <- sum(period_returns > 0, na.rm = TRUE) / sum(!is.na(period_returns))
  
  # Turnover (position changes)
  switches <- sum(position[-1] != position[-length(position)], na.rm = TRUE)
  turnover <- switches / n_rebalances
  
  # Time in market
  time_in_sp500 <- sum(position == "SP500", na.rm = TRUE) / sum(!is.na(position))
  
  metrics <- list(
    strategy_name = strategy_name,
    total_return = total_return,
    annualized_return = annualized_return,
    annualized_vol = annualized_vol,
    sharpe_ratio = sharpe_ratio,
    max_drawdown = max_drawdown,
    win_rate = win_rate,
    turnover = turnover,
    n_switches = switches,
    time_in_sp500 = time_in_sp500,
    final_value = portfolio_value[n_rebalances + 1]
  )
  
  cat("  Final value: $", round(metrics$final_value, 4), "\n")
  cat("  Total return:", percent(metrics$total_return, accuracy = 0.01), "\n")
  cat("  Sharpe ratio:", round(metrics$sharpe_ratio, 3), "\n")
  
  return(list(results = results_df, metrics = metrics))
}
```

## 2.2 Benchmark Functions

```{r benchmark-functions}
#' Buy-and-Hold Benchmark
#' 
#' @param data Data frame with date and returns
#' @param actual_col Name of actual return column  
#' @param tbill_col Name of T-bill return column (only used for T-bill benchmark)
#' @param initial_capital Starting capital (default: 1)
#' @param benchmark_name Name for reporting
#' @param asset "SP500" or "TBILL"
buy_hold_benchmark <- function(data, actual_col = NULL, tbill_col = NULL, 
                                initial_capital = 1, 
                                benchmark_name = "Buy-Hold", asset = "SP500") {
  
  cat("\n=== Benchmark:", benchmark_name, "===\n")
  
  # For T-bills, use tbill_col; for S&P500, use actual_col
  if (asset == "TBILL") {
    if (is.null(tbill_col)) stop("tbill_col must be provided for T-bill benchmark")
    backtest_df <- data %>%
      select(date, actual = all_of(tbill_col)) %>%
      arrange(date) %>%
      drop_na()
  } else {
    if (is.null(actual_col)) stop("actual_col must be provided for S&P500 benchmark")
    backtest_df <- data %>%
      select(date, actual = all_of(actual_col)) %>%
      arrange(date) %>%
      drop_na()
  }
  
  n_periods <- nrow(backtest_df)
  
  cat("  Periods:", n_periods, "\n")
  cat("  Date range:", min(backtest_df$date), "to", max(backtest_df$date), "\n")
  
  # Calculate cumulative returns - compound returns over time
  # This is correct: each period's return applies to the current portfolio value
  portfolio_value <- c(initial_capital, initial_capital * cumprod(1 + backtest_df$actual))
  
  # Start date is one quarter before first return date
  start_date <- backtest_df$date[1] %m-% months(3)
  
  results_df <- data.frame(
    date = c(start_date, backtest_df$date),
    portfolio_value = portfolio_value,
    position = asset
  )
  
  # Calculate metrics
  total_return <- (portfolio_value[n_periods + 1] / portfolio_value[1]) - 1
  
  # Annualization factor - determine from column name
  col_name <- if (asset == "TBILL") tbill_col else actual_col
  periods_per_year <- if (grepl("3m", col_name)) 4 else if (grepl("6m", col_name)) 2 else 1
  annualized_return <- (1 + total_return)^(periods_per_year / n_periods) - 1
  
  # Volatility
  returns_sd <- sd(backtest_df$actual, na.rm = TRUE)
  annualized_vol <- returns_sd * sqrt(periods_per_year)
  
  # Max drawdown
  cummax_value <- cummax(portfolio_value)
  drawdown <- (portfolio_value - cummax_value) / cummax_value
  max_drawdown <- min(drawdown)
  
  metrics <- list(
    strategy_name = benchmark_name,
    total_return = total_return,
    annualized_return = annualized_return,
    annualized_vol = annualized_vol,
    sharpe_ratio = NA,
    max_drawdown = max_drawdown,
    win_rate = NA,
    turnover = 0,
    n_switches = 0,
    time_in_sp500 = if(asset == "SP500") 1 else 0,
    final_value = portfolio_value[n_periods + 1]
  )
  
  cat("  Final value: $", round(metrics$final_value, 4), "\n")
  cat("  Total return:", percent(metrics$total_return, accuracy = 0.01), "\n")
  
  return(list(results = results_df, metrics = metrics))
}

#' Perfect Foresight (Oracle) Benchmark
perfect_foresight_benchmark <- function(data, actual_col, tbill_col, horizon = "3m",
                                         initial_capital = 1, 
                                         benchmark_name = "Perfect Foresight") {
  
  cat("\n=== Benchmark:", benchmark_name, "===\n")
  
  # Use actual returns as "forecasts" - perfect market timing
  oracle_data <- data %>%
    mutate(forecast = get(actual_col))
  
  result <- tactical_backtest(
    data = oracle_data,
    forecast_col = "forecast",
    actual_col = actual_col,
    tbill_col = tbill_col,
    horizon = horizon,
    initial_capital = initial_capital,
    strategy_name = benchmark_name
  )
  
  return(result)
}
```

---

# 3. Run Backtests - Quarterly (3-Month) Horizon

```{r backtest-3m}
cat("\n")
cat("QUARTERLY (3-MONTH) REBALANCING STRATEGIES\n")

# Models to test
models_3m <- c("rf", "lasso", "ridge", "pcr", "elastic", "ensemble")

# Initialize results list
results_3m <- list()

# Run tactical strategies
for (model in models_3m) {
  forecast_col <- paste0(model, "_forecast_3m")
  strategy_name <- paste0(toupper(model), " Tactical (3M)")
  
  results_3m[[strategy_name]] <- tactical_backtest(
    data = backtest_data,
    forecast_col = forecast_col,
    actual_col = "actual_3m",  # Always use quarterly returns
    tbill_col = "tbill_return_3m",  # Always use quarterly T-bill returns
    horizon = "3m",  # Rebalance quarterly
    strategy_name = strategy_name
  )
}

# Benchmarks
results_3m[["Buy-Hold S&P500 (3M)"]] <- buy_hold_benchmark(
  data = backtest_data,
  actual_col = "actual_3m",
  benchmark_name = "Buy-Hold S&P500 (3M)",
  asset = "SP500"
)

results_3m[["Buy-Hold T-Bill (3M)"]] <- buy_hold_benchmark(
  data = backtest_data,
  tbill_col = "tbill_return_3m",
  benchmark_name = "Buy-Hold T-Bill (3M)",
  asset = "TBILL"
)

results_3m[["Perfect Foresight (3M)"]] <- perfect_foresight_benchmark(
  data = backtest_data,
  actual_col = "actual_3m",
  tbill_col = "tbill_return_3m",
  horizon = "3m",  # Rebalance quarterly
  benchmark_name = "Perfect Foresight (3M)"
)

cat("\n✓ Quarterly (3M) backtests completed!\n")
```

---

# 4. Run Backtests - Semi-Annual (6-Month) Horizon

```{r backtest-6m}
cat("\n")
cat("SEMI-ANNUAL (6-MONTH) REBALANCING STRATEGIES\n")

# Models to test
models_6m <- c("rf", "lasso", "ridge", "pcr", "elastic", "ensemble")

# Initialize results list
results_6m <- list()

# Run tactical strategies
for (model in models_6m) {
  forecast_col <- paste0(model, "_forecast_6m")
  strategy_name <- paste0(toupper(model), " Tactical (6M)")
  
  results_6m[[strategy_name]] <- tactical_backtest(
    data = backtest_data,
    forecast_col = forecast_col,
    actual_col = "actual_3m",  # Always use quarterly returns, not overlapping 6m returns
    tbill_col = "tbill_return_3m",  # Always use quarterly T-bill returns
    horizon = "6m",  # Rebalance every 6 months (every 2 quarters)
    strategy_name = strategy_name
  )
}

# Benchmarks - NOTE: Buy-hold uses QUARTERLY returns regardless of horizon
# This ensures benchmarks are identical across all horizons (as they should be)
results_6m[["Buy-Hold S&P500 (6M)"]] <- buy_hold_benchmark(
  data = backtest_data,
  actual_col = "actual_3m",  # Always use quarterly returns, not 6m overlapping returns
  benchmark_name = "Buy-Hold S&P500 (6M)",
  asset = "SP500"
)

results_6m[["Buy-Hold T-Bill (6M)"]] <- buy_hold_benchmark(
  data = backtest_data,
  tbill_col = "tbill_return_3m",  # Always use quarterly T-bill returns
  benchmark_name = "Buy-Hold T-Bill (6M)",
  asset = "TBILL"
)

results_6m[["Perfect Foresight (6M)"]] <- perfect_foresight_benchmark(
  data = backtest_data,
  actual_col = "actual_3m",  # Always use quarterly returns to avoid overlapping
  tbill_col = "tbill_return_3m",  # Always use quarterly T-bill returns
  horizon = "6m",  # Rebalance every 6 months
  benchmark_name = "Perfect Foresight (6M)"
)

cat("\n✓ Semi-Annual (6M) backtests completed!\n")
```

---

# 5. Run Backtests - Annual (12-Month) Horizon

```{r backtest-12m}
cat("\n")
cat("ANNUAL (12-MONTH) REBALANCING STRATEGIES\n")

# Models to test
models_12m <- c("rf", "lasso", "ridge", "pcr", "elastic", "ensemble")

# Initialize results list
results_12m <- list()

# Run tactical strategies
for (model in models_12m) {
  forecast_col <- paste0(model, "_forecast_12m")
  strategy_name <- paste0(toupper(model), " Tactical (12M)")
  
  results_12m[[strategy_name]] <- tactical_backtest(
    data = backtest_data,
    forecast_col = forecast_col,
    actual_col = "actual_3m",  # Always use quarterly returns, not overlapping 12m returns
    tbill_col = "tbill_return_3m",  # Always use quarterly T-bill returns
    horizon = "12m",  # Rebalance annually (every 4 quarters)
    strategy_name = strategy_name
  )
}

# Benchmarks - NOTE: Buy-hold uses QUARTERLY returns regardless of horizon
# This ensures benchmarks are identical across all horizons (as they should be)
results_12m[["Buy-Hold S&P500 (12M)"]] <- buy_hold_benchmark(
  data = backtest_data,
  actual_col = "actual_3m",  # Always use quarterly returns, not 12m overlapping returns
  benchmark_name = "Buy-Hold S&P500 (12M)",
  asset = "SP500"
)

results_12m[["Buy-Hold T-Bill (12M)"]] <- buy_hold_benchmark(
  data = backtest_data,
  tbill_col = "tbill_return_3m",  # Always use quarterly T-bill returns
  benchmark_name = "Buy-Hold T-Bill (12M)",
  asset = "TBILL"
)

results_12m[["Perfect Foresight (12M)"]] <- perfect_foresight_benchmark(
  data = backtest_data,
  actual_col = "actual_3m",  # Always use quarterly returns to avoid overlapping
  tbill_col = "tbill_return_3m",  # Always use quarterly T-bill returns
  horizon = "12m",  # Rebalance annually
  benchmark_name = "Perfect Foresight (12M)"
)

cat("\n✓ Annual (12M) backtests completed!\n")
```

---

# 6. Performance Comparison

## 6.1 Summary Table - Quarterly (3M)

```{r performance-table-3m}
# Extract metrics from all strategies
metrics_df_3m <- bind_rows(lapply(results_3m, function(x) as.data.frame(x$metrics)))

# Format table
performance_3m <- metrics_df_3m %>%
  select(Strategy = strategy_name, 
         `Total Return` = total_return,
         `Ann. Return` = annualized_return,
         `Ann. Vol` = annualized_vol,
         `Sharpe Ratio` = sharpe_ratio,
         `Max DD` = max_drawdown,
         `Win Rate` = win_rate,
         Turnover = turnover,
         `Switches` = n_switches,
         `% in S&P500` = time_in_sp500,
         `Final Value` = final_value) %>%
  mutate(across(c(`Total Return`, `Ann. Return`, `Ann. Vol`, `Max DD`, `Win Rate`, Turnover, `% in S&P500`), 
                ~percent(., accuracy = 0.01)),
         across(c(`Sharpe Ratio`), ~round(., 3)),
         across(c(`Final Value`), ~round(., 4)))

cat("\n=== QUARTERLY (3-MONTH) REBALANCING PERFORMANCE ===\n\n")
kable(performance_3m, align = "lrrrrrrrrrrr", 
      caption = "Performance Metrics - Quarterly Strategies")
```

## 6.2 Summary Table - Semi-Annual (6M)

```{r performance-table-6m}
# Extract metrics
metrics_df_6m <- bind_rows(lapply(results_6m, function(x) as.data.frame(x$metrics)))

# Format table
performance_6m <- metrics_df_6m %>%
  select(Strategy = strategy_name, 
         `Total Return` = total_return,
         `Ann. Return` = annualized_return,
         `Ann. Vol` = annualized_vol,
         `Sharpe Ratio` = sharpe_ratio,
         `Max DD` = max_drawdown,
         `Win Rate` = win_rate,
         Turnover = turnover,
         `Switches` = n_switches,
         `% in S&P500` = time_in_sp500,
         `Final Value` = final_value) %>%
  mutate(across(c(`Total Return`, `Ann. Return`, `Ann. Vol`, `Max DD`, `Win Rate`, Turnover, `% in S&P500`), 
                ~percent(., accuracy = 0.01)),
         across(c(`Sharpe Ratio`), ~round(., 3)),
         across(c(`Final Value`), ~round(., 4)))

cat("\n=== SEMI-ANNUAL (6-MONTH) REBALANCING PERFORMANCE ===\n\n")
kable(performance_6m, align = "lrrrrrrrrrrr", 
      caption = "Performance Metrics - Semi-Annual Strategies")
```

## 6.3 Summary Table - Annual (12M)

```{r performance-table-12m}
# Extract metrics
metrics_df_12m <- bind_rows(lapply(results_12m, function(x) as.data.frame(x$metrics)))

# Format table
performance_12m <- metrics_df_12m %>%
  select(Strategy = strategy_name, 
         `Total Return` = total_return,
         `Ann. Return` = annualized_return,
         `Ann. Vol` = annualized_vol,
         `Sharpe Ratio` = sharpe_ratio,
         `Max DD` = max_drawdown,
         `Win Rate` = win_rate,
         Turnover = turnover,
         `Switches` = n_switches,
         `% in S&P500` = time_in_sp500,
         `Final Value` = final_value) %>%
  mutate(across(c(`Total Return`, `Ann. Return`, `Ann. Vol`, `Max DD`, `Win Rate`, Turnover, `% in S&P500`), 
                ~percent(., accuracy = 0.01)),
         across(c(`Sharpe Ratio`), ~round(., 3)),
         across(c(`Final Value`), ~round(., 4)))

cat("\n=== ANNUAL (12-MONTH) REBALANCING PERFORMANCE ===\n\n")
kable(performance_12m, align = "lrrrrrrrrrrr", 
      caption = "Performance Metrics - Annual Strategies")
```

---

# 7. Visualizations

## 7.1 Portfolio Value Evolution - Quarterly (3M)

```{r plot-portfolio-3m, fig.height=8}
# Combine portfolio values
portfolio_3m <- bind_rows(lapply(names(results_3m), function(name) {
  results_3m[[name]]$results %>%
    select(date, portfolio_value) %>%
    mutate(Strategy = name)
}))

# Create plot
ggplot(portfolio_3m, aes(x = date, y = portfolio_value, color = Strategy, linetype = Strategy)) +
  geom_line(linewidth = 1.2) +
  scale_y_log10(labels = dollar_format()) +
  scale_color_manual(values = c(
    "RF Tactical (3M)" = "#E41A1C",
    "LASSO Tactical (3M)" = "#377EB8",
    "RIDGE Tactical (3M)" = "#4DAF4A",
    "PCR Tactical (3M)" = "#984EA3",
    "ELASTIC Tactical (3M)" = "#FF7F00",
    "ENSEMBLE Tactical (3M)" = "#A65628",
    "Buy-Hold S&P500 (3M)" = "black",
    "Buy-Hold T-Bill (3M)" = "gray50",
    "Perfect Foresight (3M)" = "gold"
  )) +
  scale_linetype_manual(values = c(
    "RF Tactical (3M)" = "solid",
    "LASSO Tactical (3M)" = "solid",
    "RIDGE Tactical (3M)" = "solid",
    "PCR Tactical (3M)" = "solid",
    "ELASTIC Tactical (3M)" = "solid",
    "ENSEMBLE Tactical (3M)" = "solid",
    "Buy-Hold S&P500 (3M)" = "solid",
    "Buy-Hold T-Bill (3M)" = "dashed",
    "Perfect Foresight (3M)" = "dotted"
  )) +
  labs(
    title = "Portfolio Value Evolution - Quarterly Rebalancing (3-Month Horizon)",
    subtitle = paste("Test Period:", min(backtest_data$date), "to", max(backtest_data$date), "| Log Scale"),
    x = "Date",
    y = "Portfolio Value ($)",
    color = "Strategy"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    legend.text = element_text(size = 9)
  ) +
  guides(color = guide_legend(nrow = 3), linetype = guide_legend(nrow = 3))
```

## 7.2 Portfolio Value Evolution - Semi-Annual (6M)

```{r plot-portfolio-6m, fig.height=8}
# Combine portfolio values
portfolio_6m <- bind_rows(lapply(names(results_6m), function(name) {
  results_6m[[name]]$results %>%
    select(date, portfolio_value) %>%
    mutate(Strategy = name)
}))

# Create plot
ggplot(portfolio_6m, aes(x = date, y = portfolio_value, color = Strategy, linetype = Strategy)) +
  geom_line(linewidth = 1.2) +
  scale_y_log10(labels = dollar_format()) +
  scale_color_manual(values = c(
    "RF Tactical (6M)" = "#E41A1C",
    "LASSO Tactical (6M)" = "#377EB8",
    "RIDGE Tactical (6M)" = "#4DAF4A",
    "PCR Tactical (6M)" = "#984EA3",
    "ELASTIC Tactical (6M)" = "#FF7F00",
    "ENSEMBLE Tactical (6M)" = "#A65628",
    "Buy-Hold S&P500 (6M)" = "black",
    "Buy-Hold T-Bill (6M)" = "gray50",
    "Perfect Foresight (6M)" = "gold"
  )) +
  scale_linetype_manual(values = c(
    "RF Tactical (6M)" = "solid",
    "LASSO Tactical (6M)" = "solid",
    "RIDGE Tactical (6M)" = "solid",
    "PCR Tactical (6M)" = "solid",
    "ELASTIC Tactical (6M)" = "solid",
    "ENSEMBLE Tactical (6M)" = "solid",
    "Buy-Hold S&P500 (6M)" = "solid",
    "Buy-Hold T-Bill (6M)" = "dashed",
    "Perfect Foresight (6M)" = "dotted"
  )) +
  labs(
    title = "Portfolio Value Evolution - Semi-Annual Rebalancing (6-Month Horizon)",
    subtitle = paste("Test Period:", min(backtest_data$date), "to", max(backtest_data$date), "| Log Scale"),
    x = "Date",
    y = "Portfolio Value ($)",
    color = "Strategy"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    legend.text = element_text(size = 9)
  ) +
  guides(color = guide_legend(nrow = 3), linetype = guide_legend(nrow = 3))
```

## 7.3 Portfolio Value Evolution - Annual (12M)

```{r plot-portfolio-12m, fig.height=8}
# Combine portfolio values
portfolio_12m <- bind_rows(lapply(names(results_12m), function(name) {
  results_12m[[name]]$results %>%
    select(date, portfolio_value) %>%
    mutate(Strategy = name)
}))

# Create plot
ggplot(portfolio_12m, aes(x = date, y = portfolio_value, color = Strategy, linetype = Strategy)) +
  geom_line(linewidth = 1.2) +
  scale_y_log10(labels = dollar_format()) +
  scale_color_manual(values = c(
    "RF Tactical (12M)" = "#E41A1C",
    "LASSO Tactical (12M)" = "#377EB8",
    "RIDGE Tactical (12M)" = "#4DAF4A",
    "PCR Tactical (12M)" = "#984EA3",
    "ELASTIC Tactical (12M)" = "#FF7F00",
    "ENSEMBLE Tactical (12M)" = "#A65628",
    "Buy-Hold S&P500 (12M)" = "black",
    "Buy-Hold T-Bill (12M)" = "gray50",
    "Perfect Foresight (12M)" = "gold"
  )) +
  scale_linetype_manual(values = c(
    "RF Tactical (12M)" = "solid",
    "LASSO Tactical (12M)" = "solid",
    "RIDGE Tactical (12M)" = "solid",
    "PCR Tactical (12M)" = "solid",
    "ELASTIC Tactical (12M)" = "solid",
    "ENSEMBLE Tactical (12M)" = "solid",
    "Buy-Hold S&P500 (12M)" = "solid",
    "Buy-Hold T-Bill (12M)" = "dashed",
    "Perfect Foresight (12M)" = "dotted"
  )) +
  labs(
    title = "Portfolio Value Evolution - Annual Rebalancing (12-Month Horizon)",
    subtitle = paste("Test Period:", min(backtest_data$date), "to", max(backtest_data$date), "| Log Scale"),
    x = "Date",
    y = "Portfolio Value ($)",
    color = "Strategy"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    legend.text = element_text(size = 9)
  ) +
  guides(color = guide_legend(nrow = 3), linetype = guide_legend(nrow = 3))
```

## 7.4 Asset Allocation - Random Forest Example

```{r plot-allocation, fig.height=10}
# RF allocation for all horizons
p1 <- results_3m[["RF Tactical (3M)"]]$results %>%
  filter(!is.na(position)) %>%
  mutate(position_numeric = ifelse(position == "SP500", 1, 0)) %>%
  ggplot(aes(x = date, y = position_numeric)) +
  geom_step(color = "steelblue", linewidth = 1) +
  scale_y_continuous(breaks = c(0, 1), labels = c("T-Bill", "S&P 500")) +
  labs(
    title = "Random Forest - Quarterly (3M) Allocation",
    x = NULL,
    y = "Position"
  ) +
  theme_minimal()

p2 <- results_6m[["RF Tactical (6M)"]]$results %>%
  filter(!is.na(position)) %>%
  mutate(position_numeric = ifelse(position == "SP500", 1, 0)) %>%
  ggplot(aes(x = date, y = position_numeric)) +
  geom_step(color = "darkgreen", linewidth = 1) +
  scale_y_continuous(breaks = c(0, 1), labels = c("T-Bill", "S&P 500")) +
  labs(
    title = "Random Forest - Semi-Annual (6M) Allocation",
    x = NULL,
    y = "Position"
  ) +
  theme_minimal()

p3 <- results_12m[["RF Tactical (12M)"]]$results %>%
  filter(!is.na(position)) %>%
  mutate(position_numeric = ifelse(position == "SP500", 1, 0)) %>%
  ggplot(aes(x = date, y = position_numeric)) +
  geom_step(color = "darkred", linewidth = 1) +
  scale_y_continuous(breaks = c(0, 1), labels = c("T-Bill", "S&P 500")) +
  labs(
    title = "Random Forest - Annual (12M) Allocation",
    x = "Date",
    y = "Position"
  ) +
  theme_minimal()

grid.arrange(p1, p2, p3, ncol = 1)
```

## 7.5 Risk-Return Trade-off - All Horizons

```{r plot-risk-return, fig.height=8}
# Combine all metrics
all_metrics <- bind_rows(
  metrics_df_3m %>% mutate(Horizon = "3-Month"),
  metrics_df_6m %>% mutate(Horizon = "6-Month"),
  metrics_df_12m %>% mutate(Horizon = "12-Month")
)

# Create risk-return plot
ggplot(all_metrics, aes(x = annualized_vol, y = annualized_return, 
                         color = Horizon, shape = Horizon)) +
  geom_point(size = 4, alpha = 0.7) +
  geom_text(aes(label = strategy_name), hjust = -0.1, vjust = 0, size = 3) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = c("3-Month" = "#E41A1C", 
                                  "6-Month" = "#377EB8", 
                                  "12-Month" = "#4DAF4A")) +
  labs(
    title = "Risk-Return Profile: All Strategies Across Horizons",
    subtitle = "Annualized Returns vs Annualized Volatility",
    x = "Annualized Volatility",
    y = "Annualized Return",
    color = "Rebalancing Horizon",
    shape = "Rebalancing Horizon"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )
```

---

# 8. Key Findings

## 8.1 Summary Statistics by Horizon

```{r summary-by-horizon}
cat("\n=== SUMMARY BY REBALANCING HORIZON ===\n\n")

# Calculate summary stats
summary_stats <- all_metrics %>%
  filter(!grepl("Buy-Hold|Perfect", strategy_name)) %>%
  group_by(Horizon) %>%
  summarise(
    `Mean Ann. Return` = mean(annualized_return, na.rm = TRUE),
    `Mean Sharpe` = mean(sharpe_ratio, na.rm = TRUE),
    `Mean Max DD` = mean(max_drawdown, na.rm = TRUE),
    `Mean Turnover` = mean(turnover, na.rm = TRUE),
    `Mean % in S&P500` = mean(time_in_sp500, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

kable(summary_stats, align = "lrrrrr",
      caption = "Average Performance Metrics by Horizon (ML Models Only)")
```

## 8.2 Best Strategies by Metric

```{r best-strategies}
cat("\n=== TOP PERFORMERS BY METRIC ===\n\n")

# Best by Sharpe Ratio
cat("** Highest Sharpe Ratio **\n")
top_sharpe <- all_metrics %>%
  filter(!grepl("Buy-Hold T-Bill|Perfect", strategy_name)) %>%
  arrange(desc(sharpe_ratio)) %>%
  select(Strategy = strategy_name, Horizon, `Sharpe Ratio` = sharpe_ratio, 
         `Ann. Return` = annualized_return) %>%
  head(5) %>%
  mutate(`Sharpe Ratio` = round(`Sharpe Ratio`, 3),
         `Ann. Return` = percent(`Ann. Return`, accuracy = 0.01))

print(kable(top_sharpe))

cat("\n** Highest Total Return **\n")
top_return <- all_metrics %>%
  filter(!grepl("Buy-Hold T-Bill|Perfect", strategy_name)) %>%
  arrange(desc(total_return)) %>%
  select(Strategy = strategy_name, Horizon, `Total Return` = total_return,
         `Final Value` = final_value) %>%
  head(5) %>%
  mutate(`Total Return` = percent(`Total Return`, accuracy = 0.01),
         `Final Value` = round(`Final Value`, 4))

print(kable(top_return))

cat("\n** Lowest Maximum Drawdown (Best Downside Protection) **\n")
best_dd <- all_metrics %>%
  filter(!grepl("Buy-Hold T-Bill|Perfect", strategy_name)) %>%
  arrange(desc(max_drawdown)) %>%  # Less negative is better
  select(Strategy = strategy_name, Horizon, `Max DD` = max_drawdown,
         `Ann. Return` = annualized_return) %>%
  head(5) %>%
  mutate(`Max DD` = percent(`Max DD`, accuracy = 0.01),
         `Ann. Return` = percent(`Ann. Return`, accuracy = 0.01))

print(kable(best_dd))
```

---

# 9. Export Results

```{r export-results}
cat("\n=== EXPORTING RESULTS ===\n\n")

# Create output directory if it doesn't exist
dir.create("../output", showWarnings = FALSE)

# Export performance tables
write_csv(performance_3m, "../output/performance_3m_quarterly.csv")
write_csv(performance_6m, "../output/performance_6m_semiannual.csv")
write_csv(performance_12m, "../output/performance_12m_annual.csv")

# Export combined metrics
write_csv(all_metrics, "../output/all_metrics_combined.csv")

# Export portfolio values for further analysis
write_csv(portfolio_3m, "../output/portfolio_values_3m.csv")
write_csv(portfolio_6m, "../output/portfolio_values_6m.csv")
write_csv(portfolio_12m, "../output/portfolio_values_12m.csv")

cat("✓ Results exported to ../output/ directory\n\n")
cat("Files created:\n")
cat("  - performance_3m_quarterly.csv\n")
cat("  - performance_6m_semiannual.csv\n")
cat("  - performance_12m_annual.csv\n")
cat("  - all_metrics_combined.csv\n")
cat("  - portfolio_values_3m.csv\n")
cat("  - portfolio_values_6m.csv\n")
cat("  - portfolio_values_12m.csv\n")
```

---

# 10. Conclusions

## Key Takeaways:

1. **Strategy Framework**: Successfully implemented tactical asset allocation strategies across three rebalancing horizons (quarterly, semi-annual, annual) using ML forecasts.

2. **Decision Rule**: Simple comparison of forecasted S&P 500 returns vs T-bill rates drives allocation decisions, with 100% positions in either asset.

3. **Benchmark Hierarchy**: Expected ranking is T-Bill < [ML Models] ≤ Buy-Hold S&P 500 < Perfect Foresight.

4. **Performance Metrics**: Evaluated strategies using:
   - Total and annualized returns
   - Sharpe ratios (risk-adjusted performance)
   - Maximum drawdowns (downside risk)
   - Win rates and turnover (tactical behavior)
   - Time in market allocation

5. **Horizon Comparison**: Longer horizons (12M) typically show:
   - Lower turnover (fewer rebalancing decisions)
   - Smoother portfolio paths
   - Different risk-return profiles

6. **Model Comparison**: Identifies which ML models generate the most valuable forecasts for tactical allocation.

## Next Steps:

- Add transaction costs sensitivity analysis
- Test alternative allocation rules (e.g., partial positions)
- Analyze timing of position switches vs market regimes
- Compare with equal-weighted or optimized ensemble strategies

---

**End of Report**
