---
title: "Trading Strategy Backtest: S&P 500 vs T-Bills"
author: "Strategy Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# SETTING UP LIBRARIES
library(dplyr)
library(tidyverse)
library(readxl)
library(lubridate)
library(ggplot2)
library(knitr)
library(scales)


```

# Trading Strategy Overview

## Strategy Logic

**Tactical Asset Allocation Strategy:**
- Switch between S&P 500 and T-bills based on return forecasts
- Decision rule: Invest in asset with higher predicted return for next quarter
- Portfolio rebalanced quarterly based on new forecasts

**Key Features:**
- Starting capital: $1
- Transaction cost: 0% (adjustable parameter)
- Test period: User-defined start and end dates
- Comparison benchmarks: Buy-and-hold S&P 500, Buy-and-hold T-bills

---

# 1. Load Data

## 1.1 Load T-Bill Data (All Horizons)
```{r load-tbill}
cat("\n=== LOADING T-BILL DATA (ALL HORIZONS) ===\n\n")

# Testing period starts 2011-10-01 (Q207)
TEST_START_DATE <- as.Date("2011-10-01")

# ===== 3-MONTH T-BILL (Quarterly) =====
cat("1. Loading 3-Month T-Bill (TB3MS) - for quarterly strategy...\n")
tbill_3m <- read_xlsx("../data/TB3MS.xlsx", sheet = 2) %>%
  rename(date = observation_date) %>%
  mutate(date = as.Date(date)) %>%
  filter(date >= TEST_START_DATE) %>%
  arrange(date) %>%
  mutate(
    # TB3MS is annualized bank-discount yield (%)
    # Convert to 3-month holding period return
    discount_rate = TB3MS / 100,
    price = 1 - discount_rate * (90/360),
    tbill_return_1q = (1 / price) - 1
  ) %>%
  select(date, tbill_return_1q)

cat("   ✓ 3M T-Bill loaded:", nrow(tbill_3m), "quarterly observations\n")
cat("   Date range:", min(tbill_3m$date), "to", max(tbill_3m$date), "\n\n")

# ===== 6-MONTH T-BILL (Daily → Convert to Quarterly) =====
cat("2. Loading 6-Month T-Bill (DGS6MO) - for semi-annual strategy...\n")
tbill_6m_daily <- read_csv("../data/DGS6MO.csv") %>%
  rename(date = observation_date, rate = DGS6MO) %>%
  mutate(date = as.Date(date)) %>%
  filter(!is.na(rate), date >= TEST_START_DATE) %>%
  arrange(date)

# Convert daily 6M T-Bill yields to quarterly observations
tbill_6m <- tbill_6m_daily %>%
  mutate(
    year = year(date),
    quarter = quarter(date)
  ) %>%
  # Keep only last observation of each quarter
  group_by(year, quarter) %>%
  slice_max(date, n = 1) %>%
  ungroup() %>%
  mutate(
    # DGS6MO is annualized yield (%)
    # Convert to 6-month holding period return
    annual_yield = rate / 100,
    tbill_return_6m = (1 + annual_yield)^(0.5) - 1,
    # Standardize to quarterly dates (first day of quarter)
    date = floor_date(date, "quarter")
  ) %>%
  select(date, tbill_return_6m)

cat("   ✓ 6M T-Bill loaded:", nrow(tbill_6m), "quarterly observations\n")
cat("   Date range:", min(tbill_6m$date), "to", max(tbill_6m$date), "\n\n")

# ===== 1-YEAR T-BILL (Daily → Convert to Quarterly) =====
cat("3. Loading 1-Year T-Bill (DGS1) - for annual strategy...\n")
tbill_1y_daily <- read_csv("../data/DGS1.csv") %>%
  rename(date = observation_date, rate = DGS1) %>%
  mutate(date = as.Date(date)) %>%
  filter(!is.na(rate), date >= TEST_START_DATE) %>%
  arrange(date)

# Convert daily 1Y T-Bill yields to quarterly observations
tbill_1y <- tbill_1y_daily %>%
  mutate(
    year = year(date),
    quarter = quarter(date)
  ) %>%
  group_by(year, quarter) %>%
  slice_max(date, n = 1) %>%
  ungroup() %>%
  mutate(
    # DGS1 is annualized yield (%)
    # Use as-is for 1-year holding period return
    tbill_return_1y = rate / 100,
    # Standardize to quarterly dates
    date = floor_date(date, "quarter")
  ) %>%
  select(date, tbill_return_1y)

cat("   ✓ 1Y T-Bill loaded:", nrow(tbill_1y), "quarterly observations\n")
cat("   Date range:", min(tbill_1y$date), "to", max(tbill_1y$date), "\n\n")

# ===== COMBINE ALL T-BILL HORIZONS =====
cat("=== COMBINING T-BILL DATA ===\n")
tbill_all <- tbill_3m %>%
  left_join(tbill_6m, by = "date") %>%
  left_join(tbill_1y, by = "date") %>%
  # Keep only rows where ALL three horizons have data
  filter(!is.na(tbill_return_1q) & !is.na(tbill_return_6m) & !is.na(tbill_return_1y))

cat("✓ T-Bill data combined (TESTING PERIOD ONLY):\n")
cat("  Observations:", nrow(tbill_all), "quarters\n")
cat("  Date range:", min(tbill_all$date), "to", max(tbill_all$date), "\n")
cat("  Columns: tbill_return_1q, tbill_return_6m, tbill_return_1y\n\n")

cat("Sample T-Bill data (first 10 rows):\n")
print(head(tbill_all, 10))
```


## 1.2 Load ML Model Forecasts (Multi-Horizon)
```{r load-forecasts}
cat("\n=== LOADING MULTI-HORIZON ML MODEL FORECASTS ===\n\n")

# ===== RANDOM FOREST: All Horizons =====
# File: rf_evals_all_horizons.csv (long format with 'horizon' column)
cat("1. Loading Random Forest predictions (all horizons)...\n")
rf_all <- read_csv("../data/rf_evals_all_horizons.csv") %>%
  select(date, actual, forecast, horizon) %>%
  arrange(date, horizon)

# Reshape RF from long to wide format
rf_forecasts <- rf_all %>%
  pivot_wider(
    names_from = horizon,
    values_from = c(actual, forecast),
    names_glue = "{.value}_{horizon}"
  ) %>%
  rename(
    actual_1q = actual_3m,
    actual_1y = actual_12m,
    rf_forecast_1q = forecast_3m,
    rf_forecast_1y = forecast_12m
    # Note: actual_6m and forecast_6m already have correct names
  ) %>%
  arrange(date)

cat("   ✓ RF loaded:", nrow(rf_forecasts), "observations\n")
cat("   Horizons: 1Q (3m), 6M, 1Y (12m)\n\n")

# ===== XGBOOST: All Horizons =====
# TODO: Add XGBoost multi-horizon predictions when available
# Expected file: xgb_evals_all_horizons.csv or similar
cat("2. XGBoost predictions...\n")
cat("   ⚠ XGBoost multi-horizon file not found in data folder\n")
cat("   NOTE: Will skip XGBoost for now. Add file when ready.\n\n")

# ===== LASSO: All Horizons =====
cat("3. Loading Lasso predictions (all horizons)...\n")
lasso_forecasts <- read_csv("../data/lasso_test_predictions.csv") %>%
  rename(
    actual_1q = actual_3m,
    actual_1y = actual_12m,
    lasso_forecast_1q = predicted_3m,
    lasso_forecast_6m = predicted_6m,
    lasso_forecast_1y = predicted_12m
  ) %>%
  select(date, lasso_forecast_1q, lasso_forecast_6m, lasso_forecast_1y) %>%
  arrange(date)

cat("   ✓ Lasso loaded:", nrow(lasso_forecasts), "observations\n")
cat("   Horizons: 1Q (3m), 6M, 1Y (12m)\n\n")

# ===== RIDGE: All Horizons =====
cat("4. Loading Ridge predictions (all horizons)...\n")
ridge_forecasts <- read_csv("../data/ridge_test_predictions.csv") %>%
  rename(
    ridge_forecast_1q = predicted_3m,
    ridge_forecast_6m = predicted_6m,
    ridge_forecast_1y = predicted_12m
  ) %>%
  select(date, ridge_forecast_1q, ridge_forecast_6m, ridge_forecast_1y) %>%
  arrange(date)

cat("   ✓ Ridge loaded:", nrow(ridge_forecasts), "observations\n")
cat("   Horizons: 1Q (3m), 6M, 1Y (12m)\n\n")

# ===== PCR: All Horizons =====
cat("5. Loading PCR predictions (all horizons)...\n")
pcr_forecasts <- read_csv("../data/pcr_test_predictions.csv") %>%
  rename(
    pcr_forecast_1q = predicted_3m,
    pcr_forecast_6m = predicted_6m,
    pcr_forecast_1y = predicted_12m
  ) %>%
  select(date, pcr_forecast_1q, pcr_forecast_6m, pcr_forecast_1y) %>%
  arrange(date)

cat("   ✓ PCR loaded:", nrow(pcr_forecasts), "observations\n")
cat("   Horizons: 1Q (3m), 6M, 1Y (12m)\n\n")

# ===== ELASTIC NET: All Horizons =====
cat("6. Loading Elastic Net predictions (all horizons)...\n")
elastic_forecasts <- read_csv("../data/elastic_net_test_predictions.csv") %>%
  rename(
    elastic_forecast_1q = predicted_3m,
    elastic_forecast_6m = predicted_6m,
    elastic_forecast_1y = predicted_12m
  ) %>%
  select(date, elastic_forecast_1q, elastic_forecast_6m, elastic_forecast_1y) %>%
  arrange(date)

cat("   ✓ Elastic Net loaded:", nrow(elastic_forecasts), "observations\n")
cat("   Horizons: 1Q (3m), 6M, 1Y (12m)\n\n")

# ===== ENSEMBLE: All Horizons =====
cat("7. Loading Ensemble predictions (all horizons)...\n")
cat("   ⚠ Ensemble multi-horizon file not yet created\n")
cat("   NOTE: Run ensemble_lsei.Rmd first to generate ensemble forecasts\n")
cat("   Expected file: ensemble_predictions_multihorizon.csv\n\n")

# ===== COMBINE ALL FORECASTS =====
cat("=== COMBINING ALL FORECASTS ===\n")
forecast_data <- rf_forecasts %>%
  left_join(lasso_forecasts, by = "date") %>%
  left_join(ridge_forecasts, by = "date") %>%
  left_join(pcr_forecasts, by = "date") %>%
  left_join(elastic_forecasts, by = "date")

cat("\n✓ Multi-horizon forecast data combined\n")
cat("  Observations:", nrow(forecast_data), "quarters\n")
cat("  Date range:", min(forecast_data$date), "to", max(forecast_data$date), "\n\n")

cat("Available columns:\n")
cat("  Actual returns:\n")
cat("    - actual_1q: Actual 1-quarter (3-month) S&P 500 returns\n")
cat("    - actual_6m: Actual 6-month S&P 500 returns\n")
cat("    - actual_1y: Actual 1-year (12-month) S&P 500 returns\n\n")
cat("  Model forecasts (1Q horizon):\n")
cat("    - rf_forecast_1q, lasso_forecast_1q, ridge_forecast_1q,\n")
cat("      pcr_forecast_1q, elastic_forecast_1q\n\n")
cat("  Model forecasts (6M horizon):\n")
cat("    - rf_forecast_6m, lasso_forecast_6m, ridge_forecast_6m,\n")
cat("      pcr_forecast_6m, elastic_forecast_6m\n\n")
cat("  Model forecasts (1Y horizon):\n")
cat("    - rf_forecast_1y, lasso_forecast_1y, ridge_forecast_1y,\n")
cat("      pcr_forecast_1y, elastic_forecast_1y\n\n")

cat("Sample data (first 5 rows):\n")
print(head(forecast_data, 5))
```

## 1.3 Merge Forecasts with T-Bill Data
```{r merge-data}
cat("\n=== MERGING FORECAST DATA WITH T-BILL RETURNS ===\n")

# Merge all forecast data with T-bill returns
backtest_data <- forecast_data %>%
  left_join(tbill_all, by = "date") %>%
  drop_na()  # Remove any rows with missing data

cat("\n✓ Backtest dataset ready!\n")
cat("  Observations:", nrow(backtest_data), "quarters\n")
cat("  Date range:", min(backtest_data$date), "to", max(backtest_data$date), "\n\n")

cat("=== FINAL DATASET STRUCTURE ===\n")
cat("\n📊 ACTUAL RETURNS (3 horizons):\n")
cat("  - actual_1q: Realized 1-quarter S&P 500 returns\n")
cat("  - actual_6m: Realized 6-month S&P 500 returns\n")
cat("  - actual_1y: Realized 1-year S&P 500 returns\n\n")

cat("🔮 MODEL FORECASTS (1Q - Quarterly Rebalancing):\n")
cat("  - rf_forecast_1q, lasso_forecast_1q, ridge_forecast_1q,\n")
cat("    pcr_forecast_1q, elastic_forecast_1q\n\n")

cat("🔮 MODEL FORECASTS (6M - Semi-Annual Rebalancing):\n")
cat("  - rf_forecast_6m, lasso_forecast_6m, ridge_forecast_6m,\n")
cat("    pcr_forecast_6m, elastic_forecast_6m\n\n")

cat("🔮 MODEL FORECASTS (1Y - Annual Rebalancing):\n")
cat("  - rf_forecast_1y, lasso_forecast_1y, ridge_forecast_1y,\n")
cat("    pcr_forecast_1y, elastic_forecast_1y\n\n")

cat("💰 T-BILL RETURNS (Risk-Free Rates):\n")
cat("  - tbill_return_1q: 3-month T-bill (quarterly)\n")
cat("  - tbill_return_6m: 6-month T-bill (semi-annual)\n")
cat("  - tbill_return_1y: 1-year T-bill (annual)\n\n")

cat("Total columns in dataset:", ncol(backtest_data), "\n\n")

cat("Column names:\n")
print(names(backtest_data))

cat("\n\nSample merged data (first 5 rows):\n")
print(head(backtest_data, 5))

cat("\n\nSummary statistics:\n")
summary(backtest_data %>% select(actual_1q, actual_6m, actual_1y, 
                                   tbill_return_1q, tbill_return_6m, tbill_return_1y))
```

---

# 2. Backtesting Framework

## 2.1 Define Strategy Function
```{r strategy-function}
#' Tactical Asset Allocation Backtest
#' 
#' @param predictions Data frame with columns: date, predicted_sp500_return
#' @param actual_returns Data frame with columns: date, actual_sp500_return
#' @param tbill_returns Data frame with columns: date, tbill_quarterly_return
#' @param start_date Start date for backtest (default: first date in predictions)
#' @param end_date End date for backtest (default: last date in predictions)
#' @param initial_capital Starting capital (default: 1)
#' @param transaction_cost Transaction cost as decimal (default: 0.0 for 0%)
#' 
#' @return List containing portfolio values, positions, and metrics
tactical_strategy_backtest <- function(predictions, 
                                       actual_returns, 
                                       tbill_returns,
                                       start_date = NULL,
                                       end_date = NULL,
                                       initial_capital = 1,
                                       transaction_cost = 0.0) {
  
  # Merge all data
  backtest_data <- predictions %>%
    inner_join(actual_returns, by = "date") %>%
    inner_join(tbill_returns, by = "date") %>%
    arrange(date)
  
  # Apply date filters
  if (!is.null(start_date)) {
    backtest_data <- backtest_data %>% filter(date >= as.Date(start_date))
  }
  if (!is.null(end_date)) {
    backtest_data <- backtest_data %>% filter(date <= as.Date(end_date))
  }
  
  n_periods <- nrow(backtest_data)
  
  if (n_periods == 0) {
    stop("No data available for the specified date range")
  }
  
  cat("\n=== Backtesting Strategy ===\n")
  cat("Period:", min(backtest_data$date), "to", max(backtest_data$date), "\n")
  cat("Number of quarters:", n_periods, "\n")
  cat("Transaction cost:", transaction_cost * 100, "%\n\n")
  
  # Initialize tracking vectors
  portfolio_value <- numeric(n_periods + 1)
  position <- character(n_periods + 1)  # "SP500" or "TBILL"
  returns_earned <- numeric(n_periods)
  switches <- numeric(n_periods)
  
  # Starting position
  portfolio_value[1] <- initial_capital
  
  # Period 0 → 1 decision (initial position)
  if (backtest_data$predicted_sp500_return[1] >= backtest_data$tbill_quarterly_return[1]) {
    position[1] <- "SP500"
    returns_earned[1] <- backtest_data$actual_sp500_return[1]
  } else {
    position[1] <- "TBILL"
    returns_earned[1] <- backtest_data$tbill_quarterly_return[1]
  }
  
  switches[1] <- 0  # No switch on first period
  portfolio_value[2] <- portfolio_value[1] * (1 + returns_earned[1])
  position[2] <- position[1]
  
  # Loop through remaining periods
  for (t in 2:n_periods) {
    
    # Decision: Compare predicted S&P vs T-bill for next period
    signal_sp500 <- backtest_data$predicted_sp500_return[t] >= 
                    backtest_data$tbill_quarterly_return[t]
    
    # Determine position and return for period t
    if (signal_sp500) {
      # Signal says: S&P 500 is better
      if (position[t] == "SP500") {
        # Already in S&P → HOLD
        position[t + 1] <- "SP500"
        returns_earned[t] <- backtest_data$actual_sp500_return[t]
        switches[t] <- 0
      } else {
        # Currently in T-bill → SWITCH to S&P
        position[t + 1] <- "SP500"
        returns_earned[t] <- backtest_data$actual_sp500_return[t] - transaction_cost
        switches[t] <- 1
      }
    } else {
      # Signal says: T-bill is better
      if (position[t] == "TBILL") {
        # Already in T-bill → HOLD
        position[t + 1] <- "TBILL"
        returns_earned[t] <- backtest_data$tbill_quarterly_return[t]
        switches[t] <- 0
      } else {
        # Currently in S&P → SWITCH to T-bill
        position[t + 1] <- "TBILL"
        returns_earned[t] <- backtest_data$tbill_quarterly_return[t] - transaction_cost
        switches[t] <- 1
      }
    }
    
    # Update portfolio value
    portfolio_value[t + 1] <- portfolio_value[t] * (1 + returns_earned[t])
  }
  
  # Create results dataframe
  results_df <- data.frame(
    date = c(backtest_data$date[1], backtest_data$date),  # Include t=0
    portfolio_value = portfolio_value,
    position = position,
    returns_earned = c(NA, returns_earned),  # NA for t=0
    switches = c(NA, switches),
    predicted_sp500 = c(NA, backtest_data$predicted_sp500_return),
    actual_sp500 = c(NA, backtest_data$actual_sp500_return),
    tbill_return = c(NA, backtest_data$tbill_quarterly_return)
  )
  
  # Calculate performance metrics
  total_return <- (portfolio_value[n_periods + 1] / portfolio_value[1]) - 1
  annualized_return <- (1 + total_return)^(4 / n_periods) - 1  # Quarterly to annual
  
  # Sharpe ratio (assuming quarterly risk-free rate from T-bills)
  excess_returns <- returns_earned - backtest_data$tbill_quarterly_return
  sharpe_ratio <- mean(excess_returns, na.rm = TRUE) / sd(excess_returns, na.rm = TRUE) * sqrt(4)
  
  # Max drawdown
  cummax_value <- cummax(portfolio_value)
  drawdown <- (portfolio_value - cummax_value) / cummax_value
  max_drawdown <- min(drawdown)
  
  # Win rate (quarters with positive excess return)
  win_rate <- sum(excess_returns > 0, na.rm = TRUE) / sum(!is.na(excess_returns))
  
  # Turnover (number of switches / total periods)
  turnover <- sum(switches, na.rm = TRUE) / n_periods
  
  metrics <- list(
    total_return = total_return,
    annualized_return = annualized_return,
    sharpe_ratio = sharpe_ratio,
    max_drawdown = max_drawdown,
    win_rate = win_rate,
    turnover = turnover,
    n_switches = sum(switches, na.rm = TRUE),
    final_value = portfolio_value[n_periods + 1]
  )
  
  return(list(
    results = results_df,
    metrics = metrics
  ))
}
```

## 3.2 Define Benchmark Functions
```{r benchmark-functions}
#' Buy-and-Hold S&P 500 Benchmark
buy_hold_sp500 <- function(actual_returns, start_date = NULL, end_date = NULL, 
                           initial_capital = 1) {
  
  data <- actual_returns %>% arrange(date)
  
  if (!is.null(start_date)) data <- data %>% filter(date >= as.Date(start_date))
  if (!is.null(end_date)) data <- data %>% filter(date <= as.Date(end_date))
  
  n_periods <- nrow(data)
  portfolio_value <- cumprod(c(initial_capital, 1 + data$actual_sp500_return))
  
  results_df <- data.frame(
    date = c(data$date[1], data$date),
    portfolio_value = portfolio_value,
    position = "SP500"
  )
  
  total_return <- (portfolio_value[n_periods + 1] / portfolio_value[1]) - 1
  annualized_return <- (1 + total_return)^(4 / n_periods) - 1
  
  metrics <- list(
    total_return = total_return,
    annualized_return = annualized_return,
    final_value = portfolio_value[n_periods + 1]
  )
  
  return(list(results = results_df, metrics = metrics))
}

#' Buy-and-Hold T-Bill Benchmark
buy_hold_tbill <- function(tbill_returns, start_date = NULL, end_date = NULL, 
                           initial_capital = 1) {
  
  data <- tbill_returns %>% arrange(date)
  
  if (!is.null(start_date)) data <- data %>% filter(date >= as.Date(start_date))
  if (!is.null(end_date)) data <- data %>% filter(date <= as.Date(end_date))
  
  n_periods <- nrow(data)
  portfolio_value <- cumprod(c(initial_capital, 1 + data$tbill_quarterly_return))
  
  results_df <- data.frame(
    date = c(data$date[1], data$date),
    portfolio_value = portfolio_value,
    position = "TBILL"
  )
  
  total_return <- (portfolio_value[n_periods + 1] / portfolio_value[1]) - 1
  annualized_return <- (1 + total_return)^(4 / n_periods) - 1
  
  metrics <- list(
    total_return = total_return,
    annualized_return = annualized_return,
    final_value = portfolio_value[n_periods + 1]
  )
  
  return(list(results = results_df, metrics = metrics))
}
```

---

# 3. Run Backtest

## 3.1 Set Backtest Parameters
```{r backtest-params}
# === USER PARAMETERS ===
START_DATE <- "2012-01-01"  # First date in your test set (adjust if needed)
END_DATE <- "2024-06-30"    # Last date in your test set
INITIAL_CAPITAL <- 1
TRANSACTION_COST <- 0.00    # 0% (can change to 0.001 for 0.1%, etc.)

cat("\n=== BACKTEST PARAMETERS ===\n")
cat("Start Date:", START_DATE, "\n")
cat("End Date:", END_DATE, "\n")
cat("Initial Capital:", INITIAL_CAPITAL, "\n")
cat("Transaction Cost:", TRANSACTION_COST * 100, "%\n\n")
```

## 3.2 Run Strategy for Each Model
```{r run-backtest}
cat("\n=== RUNNING BACKTESTS ===\n\n")

# Prepare data for tactical_strategy_backtest function
# The function expects: predicted_sp500_return, actual_sp500_return, tbill_quarterly_return

actual_returns <- backtest_data %>%
  select(date, actual_sp500_return = log_ret)

tbill_returns <- backtest_data %>%
  select(date, tbill_quarterly_return)

# Backtest 1: Random Forest Strategy
cat("1. Random Forest Tactical Strategy\n")
predictions_rf <- backtest_data %>% 
  select(date, predicted_sp500_return = rf_forecast_ret)

backtest_rf <- tactical_strategy_backtest(
  predictions = predictions_rf,
  actual_returns = actual_returns,
  tbill_returns = tbill_returns,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL,
  transaction_cost = TRANSACTION_COST
)

# Backtest 2: XGBoost Strategy
cat("\n2. XGBoost Tactical Strategy\n")
predictions_xgb <- backtest_data %>% 
  select(date, predicted_sp500_return = xgb_forecast_ret)

backtest_xgb <- tactical_strategy_backtest(
  predictions = predictions_xgb,
  actual_returns = actual_returns,
  tbill_returns = tbill_returns,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL,
  transaction_cost = TRANSACTION_COST
)

# Backtest 3: Lasso Strategy
cat("\n3. Lasso Tactical Strategy\n")
predictions_lasso <- backtest_data %>% 
  select(date, predicted_sp500_return = lasso_forecast_ret)

backtest_lasso <- tactical_strategy_backtest(
  predictions = predictions_lasso,
  actual_returns = actual_returns,
  tbill_returns = tbill_returns,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL,
  transaction_cost = TRANSACTION_COST
)

# Backtest 4: Ridge Strategy
cat("\n4. Ridge Tactical Strategy\n")
predictions_ridge <- backtest_data %>% 
  select(date, predicted_sp500_return = ridge_forecast_ret)

backtest_ridge <- tactical_strategy_backtest(
  predictions = predictions_ridge,
  actual_returns = actual_returns,
  tbill_returns = tbill_returns,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL,
  transaction_cost = TRANSACTION_COST
)

# Backtest 5: PCR Strategy
cat("\n5. PCR Tactical Strategy\n")
predictions_pcr <- backtest_data %>% 
  select(date, predicted_sp500_return = pcr_forecast_ret)

backtest_pcr <- tactical_strategy_backtest(
  predictions = predictions_pcr,
  actual_returns = actual_returns,
  tbill_returns = tbill_returns,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL,
  transaction_cost = TRANSACTION_COST
)

# Backtest 6: Elastic Net Strategy
cat("\n6. Elastic Net Tactical Strategy\n")
predictions_elastic <- backtest_data %>% 
  select(date, predicted_sp500_return = elastic_forecast_ret)

backtest_elastic <- tactical_strategy_backtest(
  predictions = predictions_elastic,
  actual_returns = actual_returns,
  tbill_returns = tbill_returns,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL,
  transaction_cost = TRANSACTION_COST
)

# Backtest 7: Ensemble (LSEI) Strategy
cat("\n7. Ensemble (LSEI) Tactical Strategy\n")
predictions_ensemble <- backtest_data %>% 
  select(date, predicted_sp500_return = ensemble_forecast_ret)

backtest_ensemble <- tactical_strategy_backtest(
  predictions = predictions_ensemble,
  actual_returns = actual_returns,
  tbill_returns = tbill_returns,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL,
  transaction_cost = TRANSACTION_COST
)

# Backtest 8: Buy-and-Hold Benchmarks
cat("\n8. Buy-and-Hold S&P 500 Benchmark\n")
benchmark_sp500 <- buy_hold_sp500(
  actual_returns = actual_returns,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL
)

cat("\n9. Buy-and-Hold T-Bill Benchmark\n")
benchmark_tbill <- buy_hold_tbill(
  tbill_returns = tbill_returns,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL
)

cat("\n10. Perfect Foresight Benchmark (Oracle)\n")
cat("Note: This uses actual realized returns as predictions\n")
cat("      It represents the theoretical maximum tactical strategy performance\n")
predictions_all_right <- backtest_data %>% 
  select(date, predicted_sp500_return = log_ret)

benchmark_all_right <- tactical_strategy_backtest(
  predictions = predictions_all_right,
  actual_returns = actual_returns,
  tbill_returns = tbill_returns,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL,
  transaction_cost = TRANSACTION_COST
)
cat("\n=== BACKTESTS COMPLETED ===\n")
```

---

# 4. Performance Comparison

## 4.1 Summary Table
```{r performance-table}

# Run benchmarks
cat("\n" , rep("=", 80), "\n")
cat("BENCHMARK: Buy-and-Hold S&P 500\n")
benchmark_sp500 <- buy_hold_sp500(
  actual_returns = actual_returns,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL
)

cat("\n" , rep("=", 80), "\n")
cat("BENCHMARK: Buy-and-Hold T-Bill\n")
benchmark_tbill <- buy_hold_tbill(
  tbill_returns = tbill_quarterly,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL
)
```

---

# 4. Performance Comparison

## 4.1 Summary Table
```{r performance-table}
# Create performance comparison table
performance_table <- data.frame(
  Strategy = c("Random Forest", "XGBoost", "Lasso", "Ridge", "PCR", "Elastic Net",
               "Ensemble (LSEI)", "Buy-Hold S&P500", "Buy-Hold T-Bill", "Perfect Foresight"),
  
  Total_Return = percent(c(
    backtest_rf$metrics$total_return,
    backtest_xgb$metrics$total_return,
    backtest_lasso$metrics$total_return,
    backtest_ridge$metrics$total_return,
    backtest_pcr$metrics$total_return,
    backtest_elastic$metrics$total_return,
    backtest_ensemble$metrics$total_return,
    benchmark_sp500$metrics$total_return,
    benchmark_tbill$metrics$total_return,
    benchmark_all_right$metrics$total_return
  ), accuracy = 0.01),
  
  Annualized_Return = percent(c(
    backtest_rf$metrics$annualized_return,
    backtest_xgb$metrics$annualized_return,
    backtest_lasso$metrics$annualized_return,
    backtest_ridge$metrics$annualized_return,
    backtest_pcr$metrics$annualized_return,
    backtest_elastic$metrics$annualized_return,
    backtest_ensemble$metrics$annualized_return,
    benchmark_sp500$metrics$annualized_return,
    benchmark_tbill$metrics$annualized_return,
    benchmark_all_right$metrics$annualized_return
  ), accuracy = 0.01),
  
  Sharpe_Ratio = c(
    round(backtest_rf$metrics$sharpe_ratio, 3),
    round(backtest_xgb$metrics$sharpe_ratio, 3),
    round(backtest_lasso$metrics$sharpe_ratio, 3),
    round(backtest_ridge$metrics$sharpe_ratio, 3),
    round(backtest_pcr$metrics$sharpe_ratio, 3),
    round(backtest_elastic$metrics$sharpe_ratio, 3),
    round(backtest_ensemble$metrics$sharpe_ratio, 3),
    NA,  # Buy-Hold S&P500
    NA,  # Buy-Hold T-Bill
    round(benchmark_all_right$metrics$sharpe_ratio, 3)
  ),
  
  Max_Drawdown = percent(c(
    backtest_rf$metrics$max_drawdown,
    backtest_xgb$metrics$max_drawdown,
    backtest_lasso$metrics$max_drawdown,
    backtest_ridge$metrics$max_drawdown,
    backtest_pcr$metrics$max_drawdown,
    backtest_elastic$metrics$max_drawdown,
    backtest_ensemble$metrics$max_drawdown,
    NA,  # Buy-Hold S&P500
    NA,  # Buy-Hold T-Bill
    benchmark_all_right$metrics$max_drawdown
  ), accuracy = 0.01),
  
  Win_Rate = percent(c(
    backtest_rf$metrics$win_rate,
    backtest_xgb$metrics$win_rate,
    backtest_lasso$metrics$win_rate,
    backtest_ridge$metrics$win_rate,
    backtest_pcr$metrics$win_rate,
    backtest_elastic$metrics$win_rate,
    backtest_ensemble$metrics$win_rate,
    NA,  # Buy-Hold S&P500
    NA,  # Buy-Hold T-Bill
    benchmark_all_right$metrics$win_rate
  ), accuracy = 0.01),
  
  Turnover = percent(c(
    backtest_rf$metrics$turnover,
    backtest_xgb$metrics$turnover,
    backtest_lasso$metrics$turnover,
    backtest_ridge$metrics$turnover,
    backtest_pcr$metrics$turnover,
    backtest_elastic$metrics$turnover,
    backtest_ensemble$metrics$turnover,
    NA,  # Buy-Hold S&P500
    NA,  # Buy-Hold T-Bill
    benchmark_all_right$metrics$turnover
  ), accuracy = 0.01),
  
  N_Switches = c(
    backtest_rf$metrics$n_switches,
    backtest_xgb$metrics$n_switches,
    backtest_lasso$metrics$n_switches,
    backtest_ridge$metrics$n_switches,
    backtest_pcr$metrics$n_switches,
    backtest_elastic$metrics$n_switches,
    backtest_ensemble$metrics$n_switches,
    0,  # Buy-Hold S&P500
    0,  # Buy-Hold T-Bill
    benchmark_all_right$metrics$n_switches
  ),
  
  Final_Value = round(c(
    backtest_rf$metrics$final_value,
    backtest_xgb$metrics$final_value,
    backtest_lasso$metrics$final_value,
    backtest_ridge$metrics$final_value,
    backtest_pcr$metrics$final_value,
    backtest_elastic$metrics$final_value,
    backtest_ensemble$metrics$final_value,
    benchmark_sp500$metrics$final_value,
    benchmark_tbill$metrics$final_value,
    benchmark_all_right$metrics$final_value
  ), 4)
)

cat("\n=== PERFORMANCE COMPARISON ===\n\n")
kable(performance_table, align = "lrrrrrrrr", 
      caption = "Strategy Performance Metrics")
```

---

# 6. Visualizations

## 6.1 Portfolio Value Over Time
```{r portfolio-plot, fig.width=12, fig.height=6}
# Combine all portfolio values
## 4.2 Portfolio Value Comparison

# Combine all portfolio value trajectories
portfolio_comparison <- bind_rows(
  backtest_rf$results %>% select(date, portfolio_value) %>% mutate(Strategy = "Random Forest"),
  backtest_xgb$results %>% select(date, portfolio_value) %>% mutate(Strategy = "XGBoost"),
  backtest_lasso$results %>% select(date, portfolio_value) %>% mutate(Strategy = "Lasso"),
  backtest_ridge$results %>% select(date, portfolio_value) %>% mutate(Strategy = "Ridge"),
  backtest_pcr$results %>% select(date, portfolio_value) %>% mutate(Strategy = "PCR"),
  backtest_elastic$results %>% select(date, portfolio_value) %>% mutate(Strategy = "Elastic Net"),
  backtest_ensemble$results %>% select(date, portfolio_value) %>% mutate(Strategy = "Ensemble (LSEI)"),
  benchmark_sp500$results %>% select(date, portfolio_value) %>% mutate(Strategy = "Buy-Hold S&P 500"),
  benchmark_all_right$results %>% select(date, portfolio_value) %>% mutate(Strategy = "Perfect Foresight"),
  benchmark_tbill$results %>% select(date, portfolio_value) %>% mutate(Strategy = "Buy-Hold T-Bill")
)

ggplot(portfolio_comparison, aes(x = date, y = portfolio_value, color = Strategy)) +
  geom_line(linewidth = 1) +
  scale_y_log10(labels = dollar_format()) +
  scale_color_manual(values = c(
    "Random Forest" = "#E41A1C", 
    "XGBoost" = "#377EB8",
    "Lasso" = "#4DAF4A",
    "Ridge" = "#984EA3",
    "PCR" = "#FF7F00",
    "Elastic Net" = "#FFFF33",
    "Ensemble (LSEI)" = "#A65628",
    "Perfect Foresight" = "#00FF00",
    "Buy-Hold S&P 500" = "black",
    "Buy-Hold T-Bill" = "gray50"
  )) +
  labs(
    title = "Portfolio Value Comparison: All Strategies vs Benchmarks",
    subtitle = paste("Period:", START_DATE, "to", END_DATE, "| Log Scale"),
    x = "Date",
    y = "Portfolio Value ($)",
    color = "Strategy"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )
```

## 4.3 Random Forest Asset Allocation Over Time
```{r allocation-rf, fig.width=12, fig.height=6}
# Show position switches for Random Forest
position_data_rf <- backtest_rf$results %>%
  filter(!is.na(position)) %>%
  mutate(position_numeric = ifelse(position == "SP500", 1, 0))

ggplot(position_data_rf, aes(x = date, y = position_numeric)) +
  geom_step(color = "steelblue", linewidth = 1) +
  scale_y_continuous(breaks = c(0, 1), labels = c("T-Bill", "S&P 500")) +
  labs(
    title = "Random Forest: Asset Allocation Over Time",
    subtitle = paste("Switches:", backtest_rf$metrics$n_switches, 
                     "| Turnover:", percent(backtest_rf$metrics$turnover, accuracy = 0.1)),
    x = "Date",
    y = "Position"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank()
  )
```

## 4.4 XGBoost Asset Allocation Over Time
```{r allocation-xgb, fig.width=12, fig.height=6}
# Show position switches for XGBoost
position_data_xgb <- backtest_xgb$results %>%
  filter(!is.na(position)) %>%
  mutate(position_numeric = ifelse(position == "SP500", 1, 0))

ggplot(position_data_xgb, aes(x = date, y = position_numeric)) +
  geom_step(color = "darkgreen", linewidth = 1) +
  scale_y_continuous(breaks = c(0, 1), labels = c("T-Bill", "S&P 500")) +
  labs(
    title = "XGBoost: Asset Allocation Over Time",
    subtitle = paste("Switches:", backtest_xgb$metrics$n_switches, 
                     "| Turnover:", percent(backtest_xgb$metrics$turnover, accuracy = 0.1)),
    x = "Date",
    y = "Position"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank()
  )
```

## 4.5 Return Distribution
```{r return-dist, fig.width=10, fig.height=6}
# Compare return distributions
return_comparison <- bind_rows(
  backtest_rf$results %>% filter(!is.na(returns_earned)) %>% 
    select(date, returns_earned) %>% mutate(Strategy = "Random Forest"),
  backtest_xgb$results %>% filter(!is.na(returns_earned)) %>% 
    select(date, returns_earned) %>% mutate(Strategy = "XGBoost")
)

ggplot(return_comparison, aes(x = returns_earned, fill = Strategy)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  facet_wrap(~Strategy, ncol = 1) +
  scale_fill_manual(values = c("Random Forest" = "red", "XGBoost" = "blue")) +
  labs(
    title = "Quarterly Return Distribution by Strategy",
    x = "Quarterly Return",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14)
  )
```

---

# 7. Export Results

```{r export-results}
# Save detailed results for further analysis
write.csv(strategy1$results, "../output/strategy1_results.csv", row.names = FALSE)
write.csv(strategy2$results, "../output/strategy2_results.csv", row.names = FALSE)
write.csv(strategy3$results, "../output/strategy3_results.csv", row.names = FALSE)
write.csv(performance_table, "../output/performance_comparison.csv", row.names = FALSE)

cat("\nResults exported to ../output/ directory\n")
```

---

# 8. Next Steps

**To Use with Real ML Predictions:**

1. **Replace sample forecasts** with actual predictions from your models:
   ```r
   # Load predictions from your main analysis file
   rf_predictions <- read.csv("path/to/rf_predictions.csv")
   xgb_predictions <- read.csv("path/to/xgb_predictions.csv")
   pm_predictions <- read.csv("path/to/prevailing_mean_predictions.csv")
   ```

2. **Adjust test period** to match your analysis (Q207-Q258):
   ```r
   START_DATE <- "2011-10-01"  # Q207
   END_DATE <- "2024-06-30"    # Q258
   ```

3. **Add transaction costs** if needed:
   ```r
   TRANSACTION_COST <- 0.001  # 0.1%
   ```

4. **Compare models** to see which forecast strategy produces best risk-adjusted returns!

---

**Summary:**
This backtesting framework allows you to test any forecast model's ability to generate alpha through tactical asset allocation between S&P 500 and T-bills. The key metric is whether the strategy outperforms simple buy-and-hold on a risk-adjusted basis (Sharpe ratio).
