---
title: "Trading Strategy Backtest: S&P 500 vs T-Bills"
author: "Strategy Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# SETTING UP LIBRARIES
library(dplyr)
library(tidyverse)
library(readxl)
library(lubridate)
library(ggplot2)
library(knitr)
library(scales)
```

# Trading Strategy Overview

## Strategy Logic

**Tactical Asset Allocation Strategy:**
- Switch between S&P 500 and T-bills based on return forecasts
- Decision rule: Invest in asset with higher predicted return for next quarter
- Portfolio rebalanced quarterly based on new forecasts

**Key Features:**
- Starting capital: $1
- Transaction cost: 0% (adjustable parameter)
- Test period: User-defined start and end dates
- Comparison benchmarks: Buy-and-hold S&P 500, Buy-and-hold T-bills

---

# 1. Load Data

## 1.1 Load T-Bill Data
```{r load-tbill}
cat("Loading T-Bill data...\n")

# Load T-Bill 3-month rate (TB3MS)
# Data is already quarterly: Jan, Apr, Jul, Oct (same as S&P 500)
tbill_data <- read_xlsx("../data/TB3MS.xlsx", sheet = 2) %>%
  rename(date = observation_date) %>%
  filter(year(date) >= 1960) %>%
  arrange(date)

# Convert bank-discount yield to quarterly holding-period return
# TB3MS is annualized bank-discount yield (%)
# 
# Formula (from FRED-MD documentation):
# Price per $1 face = 1 - (discount_rate × 90/360)
# Final value = 1 / Price = 1 / (1 - discount_rate × 90/360)
# Quarterly return = Final value - 1
#
# Simplified: return = (discount_rate × 90/360) / (1 - discount_rate × 90/360)
#
# Example: If TB3MS = 3.8% (0.038 as decimal):
#   Price = 1 - 0.038 × (90/360) = 0.9905
#   Return = (1/0.9905) - 1 ≈ 0.00959 (0.959% for 3 months)

tbill_data <- tbill_data %>%
  mutate(
    # Convert percentage to decimal
    discount_rate = TB3MS / 100,
    
    # Calculate price per $1 face value (90-day T-bill, 360-day basis)
    price = 1 - discount_rate * (90/360),
    
    # Calculate 3-month holding period return
    tbill_quarterly_return = (1 / price) - 1
  ) %>%
  select(date, TB3MS, discount_rate, price, tbill_quarterly_return)

cat("T-Bill data loaded:", nrow(tbill_data), "quarterly observations\n")
cat("Date range:", min(tbill_data$date), "to", max(tbill_data$date), "\n")
cat("\nSample data (showing calculation):\n")
head(tbill_data, 10)
```

## 1.2 Load Actual S&P 500 Returns
```{r load-sp500}
cat("\nLoading actual S&P 500 returns...\n")

# Load from merged data (created in data cleaning script)
actual_returns <- read.csv("../data/data_merged.csv") %>%
  mutate(date = as.Date(date)) %>%
  select(date, log_ret) %>%
  rename(actual_sp500_return = log_ret) %>%
  filter(!is.na(actual_sp500_return))

cat("Actual S&P 500 returns loaded:", nrow(actual_returns), "quarterly observations\n")
cat("Date range:", min(actual_returns$date), "to", max(actual_returns$date), "\n")
head(actual_returns)
```

## 1.3 Prepare T-Bill Data for Backtesting
```{r prepare-tbill}
# T-Bill data is already quarterly - just rename for clarity
tbill_quarterly <- tbill_data %>%
  select(date, tbill_quarterly_return) %>%
  arrange(date)

cat("\nT-Bill data ready for backtesting:", nrow(tbill_quarterly), "quarterly observations\n")
cat("Date range:", min(tbill_quarterly$date), "to", max(tbill_quarterly$date), "\n")
cat("Sample:\n")
head(tbill_quarterly, 10)
```

---

# 2. Create Sample Forecast Data

## 2.1 Sample Predictions (Placeholder)
```{r sample-predictions}
cat("\n=== CREATING SAMPLE FORECAST DATA ===\n")
cat("Note: Replace this with actual ML model predictions later\n\n")

# Create sample forecast data structure
# In practice, this will come from your RF, XGBoost, Prevailing Mean models

set.seed(123)

# Generate sample predictions for last 52 quarters (matching your test set)
# Dates: Q207-Q258 (approximately 2011-2024)
sample_dates <- tail(actual_returns$date, 52)

sample_forecasts <- data.frame(
  date = sample_dates,
  
  # Sample Model 1: Optimistic predictions (higher than actual on average)
  model1_predicted_return = actual_returns$actual_sp500_return[
    (nrow(actual_returns)-51):nrow(actual_returns)
  ] + rnorm(52, mean = 0.01, sd = 0.02),
  
  # Sample Model 2: Conservative predictions (lower than actual on average)
  model2_predicted_return = actual_returns$actual_sp500_return[
    (nrow(actual_returns)-51):nrow(actual_returns)
  ] - rnorm(52, mean = 0.005, sd = 0.015),
  
  # Sample Model 3: Prevailing Mean (simple average)
  model3_predicted_return = actual_returns$actual_sp500_return[
    (nrow(actual_returns)-51):nrow(actual_returns)
  ] + rnorm(52, mean = 0, sd = 0.01)
)

cat("Sample forecasts created for", nrow(sample_forecasts), "quarters\n")
cat("Date range:", min(sample_forecasts$date), "to", max(sample_forecasts$date), "\n\n")

print(head(sample_forecasts, 10))
```

---

# 3. Backtesting Framework

## 3.1 Define Strategy Function
```{r strategy-function}
#' Tactical Asset Allocation Backtest
#' 
#' @param predictions Data frame with columns: date, predicted_sp500_return
#' @param actual_returns Data frame with columns: date, actual_sp500_return
#' @param tbill_returns Data frame with columns: date, tbill_quarterly_return
#' @param start_date Start date for backtest (default: first date in predictions)
#' @param end_date End date for backtest (default: last date in predictions)
#' @param initial_capital Starting capital (default: 1)
#' @param transaction_cost Transaction cost as decimal (default: 0.0 for 0%)
#' 
#' @return List containing portfolio values, positions, and metrics
tactical_strategy_backtest <- function(predictions, 
                                       actual_returns, 
                                       tbill_returns,
                                       start_date = NULL,
                                       end_date = NULL,
                                       initial_capital = 1,
                                       transaction_cost = 0.0) {
  
  # Merge all data
  backtest_data <- predictions %>%
    inner_join(actual_returns, by = "date") %>%
    inner_join(tbill_returns, by = "date") %>%
    arrange(date)
  
  # Apply date filters
  if (!is.null(start_date)) {
    backtest_data <- backtest_data %>% filter(date >= as.Date(start_date))
  }
  if (!is.null(end_date)) {
    backtest_data <- backtest_data %>% filter(date <= as.Date(end_date))
  }
  
  n_periods <- nrow(backtest_data)
  
  if (n_periods == 0) {
    stop("No data available for the specified date range")
  }
  
  cat("\n=== Backtesting Strategy ===\n")
  cat("Period:", min(backtest_data$date), "to", max(backtest_data$date), "\n")
  cat("Number of quarters:", n_periods, "\n")
  cat("Transaction cost:", transaction_cost * 100, "%\n\n")
  
  # Initialize tracking vectors
  portfolio_value <- numeric(n_periods + 1)
  position <- character(n_periods + 1)  # "SP500" or "TBILL"
  returns_earned <- numeric(n_periods)
  switches <- numeric(n_periods)
  
  # Starting position
  portfolio_value[1] <- initial_capital
  
  # Period 0 → 1 decision (initial position)
  if (backtest_data$predicted_sp500_return[1] >= backtest_data$tbill_quarterly_return[1]) {
    position[1] <- "SP500"
    returns_earned[1] <- backtest_data$actual_sp500_return[1]
  } else {
    position[1] <- "TBILL"
    returns_earned[1] <- backtest_data$tbill_quarterly_return[1]
  }
  
  switches[1] <- 0  # No switch on first period
  portfolio_value[2] <- portfolio_value[1] * (1 + returns_earned[1])
  position[2] <- position[1]
  
  # Loop through remaining periods
  for (t in 2:n_periods) {
    
    # Decision: Compare predicted S&P vs T-bill for next period
    signal_sp500 <- backtest_data$predicted_sp500_return[t] >= 
                    backtest_data$tbill_quarterly_return[t]
    
    # Determine position and return for period t
    if (signal_sp500) {
      # Signal says: S&P 500 is better
      if (position[t] == "SP500") {
        # Already in S&P → HOLD
        position[t + 1] <- "SP500"
        returns_earned[t] <- backtest_data$actual_sp500_return[t]
        switches[t] <- 0
      } else {
        # Currently in T-bill → SWITCH to S&P
        position[t + 1] <- "SP500"
        returns_earned[t] <- backtest_data$actual_sp500_return[t] - transaction_cost
        switches[t] <- 1
      }
    } else {
      # Signal says: T-bill is better
      if (position[t] == "TBILL") {
        # Already in T-bill → HOLD
        position[t + 1] <- "TBILL"
        returns_earned[t] <- backtest_data$tbill_quarterly_return[t]
        switches[t] <- 0
      } else {
        # Currently in S&P → SWITCH to T-bill
        position[t + 1] <- "TBILL"
        returns_earned[t] <- backtest_data$tbill_quarterly_return[t] - transaction_cost
        switches[t] <- 1
      }
    }
    
    # Update portfolio value
    portfolio_value[t + 1] <- portfolio_value[t] * (1 + returns_earned[t])
  }
  
  # Create results dataframe
  results_df <- data.frame(
    date = c(backtest_data$date[1], backtest_data$date),  # Include t=0
    portfolio_value = portfolio_value,
    position = position,
    returns_earned = c(NA, returns_earned),  # NA for t=0
    switches = c(NA, switches),
    predicted_sp500 = c(NA, backtest_data$predicted_sp500_return),
    actual_sp500 = c(NA, backtest_data$actual_sp500_return),
    tbill_return = c(NA, backtest_data$tbill_quarterly_return)
  )
  
  # Calculate performance metrics
  total_return <- (portfolio_value[n_periods + 1] / portfolio_value[1]) - 1
  annualized_return <- (1 + total_return)^(4 / n_periods) - 1  # Quarterly to annual
  
  # Sharpe ratio (assuming quarterly risk-free rate from T-bills)
  excess_returns <- returns_earned - backtest_data$tbill_quarterly_return
  sharpe_ratio <- mean(excess_returns, na.rm = TRUE) / sd(excess_returns, na.rm = TRUE) * sqrt(4)
  
  # Max drawdown
  cummax_value <- cummax(portfolio_value)
  drawdown <- (portfolio_value - cummax_value) / cummax_value
  max_drawdown <- min(drawdown)
  
  # Win rate (quarters with positive excess return)
  win_rate <- sum(excess_returns > 0, na.rm = TRUE) / sum(!is.na(excess_returns))
  
  # Turnover (number of switches / total periods)
  turnover <- sum(switches, na.rm = TRUE) / n_periods
  
  metrics <- list(
    total_return = total_return,
    annualized_return = annualized_return,
    sharpe_ratio = sharpe_ratio,
    max_drawdown = max_drawdown,
    win_rate = win_rate,
    turnover = turnover,
    n_switches = sum(switches, na.rm = TRUE),
    final_value = portfolio_value[n_periods + 1]
  )
  
  return(list(
    results = results_df,
    metrics = metrics
  ))
}
```

## 3.2 Define Benchmark Functions
```{r benchmark-functions}
#' Buy-and-Hold S&P 500 Benchmark
buy_hold_sp500 <- function(actual_returns, start_date = NULL, end_date = NULL, 
                           initial_capital = 1) {
  
  data <- actual_returns %>% arrange(date)
  
  if (!is.null(start_date)) data <- data %>% filter(date >= as.Date(start_date))
  if (!is.null(end_date)) data <- data %>% filter(date <= as.Date(end_date))
  
  n_periods <- nrow(data)
  portfolio_value <- cumprod(c(initial_capital, 1 + data$actual_sp500_return))
  
  results_df <- data.frame(
    date = c(data$date[1], data$date),
    portfolio_value = portfolio_value,
    position = "SP500"
  )
  
  total_return <- (portfolio_value[n_periods + 1] / portfolio_value[1]) - 1
  annualized_return <- (1 + total_return)^(4 / n_periods) - 1
  
  metrics <- list(
    total_return = total_return,
    annualized_return = annualized_return,
    final_value = portfolio_value[n_periods + 1]
  )
  
  return(list(results = results_df, metrics = metrics))
}

#' Buy-and-Hold T-Bill Benchmark
buy_hold_tbill <- function(tbill_returns, start_date = NULL, end_date = NULL, 
                           initial_capital = 1) {
  
  data <- tbill_returns %>% arrange(date)
  
  if (!is.null(start_date)) data <- data %>% filter(date >= as.Date(start_date))
  if (!is.null(end_date)) data <- data %>% filter(date <= as.Date(end_date))
  
  n_periods <- nrow(data)
  portfolio_value <- cumprod(c(initial_capital, 1 + data$tbill_quarterly_return))
  
  results_df <- data.frame(
    date = c(data$date[1], data$date),
    portfolio_value = portfolio_value,
    position = "TBILL"
  )
  
  total_return <- (portfolio_value[n_periods + 1] / portfolio_value[1]) - 1
  annualized_return <- (1 + total_return)^(4 / n_periods) - 1
  
  metrics <- list(
    total_return = total_return,
    annualized_return = annualized_return,
    final_value = portfolio_value[n_periods + 1]
  )
  
  return(list(results = results_df, metrics = metrics))
}
```

---

# 4. Run Backtest

## 4.1 Set Backtest Parameters
```{r backtest-params}
# === USER PARAMETERS ===
START_DATE <- "2011-01-01"  # Adjust to your test period
END_DATE <- "2024-06-30"    # Adjust to your test period
INITIAL_CAPITAL <- 1
TRANSACTION_COST <- 0.00    # 0% (can change to 0.001 for 0.1%, etc.)

cat("\n=== BACKTEST PARAMETERS ===\n")
cat("Start Date:", START_DATE, "\n")
cat("End Date:", END_DATE, "\n")
cat("Initial Capital:", INITIAL_CAPITAL, "\n")
cat("Transaction Cost:", TRANSACTION_COST * 100, "%\n\n")
```

## 4.2 Run Strategy for Each Model
```{r run-backtest}
# Prepare prediction data for each model
predictions_model1 <- sample_forecasts %>% 
  select(date, predicted_sp500_return = model1_predicted_return)

predictions_model2 <- sample_forecasts %>% 
  select(date, predicted_sp500_return = model2_predicted_return)

predictions_model3 <- sample_forecasts %>% 
  select(date, predicted_sp500_return = model3_predicted_return)

# Run tactical strategy for each model
cat("\n" , rep("=", 80), "\n")
cat("MODEL 1: Optimistic Forecaster\n")
strategy1 <- tactical_strategy_backtest(
  predictions = predictions_model1,
  actual_returns = actual_returns,
  tbill_returns = tbill_quarterly,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL,
  transaction_cost = TRANSACTION_COST
)

cat("\n" , rep("=", 80), "\n")
cat("MODEL 2: Conservative Forecaster\n")
strategy2 <- tactical_strategy_backtest(
  predictions = predictions_model2,
  actual_returns = actual_returns,
  tbill_returns = tbill_quarterly,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL,
  transaction_cost = TRANSACTION_COST
)

cat("\n" , rep("=", 80), "\n")
cat("MODEL 3: Prevailing Mean\n")
strategy3 <- tactical_strategy_backtest(
  predictions = predictions_model3,
  actual_returns = actual_returns,
  tbill_returns = tbill_quarterly,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL,
  transaction_cost = TRANSACTION_COST
)

# Run benchmarks
cat("\n" , rep("=", 80), "\n")
cat("BENCHMARK: Buy-and-Hold S&P 500\n")
benchmark_sp500 <- buy_hold_sp500(
  actual_returns = actual_returns,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL
)

cat("\n" , rep("=", 80), "\n")
cat("BENCHMARK: Buy-and-Hold T-Bill\n")
benchmark_tbill <- buy_hold_tbill(
  tbill_returns = tbill_quarterly,
  start_date = START_DATE,
  end_date = END_DATE,
  initial_capital = INITIAL_CAPITAL
)
```

---

# 5. Performance Comparison

## 5.1 Summary Table
```{r performance-table}
# Create performance comparison table
performance_table <- data.frame(
  Strategy = c("Model 1 (Optimistic)", "Model 2 (Conservative)", 
               "Model 3 (Prevailing)", "Buy-Hold S&P500", "Buy-Hold T-Bill"),
  
  Total_Return = percent(c(
    strategy1$metrics$total_return,
    strategy2$metrics$total_return,
    strategy3$metrics$total_return,
    benchmark_sp500$metrics$total_return,
    benchmark_tbill$metrics$total_return
  ), accuracy = 0.01),
  
  Annualized_Return = percent(c(
    strategy1$metrics$annualized_return,
    strategy2$metrics$annualized_return,
    strategy3$metrics$annualized_return,
    benchmark_sp500$metrics$annualized_return,
    benchmark_tbill$metrics$annualized_return
  ), accuracy = 0.01),
  
  Sharpe_Ratio = c(
    round(strategy1$metrics$sharpe_ratio, 3),
    round(strategy2$metrics$sharpe_ratio, 3),
    round(strategy3$metrics$sharpe_ratio, 3),
    NA,
    NA
  ),
  
  Max_Drawdown = percent(c(
    strategy1$metrics$max_drawdown,
    strategy2$metrics$max_drawdown,
    strategy3$metrics$max_drawdown,
    NA,
    NA
  ), accuracy = 0.01),
  
  Win_Rate = percent(c(
    strategy1$metrics$win_rate,
    strategy2$metrics$win_rate,
    strategy3$metrics$win_rate,
    NA,
    NA
  ), accuracy = 0.01),
  
  Turnover = percent(c(
    strategy1$metrics$turnover,
    strategy2$metrics$turnover,
    strategy3$metrics$turnover,
    NA,
    NA
  ), accuracy = 0.01),
  
  N_Switches = c(
    strategy1$metrics$n_switches,
    strategy2$metrics$n_switches,
    strategy3$metrics$n_switches,
    0,
    0
  ),
  
  Final_Value = round(c(
    strategy1$metrics$final_value,
    strategy2$metrics$final_value,
    strategy3$metrics$final_value,
    benchmark_sp500$metrics$final_value,
    benchmark_tbill$metrics$final_value
  ), 4)
)

cat("\n=== PERFORMANCE COMPARISON ===\n\n")
kable(performance_table, align = "lrrrrrrrr", 
      caption = "Strategy Performance Metrics")
```

---

# 6. Visualizations

## 6.1 Portfolio Value Over Time
```{r portfolio-plot, fig.width=12, fig.height=6}
# Combine all portfolio values
portfolio_comparison <- bind_rows(
  strategy1$results %>% select(date, portfolio_value) %>% mutate(Strategy = "Model 1"),
  strategy2$results %>% select(date, portfolio_value) %>% mutate(Strategy = "Model 2"),
  strategy3$results %>% select(date, portfolio_value) %>% mutate(Strategy = "Model 3"),
  benchmark_sp500$results %>% select(date, portfolio_value) %>% mutate(Strategy = "Buy-Hold SP500"),
  benchmark_tbill$results %>% select(date, portfolio_value) %>% mutate(Strategy = "Buy-Hold T-Bill")
)

ggplot(portfolio_comparison, aes(x = date, y = portfolio_value, color = Strategy)) +
  geom_line(linewidth = 1) +
  scale_y_log10(labels = dollar_format()) +
  scale_color_manual(values = c("Model 1" = "red", "Model 2" = "blue", 
                                 "Model 3" = "darkgreen", "Buy-Hold SP500" = "black",
                                 "Buy-Hold T-Bill" = "gray50")) +
  labs(
    title = "Portfolio Value Comparison: Tactical Strategies vs Benchmarks",
    subtitle = paste("Period:", START_DATE, "to", END_DATE, "| Log Scale"),
    x = "Date",
    y = "Portfolio Value ($)",
    color = "Strategy"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )
```

## 6.2 Asset Allocation Over Time (Model 1 Example)
```{r allocation-plot, fig.width=12, fig.height=6}
# Show position switches for Model 1
position_data <- strategy1$results %>%
  filter(!is.na(position)) %>%
  mutate(position_numeric = ifelse(position == "SP500", 1, 0))

ggplot(position_data, aes(x = date, y = position_numeric)) +
  geom_step(color = "steelblue", linewidth = 1) +
  scale_y_continuous(breaks = c(0, 1), labels = c("T-Bill", "S&P 500")) +
  labs(
    title = "Model 1: Asset Allocation Over Time",
    subtitle = paste("Switches:", strategy1$metrics$n_switches, 
                     "| Turnover:", percent(strategy1$metrics$turnover, accuracy = 0.1)),
    x = "Date",
    y = "Position"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank()
  )
```

## 6.3 Return Distribution
```{r return-dist, fig.width=10, fig.height=6}
# Compare return distributions
return_comparison <- bind_rows(
  strategy1$results %>% filter(!is.na(returns_earned)) %>% 
    select(date, returns_earned) %>% mutate(Strategy = "Model 1"),
  strategy2$results %>% filter(!is.na(returns_earned)) %>% 
    select(date, returns_earned) %>% mutate(Strategy = "Model 2"),
  strategy3$results %>% filter(!is.na(returns_earned)) %>% 
    select(date, returns_earned) %>% mutate(Strategy = "Model 3")
)

ggplot(return_comparison, aes(x = returns_earned, fill = Strategy)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  facet_wrap(~Strategy, ncol = 1) +
  scale_fill_manual(values = c("Model 1" = "red", "Model 2" = "blue", 
                                "Model 3" = "darkgreen")) +
  labs(
    title = "Quarterly Return Distribution by Strategy",
    x = "Quarterly Return",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14)
  )
```

---

# 7. Export Results

```{r export-results}
# Save detailed results for further analysis
write.csv(strategy1$results, "../output/strategy1_results.csv", row.names = FALSE)
write.csv(strategy2$results, "../output/strategy2_results.csv", row.names = FALSE)
write.csv(strategy3$results, "../output/strategy3_results.csv", row.names = FALSE)
write.csv(performance_table, "../output/performance_comparison.csv", row.names = FALSE)

cat("\nResults exported to ../output/ directory\n")
```

---

# 8. Next Steps

**To Use with Real ML Predictions:**

1. **Replace sample forecasts** with actual predictions from your models:
   ```r
   # Load predictions from your main analysis file
   rf_predictions <- read.csv("path/to/rf_predictions.csv")
   xgb_predictions <- read.csv("path/to/xgb_predictions.csv")
   pm_predictions <- read.csv("path/to/prevailing_mean_predictions.csv")
   ```

2. **Adjust test period** to match your analysis (Q207-Q258):
   ```r
   START_DATE <- "2011-10-01"  # Q207
   END_DATE <- "2024-06-30"    # Q258
   ```

3. **Add transaction costs** if needed:
   ```r
   TRANSACTION_COST <- 0.001  # 0.1%
   ```

4. **Compare models** to see which forecast strategy produces best risk-adjusted returns!

---

**Summary:**
This backtesting framework allows you to test any forecast model's ability to generate alpha through tactical asset allocation between S&P 500 and T-bills. The key metric is whether the strategy outperforms simple buy-and-hold on a risk-adjusted basis (Sharpe ratio).
